---
title: "MPA Exposure Analysis"
format: 
  html:
    code-fold: true
editor: visual
---

```{r}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

------------------------------------------------------------------------

# CA MPA Exposure to Future Climate

*Characterization of exposure of California MPAs to future pH, DO, and temperature using IPSL model predictions.*

## Setup

Load required libraries, read IPSL MPA projections file, and assign MPAs to regions (Northern CA, Central CA, Southern CA, Channel Islands).

```{r}
#| label: libraries
#| message: false

library(tidyverse)
library(lubridate)
library(data.table)
library(factoextra)
library(broom)
library(cowplot)
library(respR)
library(here)
library(lattice)
library(RcppRoll)
library(RColorBrewer)
library(gplots)
library(ggpmisc)
library(dendextend)
library(colourvalues)
library(zoo)
library(terra) #needed for tmap?
library(tmap) 
library(sf) 
library(heatmaply)
```

```{r}
#| label: load_data
#| message: false

#Read mpa data, convert DO units, add a column for time period, and one for difference between mid/end cen and historic.
mpa <- read_csv(here("data/processeddata/model/IPSLmpa.csv")) %>% 
  mutate(DO_mmolL = DO_surf/1000,
         DO_mgL = convert_DO(DO_mmolL, from = "mmol/L", to = "mg/L")) %>% 
  filter(Year <= 2020|
         Year >= 2040 & Year <= 2060|
         Year >= 2080 & Year <= 2100) %>%
  mutate(period = case_when((Year <= 2020) ~ "historic", 
                           (Year %in% c(2040:2060)) ~ "midcen",
                           (Year %in% c(2080:2100)) ~ "endcen")) %>%
  select(-T_bot, -DO_bot, -pH_bot,-DO_surf,-DO_mmolL) %>%
  rename(Temp = T_surf,
         DO = DO_mgL,
         pH = pH_surf) %>%
  mutate(Date = make_date(Year, Month, Day)) %>%
  mutate(julianday = yday(Date))

mpa$File <- substr(mpa$File, 1, nchar(mpa$File)-4)

```

```{r}
#| label: regions
#| message: false
#add regions here!
mpa_centroids<- read_csv(here("data/rawdata/MPA_polygons.csv")) %>%
  select(-Area_sq_mi, -Type)
mpa_centroids$File <- sub("^", "tphdo_mpa_", mpa_centroids$OBJECTID )

mpa_centroids <- mpa_centroids %>%
  mutate(region = ifelse(degy >= 37.29, "norca", 
                         ifelse(degy > 34.8, "centralca", 
                                ifelse("degy" < 34.274 & "degx" < -119.220, "channelisl" , 
                                       "socal"))))

channel <- c("Anacapa Island FMCA", "Anacapa Island FMR", "Anacapa Island SMCA", "Anacapa Island SMR", "Anacapa Island Special Closure", 
             "Arrow Point to Lion Head Point SMCA", "Begg Rock SMR", "Blue Cavern Offshore SMCA", "Blue Cavern Onshore SMCA (No-Take)", 
             "Carrington Point SMR", "Casino Point SMCA (No-Take)", "Cat Harbor SMCA", "Farnsworth Offshore SMCA", "Farnsworth Onshore SMCA", 
             "Footprint FMR", "Footprint SMR", "Gull Island FMR", "Gull Island SMR", "Harris Point FMR", "Harris Point SMR", "Judith Rock SMR", 
             "Long Point SMR", "Lover’s Cove SMCA", "Painted Cave SMCA", "Richardson Rock FMR", "Richardson Rock SMR", "San Miguel Island Special Closure", 
             "Santa Barbara Island FMR", "Santa Barbara Island SMR", "Scorpion FMR", "Scorpion SMR", "Skunk Point SMR", "South Point FMR", "South Point SMR")

mpa_centroids$region[mpa_centroids$NAME %in% channel]  <- "channel" 
mpa <- merge(mpa, mpa_centroids,  by = "File")

```

## Calculate climatology

Create a climatology (for each MPA within a time period, Jan values averaged over all years - same for each month), so each MPA has one value per month that is averaged over all of the years. Seasonal SD is calculated later in summary stat section, and is the variation across these 12 climatologies for each MPA, representing the seasonal cycle.

```{r}
#| label: climatology
#| message: false

#calculate climatology for each MPA for each time period
mpa_climatology <- mpa %>%
  group_by(File, Month, period) %>%
    summarise(T_clim = mean(Temp), 
              pH_clim = mean(pH), 
              DO_clim =   mean(DO))

```

## Calculate event-based variation

To understand event-based variation (not attributed to the seasonal cycle), we need to find the deviation of actual daily values from a climatological value. To do so, a mock dataset with julian day climatologies is created by interpolating monthly climatology values. Actual daily values are subtracted from this interpolated dataset at the daily scale. The standard deviation of this difference is the event-based SD, which is calculated in the summary stat section.

*Q: instead of interpolating from monthly climatologies to create daily climatologies, why not average daily values to get daily climatologies?*

**Setup for interpolation:**

Need to create a day 1 and day 365 proxy. The approx function can only interpolate not extrapolate. The approx function is used to interpolate between the 15th day of each month, and day 1 and day 365 are used to interpolate the first 15 and last 15 days of the year.

```{r}
#| label: inter_setup
#| message: false
#| warning: false

##below creates empty vectors as big as we need (365 days per mpa) for each variable and each mpa. julianday is dates 1-365 as many mpa times

mpaslist = unique(mpa$File) 
mpas = rep(NA, 365*121)
julianday = rep(1:365, 121)
interp = rep(NA, 365*121)

#Set up a vector of julian day assignment for the 15th of each month and the first and last day of the year
x_in <- yday(as.Date(c("2001-01-01", 
                       "2001-01-15","2001-02-15","2001-03-15","2001-04-15",
                       "2001-05-15","2001-06-15","2001-07-15","2001-08-15",
                       "2001-09-15","2001-10-15","2001-11-15","2001-12-15", 
                       "2001-12-31")))

# creating a list of all the days of the year to interpolate to.
x_out <- (1:365) 

```

```{r}
#| label: interpolation

interpolate <- function(periodt, variable){
  
  mpa_period_climatology <- mpa_climatology %>%
    filter(period == periodt)
  
  for (i in 1:length(mpaslist)){ #for each mpa
    
    d <- mpa_period_climatology %>% #for given mpa iterating through, selects var from clim
      filter(File == mpaslist[i]) %>% 
      select(File,Month, !!sym(variable))
    
#use a weighted average. for dec 31, slightly more than half weight to dec (denoted by [12,3] index - 12 is the month, third column is the variable's clim value), little less than   half weight to jan. for jan 1, slightly more weight to jan, little less weight to dec. 
    Dec31 = as.numeric(( ((16/30) * (d[12,3])) +  
                           ((14/30) * d[1,3]) )) 
    Jan1 = as.numeric(( ((14/30) * (d[12,3])) + 
                          ((16/30) * d[1,3]) ))

    y_in <- c(Jan1, d[[variable]], Dec31)
    
    mod <- approx(x = x_in, #days to interpolate from
                 y = y_in, #temp values per day in x_in
                 xout = x_out) #1-365 days to interpolate out to
    
    #I don't understand this part
    mpas[((i-1)*365+1):(i*365)] <- mpaslist[i] #rep(mpaslist[1], 365)
    interp[((i-1)*365+1):(i*365)] <- mod$y
  }
  
  df <- data.frame(mpas, julianday, interp) %>%
    rename(File = mpas) %>%
    mutate(period = periodt) %>%
    rename(!!sym(variable) := interp)
  
  assign(x = paste("interp",variable,periodt, sep="_"), value = df, envir = globalenv())
  
}

interpolate("historic", "T_clim")
interpolate("midcen", "T_clim")
interpolate("endcen", "T_clim")

interpolate("historic", "pH_clim")
interpolate("midcen", "pH_clim")
interpolate("endcen", "pH_clim")

interpolate("historic", "DO_clim")
interpolate("midcen", "DO_clim")
interpolate("endcen", "DO_clim")
```

```{r}
#| label: combine_interp_dataframes

interpolated_temp <- rbind(interp_T_clim_historic,
                           interp_T_clim_midcen,
                           interp_T_clim_endcen)

interpolated_pH <- rbind(interp_pH_clim_historic,
                           interp_pH_clim_midcen,
                           interp_pH_clim_endcen)

interpolated_DO <- rbind(interp_DO_clim_historic,
                           interp_DO_clim_midcen,
                           interp_DO_clim_endcen)
```

## Create summary statistics

Generate summary statistics for T, pH, and DO.

*Seasonal SD* is across climatologies (monthly values averaged over all years resulting in 12 values per MPA - one per month - the seasonal SD is the variation across these values).

*Event SD* is calculated based on the deviation of daily values from interpolated daily values. The standard deviation per MPA is across all of the daily differences between daily value and interpolated values filtered for upwelling months May, June, July, August, and September (*filtered*, not *grouped by* upwelling month - the Event SD captures variation of the deviations at a daily scale).

Additional summary stats include the mean, lower 10th percentile, and upper 10th percentile.

```{r}
#| label: summary_stats
#| warning: false

#calculate seasonal SD 
seasonal_SD <- mpa_climatology %>%
  group_by(File, period) %>%
  summarise(T_seasonalSD = sd(T_clim),
            pH_seasonalSD = sd(pH_clim),
            DO_seasonalSD = sd(DO_clim))

#put all summary stats together - first, calculate event SD
sum <- mpa %>%
 
  #merge interpolated climatologies to mpa. (used to calculate event sd)
  merge(interpolated_temp, by = c("File", "julianday", "period")) %>%
  merge(interpolated_pH, by = c("File", "julianday", "period")) %>%
  merge(interpolated_DO, by = c("File", "julianday", "period")) %>%
  
  #subtract actual values - interpolated climatology values 
  #filtered for only upwelling months!!
  mutate(temp_deviation = Temp - T_clim,
         pH_deviation = pH - pH_clim,
         DO_deviation = DO - DO_clim) %>%
  filter(Month == c(5,6,7,8,9)) %>%
  
  #find event based SD 
  group_by(File, period) %>%
    mutate(T_eventSD = sd(temp_deviation),
            pH_eventSD = sd(pH_deviation),
            DO_eventSD = sd(DO_deviation)) %>%
  ungroup() %>%
    select(-T_clim, -pH_clim, -DO_clim, -temp_deviation,
         -pH_deviation, -DO_deviation) %>%
  
    #merge seasonal SD 
    merge(seasonal_SD, by = c("File", "period")) %>%
  
    #adding other summary stats - mean, low 10th and upper 10th percentiles
    group_by(File, period) %>%
    mutate(across(c(Temp, DO, pH), 
                   list(mean = mean, 
                        low10 = ~ quantile(.x, 0.1),
                        high10 = ~quantile(.x,0.9)))) %>%
  
  # to get to scale of one row per mpa
  select(-...1, -Temp, -pH, -DO, -julianday, -Year, -Month, -Day) %>%
  distinct(File, .keep_all = TRUE) %>% ungroup()

```

## PCA

PCA of all summary stats (mean, lower 10th percentile, upper 10th percentile, seasonal SD, event SD) clustered by regions.

```{r}
#| label: pca_function

make_pca <- function(df, periodt){
  sum_period <- df %>% filter(period == periodt) 
  sumsub <- sum_period %>% select(-OBJECTID,-NAME, -File, -SHORTNAME, -degx, 
                                  -degy, -region, -period, -Date) 
  
  pca <- prcomp(sumsub, scale = TRUE)
  
  plot <- fviz_pca_biplot(pca, repel = TRUE,
                  col.var = "black",
                  col.ind = sum_period$region,
                  label ="var",
                  labelsize = 3,
                  addEllipses = TRUE,
                  title = paste("PCA of summary statistics in", periodt,"period")) 
  #fviz_pca_ind(pca, label="none", habillage=sum_period$region,
               #addEllipses=TRUE,  col.ind = sum_period$region)
  
  ggsave(here(paste("./figs/pca/pca",periodt,".png")), plot)
}
```

```{r}
#| label: plot_pca

make_pca(sum, "historic")
make_pca(sum, "midcen")
make_pca(sum, "endcen") #need to rotate by changing sumsub to sumsub[,-1] inside prcomp
```

## Heatmaps by variable

Create heatmaps per summary stat showing time periods on x-axis, and all MPAs on y-axis to pick up on finer scale geographical patterns in change of summary stats over time.

```{r}
#| label: heatmap_function

col <- colorRampPalette(brewer.pal(9,"YlOrRd"))(256)
invert_col <- colorRampPalette(rev(brewer.pal(9,"YlOrRd")))(256)


make_heatmap <- function(sumstat, color_palette = invert_col, title){
  
  #make matrix for variable of interest
  matrix <- sum %>%
  mutate(color = case_when((region == "centralca") ~ "#ffff99", 
                             (region == "norca") ~ "#beaed4",
                             (region == "socal") ~ "#fdc086",
                             (region == "channel") ~ "#7fc97f")) %>%
  select(!!sym(sumstat), NAME, degy, period, region, color) %>%
  pivot_wider(names_from = period, values_from = sumstat) %>%
  arrange(degy)
  
   

  #making matrix numeric for heatmap to work
  matrix_numeric <- matrix %>%
  arrange(degy) %>%
  select(-NAME, -region, -degy, -color) %>%
  select(historic, midcen, endcen) %>%
  as.matrix()

  row.names(matrix_numeric) <- matrix$NAME
  
  #jpeg(file=here(paste("./figs/heatmap/heatmap",sumstat,".png", 
                       #sep = "_")))
  
  heatmap.2(matrix_numeric, Rowv = FALSE, Colv = FALSE, dendrogram = "none", 
          main = title, tracecol=NA, revC= TRUE,
          margins = c(3,8), col= color_palette, srtCol = 360, 
          labRow = matrix$NAME,cexRow = 0.5, cexCol = 1, keysize = 0.95, 
          key.title = sumstat, key.xlab = NA, key.ylab = NA, par(cex.main=0.5)) #+
    
  #RowSideColors = matrix$color - add this within above - but it is inverted need to fix.
    
    #legend("left",      
    #legend = c("Central CA", "Northern CA", "Southern CA", "Channel Isl"),
    #col = unique(matrix$color), 
    #lty= 1,             
    #lwd = 5,           
    #cex=.7
    #)
  
  #ggsave(here(paste("./figs/heatmap/heatmapTEST.png")), plot)
  
  #dev.off()

}
```

```{r}
#| label: plot_heatmap

#Temp heatmaps
make_heatmap("Temp_mean", col, title = "Temp (C) mean across periods" )
make_heatmap("Temp_high10", col, title = "Temp highest 10% across periods")
make_heatmap("Temp_low10", col, title = "Temp lowest 10% across periods")
make_heatmap("T_eventSD", col, title = "Temp event SD across periods")
make_heatmap("T_seasonalSD", col, title = "Temp seasonal SD across periods")

#pH heatmaps
make_heatmap("pH_mean", title = "pH mean across periods")
make_heatmap("pH_high10",title = "pH highest 10% across periods" )
make_heatmap("pH_low10", title = "pH lowest 10% across periods")
make_heatmap("pH_eventSD",col, title = "pH event SD across periods")
make_heatmap("pH_seasonalSD",col, title = "pH seasonal SD across periods")

#DO heatmaps
make_heatmap("DO_mean", title = "DO mean across periods")
make_heatmap("DO_high10",title = "DO highest 10% across periods" )
make_heatmap("DO_low10", title = "DO lowest 10% across periods")
make_heatmap("DO_eventSD",col, title = "DO event SD across periods")
make_heatmap("DO_seasonalSD",col, title = "DO seasonal SD across periods")

```

## Regressions

Supplements heatmaps by looking at the correlation between historic and future time periods for a given summary stat. Each point is an MPA, and values for given summary stat within specified time periods are on the axes.

```{r}
#| label: regression_function

make_regression <- function(sumstat, period1, period2,title){
 plot_matrix <- sum %>%
   filter(period == period1 | period == period2) %>%
   select(sumstat, period, NAME, File) %>%
   pivot_wider(names_from = period, values_from = sumstat) 
 
 plot <- ggplot(data = plot_matrix, aes_string(x = period1, y = period2)) +
   stat_poly_line() +
   stat_poly_eq(use_label(c("eq", "R2"))) +
   geom_point() +
   theme_classic() +
   ggtitle(title)
 
  ggsave(here(paste("./figs/regression/regression",sumstat,period2,".png")), plot)

}
```

```{r}
#| label: plot_regression

#temp
    make_regression("Temp_mean", "historic", "midcen", "Historic vs. mid century temp mean")
    make_regression("Temp_mean", "historic", "endcen", "Historic vs. end century temp mean")
    make_regression("T_seasonalSD", "historic", "midcen", "Historic vs. mid century temp seasonal SD")
    make_regression("T_seasonalSD", "historic", "endcen", "Historic vs. end century temp seasonal SD")
    make_regression("T_eventSD", "historic", "midcen", "Historic vs. mid century temp event SD")
    make_regression("T_eventSD", "historic", "endcen", "Historic vs. end century temp event SD")
    
    #pH
    make_regression("pH_mean", "historic", "midcen", "Historic vs. mid century pH mean")
    make_regression("pH_mean", "historic", "endcen", "Historic vs. end century pH mean")
    make_regression("pH_seasonalSD", "historic", "midcen", "Historic vs. mid century pH seasonal SD")
    make_regression("pH_seasonalSD", "historic", "endcen", "Historic vs. end century pH seasonal SD")
    make_regression("pH_eventSD", "historic", "midcen", "Historic vs. mid century pH event SD")
    make_regression("pH_eventSD", "historic", "endcen", "Historic vs. end century pH event SD")
    
    #DO
    make_regression("DO_mean", "historic", "midcen", "Historic vs. mid century DO mean")
    make_regression("DO_mean", "historic", "endcen", "Historic vs. end century DO mean")
    make_regression("DO_seasonalSD", "historic", "midcen", "Historic vs. mid century DO seasonal SD")
    make_regression("DO_seasonalSD", "historic", "endcen", "Historic vs. end century DO seasonal SD")
    make_regression("DO_eventSD", "historic", "midcen", "Historic vs. mid century DO event SD")
    make_regression("DO_eventSD", "historic", "endcen", "Historic vs. end century DO event SD")

```

## Anomalous pH/DO/temp Event Analysis

Identify pH/DO/temp anomalous events. Events for pH are defined as 1+ days where pH \< 7.75, and events for DO are 1+ days where DO \< 4.6. Events for temp defined as 1+ days where daily temp exceeds historical climatology's temp for that month + 2\* historical climatology SD.

For each [event]{.underline}, the following are calculated:

-   Duration of days

-   Date event began

-   Mean pH/DO/T during the event

-   Intensity of the event (threshold - event mean)

-   Severity of the event (intensity\*duration).

For each [MPA]{.underline}, the following are calculated:

-   Number of events

-   Mean event duration

-   Mean of average event pH/DO/T

-   Mean event intensity

-   Mean event severity

-   Average number of events within a year

-   Average annual pH/DO/T during event

```{r}
#| label: label_events
# Create labels for events (e.g., consecutive days with low pH or low DO) using run 
# length encoding

label_events <- function(is_event) {
  event_rle <- rle(is_event)
  labels <- ifelse(event_rle$values, cumsum(event_rle$values), NA)
  rep(labels, event_rle$lengths)
}

#all event categorization for pH and DO
events <- mpa %>%
  group_by(period) %>%
  select(File, Date, Year,NAME, pH, DO, period, degx, degy, julianday) %>%
  mutate(is_pH_low = pH < 7.75,
         pH_event = label_events(is_pH_low),
        
         is_DO_low = DO < 4.6,
         DO_event = label_events(is_DO_low),
         
         is_pH_and_DO_low = ifelse(is_pH_low == TRUE & 
                                     is_DO_low == TRUE, TRUE, FALSE),
         pH_and_DO_event = label_events(is_pH_and_DO_low)) 
```

```{r}
#| label: pH_event_summary

#pH event summary
pH_event_summary <- events %>%
  filter(is_pH_low == TRUE) %>% 
  select(-DO_event, -DO, -is_DO_low, -is_pH_and_DO_low, -pH_and_DO_event) %>%
  group_by(File, pH_event) %>% 
  mutate(duration_days = n(),
         event_begin = min(Date),
         event_mean = mean(pH), #mean during the individual event 
         intensity = 7.75 - event_mean,
         severity = intensity*duration_days) %>%
  ungroup() %>%

  #mpa summary (across periods)
  group_by(File, period) %>%
  mutate(num_event = n_distinct(pH_event), 
         mean_event_duration = mean(duration_days),  
         max_event_duration = max(duration_days),
         mean_event_mean_pH = mean(event_mean), 
         mean_event_intensity = mean(intensity),
         mean_event_severity = mean(severity)) %>%
  ungroup() %>%

  #mpa summary (annual)
  group_by(File, Year, period) %>%
  mutate(annual_days_belowthresh = n(), annual_avg_ph_belowthresh = mean(pH)) %>%
  ungroup() %>%
  group_by(File, period) %>%
  mutate(meanannual_days_belowthresh = mean(annual_days_belowthresh),
         meanannual_avg_pH_belowthresh = mean(annual_avg_ph_belowthresh)) 
  
#now removing event-scale variables not needed to condense to mpa scale
pH_mpa_summary <- pH_event_summary %>%
  select(-duration_days, -event_begin, -event_mean, -intensity, 
         -severity, -is_pH_low,-pH_event, -julianday) %>%
  distinct(period, File, .keep_all = TRUE) %>%
  #renaming vars just for maps in next section
  rename("Mean num of events in a year" = "meanannual_days_belowthresh",
         "Mean pH during event"="meanannual_avg_pH_belowthresh")

#used later
pH_all_events <- left_join(events, pH_event_summary,
                         by = c("File", "Date", "Year", "NAME", "pH", 
                                "period", "degx", "degy", "julianday",
                                "is_pH_low", "pH_event"))
```

```{r}
#| label: DO_event_summary


#DO event summary
DO_event_summary <- events %>%
  filter(is_DO_low == TRUE) %>% 
  select(-pH_event, -pH, -is_pH_low, -is_pH_and_DO_low, -pH_and_DO_event) %>%
  group_by(File, DO_event) %>% 
  mutate(duration_days = n(),
         event_begin = min(Date),
         event_mean = mean(DO), #mean during the individual event 
         intensity = 7.75 - event_mean,
         severity = intensity*duration_days) %>%
  ungroup() %>%

  #mpa summary (across periods)
  group_by(File, period) %>%
  mutate(num_event = n_distinct(DO_event), 
         mean_event_duration = mean(duration_days),  
         max_event_duration = max(duration_days),
         mean_event_mean_pH = mean(event_mean), 
         mean_event_intensity = mean(intensity),
         mean_event_severity = mean(severity)) %>%
  ungroup() %>%

  #mpa summary (annual)
  group_by(File, Year, period) %>%
  mutate(annual_days_belowthresh = n(), annual_avg_DO_belowthresh = mean(DO)) %>%
  ungroup() %>%
  group_by(File, period) %>%
  mutate(meanannual_days_belowthresh = mean(annual_days_belowthresh),
         meanannual_avg_DO_belowthresh = mean(annual_avg_DO_belowthresh)) 
  
#now removing event-scale variables not needed to condense to mpa scale
DO_mpa_summary <- DO_event_summary %>% 
  select(-duration_days, -event_begin, -event_mean, -intensity, 
         -severity, -is_DO_low, -DO_event, -julianday) %>%
   distinct(File, period, .keep_all = TRUE) %>%
  #renaming vars just for maps in next section
  rename("Mean num of events in a year" = "meanannual_days_belowthresh",
         "Mean DO during event"= "meanannual_avg_DO_belowthresh")

#used later
DO_all_events <- left_join(events, DO_event_summary,
                         by = c("File", "Date", "Year", "NAME", "DO", 
                                "period", "degx", "degy", "julianday",
                                "is_DO_low", "DO_event"))
```

```{r}
#| label: temp_event_summary

#temp:
mpa_with_histclimsd <- mpa %>%
  filter(period == "historic") %>%
  group_by(File, Month) %>%
  summarise(hist_T_clim = mean(Temp), hist_T_clim_sd = sd(Temp)) %>% #this sd is temp over all days in Jan - rather than average of each jan's SD - does it make a difference? 
  merge(mpa, by = c("File", "Month"))

temp_event_summary <- mpa_with_histclimsd %>% 
  group_by(period) %>%
  select(File, Date, Year,NAME, Temp, hist_T_clim, hist_T_clim_sd, period, 
         degx, degy, julianday) %>%
  mutate(is_temp_high = Temp > (hist_T_clim + 2*hist_T_clim_sd),
         temp_event = label_events(is_temp_high)) %>%
  
  filter(is_temp_high == TRUE) %>%
  group_by(File, temp_event, period) %>% 
  mutate(duration_days = n(),
         event_begin = min(Date),
         event_mean = mean(Temp),  
         intensity = Temp - (hist_T_clim + 2*hist_T_clim_sd),
         severity = intensity*duration_days) %>%
  ungroup() %>%

  #mpa summary (across periods)
  group_by(File, period) %>%
  mutate(num_event = n_distinct(temp_event), 
         mean_event_duration = mean(duration_days),  
         max_event_duration = max(duration_days),
         mean_event_mean_temp = mean(event_mean), 
         mean_event_intensity = mean(intensity),
         mean_event_severity = mean(severity)) %>%
  ungroup() %>%
  
  #mpa summary (annual)
  group_by(File, Year, period) %>%
  mutate(annual_days_abovethresh = n(), 
         annual_avg_temp_abovethresh = mean(Temp)) %>%
  ungroup() %>%
  group_by(File, period) %>%
  mutate(meanannual_days_abovethresh = mean(annual_days_abovethresh),
         meanannual_avg_temp_abovethresh = mean(annual_avg_temp_abovethresh)) 
  
  #now removing event-scale variables not needed to condense to mpa scale
temp_mpa_summary <- temp_event_summary %>% 
  select(-Temp,-duration_days, -event_begin, -event_mean, -intensity, 
         -severity, -is_temp_high, -temp_event, -julianday) %>%
  distinct(period, File, .keep_all = TRUE) %>%
  #renaming vars just for maps in next section
  rename("Mean num of events in a year" = "meanannual_days_abovethresh",
         "Mean temperature during event"="meanannual_avg_temp_abovethresh")

#used later
temp_all_events <- left_join(mpa, temp_event_summary,
                         by = c("File", "Date", "Year", "NAME", "Temp", "period", 
                                "degx", "degy", "julianday"))
```

## Map of anomalous events 

Maps showing anomalous event summaries in CA, including average number days crossed threshold in a year, and average value when threshold is crossed. Same color scheme fixed across time periods, using equal breaks classification of values across all periods.

```{r}
#| label: anom_map_function


CA_shp <- st_read(here("./data/rawdata/shp/CA_Counties/CA_Counties_TIGER2016.shp"))


 make_anom_map <- function(mpa_summary_df, sumstat, periodt, title, 
                           filename, palette = "YlOrRd"){
    
    anom_points <- st_as_sf((mpa_summary_df %>% filter(period == periodt)), 
                            coords = c("degx","degy"))

    #finding equal interval breaks based on all time periods
    all_breaks <- classInt::classIntervals(mpa_summary_df[[sumstat]], 
                                           8, "equal")
    
    #map with equal breaks across periods
   map <- tm_shape(CA_shp) + #basemap
      tm_fill(col = "#ccebc5") +
      
      tm_shape(anom_points) +
      tm_dots(col = sumstat, palette = palette, size = 0.2, 
              breaks = all_breaks[["brks"]]) +
      
      tm_layout(bg.color = "#a6cee3", 
                main.title.position = "center",
                main.title.size = 1.3,
                main.title = title,
                legend.title.size = 1,
                legend.text.size = .8) 
    
tmap_save(map, here(paste("./figs/anom_map/",filename)))
    
 }
 
```

```{r}
#| label: plot_anom_map

#TEMP-----------------------------------------------------
#mean annual num events for all periods for temp 
make_anom_map(temp_mpa_summary, sumstat = "Mean num of events in a year", 
                periodt =  "historic", 
                title = "Annual # of anomalous heat events in historic period",
                filename = "temp_historic_annual_event.png")

make_anom_map(temp_mpa_summary, sumstat = "Mean num of events in a year",
              periodt = "midcen",
              title = "Annual # of anomalous heat events in mid century period",
              filename = "temp_midcen_annual_event.png") 

make_anom_map(temp_mpa_summary, sumstat = "Mean num of events in a year",
              periodt = "endcen",
              title = "Annual # of anomalous heat events in end century period",
              filename = "temp_endcen_annual_event.png")

#mean temp during event for all periods
make_anom_map(temp_mpa_summary, "Mean temperature during event", "historic", 
              "Mean heat event temperature in historic period",
              "temp_val_historic_event.png")

make_anom_map(temp_mpa_summary, "Mean temperature during event","midcen",
              "Mean heat event temperature in mid century period",
              "temp_val_midcen_event.png") 

make_anom_map(temp_mpa_summary, "Mean temperature during event",
              "endcen","Mean heat event temperature in end century period",
              "temp_val_endcen_event.png")

#pH---------------------------------------------------------
make_anom_map(pH_mpa_summary, "Mean num of events in a year", "historic",
              "Annual # of anomalous pH events in historic period",
              "pH_historic_annual_event.png")

make_anom_map(pH_mpa_summary, "Mean num of events in a year","midcen",
              "Annual # of anomalous pH events in mid century period",
              "pH_midcen_annual_event.png") 

make_anom_map(pH_mpa_summary, "Mean num of events in a year","endcen",
              "Annual # of anomalous pH events in end century period",
              "pH_endcen_annual_event.png")

#mean pH during event for all periods 
make_anom_map(pH_mpa_summary, "Mean pH during event", "historic", 
              "Mean event pH in historic period","pH_val_historic_event.png",
              palette = "-YlOrRd")

make_anom_map(pH_mpa_summary, "Mean pH during event","midcen",
              "Mean event pH in mid century period","pH_val_midcen_event.png",
              palette = "-YlOrRd") 

make_anom_map(pH_mpa_summary, "Mean pH during event","endcen",
              "Mean event pH in end century period","pH_val_endcen_event.png",
              palette = "-YlOrRd")

#DO---------------------------------------------------------
make_anom_map(DO_mpa_summary, "Mean num of events in a year", "historic",
              "Annual # of anomalous DO events in historic period",
              "DO_historic_annual_event.png")

make_anom_map(DO_mpa_summary, "Mean num of events in a year","midcen",
              "Annual # of anomalous DO events in mid century period",
              "DO_midcen_annual_event.png") 

make_anom_map(DO_mpa_summary, "Mean num of events in a year","endcen",
              "Annual # of anomalous DO events in end century period",
              "DO_endcen_annual_event.png")

#mean DO during event for all periods 
make_anom_map(DO_mpa_summary, "Mean DO during event", "historic", 
              "Mean event DO in historic period","DO_val_historic_event.png",
              palette = "-YlOrRd")

make_anom_map(DO_mpa_summary, "Mean DO during event","midcen",
              "Mean event DO in mid century period","DO_val_midcen_event.png",
              palette = "-YlOrRd") 

make_anom_map(DO_mpa_summary, "Mean DO during event","endcen",
              "Mean event DO in end century period","DO_val_endcen_event.png",
              palette = "-YlOrRd")

```

## Severity, Intensity, Duration

Create figures displaying occurrences of events throughout the year, to determine monthly occurrences. Julian days on x-axis, years on y-axis, and events are shaded by intensity (threshold - event mean).

```{r}
#| label: temp_sev_int_dur_function

#temp
temp_sev_int_dur <- function(periodt, site){
  
  plot <- temp_all_events %>%
    filter(period == periodt, NAME == site) %>%
    ggplot(aes(julianday, Year, fill = intensity)) +
    geom_tile() +
    scale_fill_distiller(palette = "YlOrRd", direction = 1, 
                         na.value = "skyblue") +
    theme_classic() +
    ggtitle(paste(periodt, "Heat events at", site)) 
    
ggsave(here(paste("./figs/sevintdur/","temp_events", periodt, site,".png",sep="_")), plot)

}

#central CA
temp_sev_int_dur("historic", "Ano Nuevo SMR")
temp_sev_int_dur("midcen", "Ano Nuevo SMR")
temp_sev_int_dur("endcen", "Ano Nuevo SMR")

#northern CA
temp_sev_int_dur("historic", "Point Arena SMR")
temp_sev_int_dur("midcen", "Point Arena SMR")
temp_sev_int_dur("endcen", "Point Arena SMR")

#southern CA
temp_sev_int_dur("historic", "South La Jolla SMR")
temp_sev_int_dur("midcen", "South La Jolla SMR")
temp_sev_int_dur("endcen", "South La Jolla SMR")

#channel
temp_sev_int_dur("historic", "Anacapa Island FMR")
temp_sev_int_dur("midcen", "Anacapa Island FMR")
temp_sev_int_dur("endcen", "Anacapa Island FMR")
```

```{r}
#pH
pH_sev_int_dur <- function(periodt, site){
  
  plot <- pH_all_events %>%
    filter(period == periodt, NAME == site) %>%
    ggplot(aes(julianday, Year, fill = intensity)) +
    geom_tile() +
    scale_fill_distiller(palette = "YlOrRd", direction = 1, 
                         na.value = "skyblue") +
    theme_classic() +
    ggtitle(paste(periodt, "pH events at", site)) 
    
ggsave(here(paste("./figs/sevintdur/","pH_events", periodt, site,".png",sep="_")), plot)

}




#central CA
pH_sev_int_dur("historic", "Ano Nuevo SMR")
pH_sev_int_dur("midcen", "Ano Nuevo SMR")
pH_sev_int_dur("endcen", "Ano Nuevo SMR")

#northern CA
pH_sev_int_dur("historic", "Point Arena SMR")
pH_sev_int_dur("midcen", "Point Arena SMR")
pH_sev_int_dur("endcen", "Point Arena SMR")

#southern CA
pH_sev_int_dur("historic", "South La Jolla SMR")
pH_sev_int_dur("midcen", "South La Jolla SMR")
pH_sev_int_dur("endcen", "South La Jolla SMR")

#channel
pH_sev_int_dur("historic", "Anacapa Island FMR")
pH_sev_int_dur("midcen", "Anacapa Island FMR")
pH_sev_int_dur("endcen", "Anacapa Island FMR")
```

```{r}
#DO
DO_sev_int_dur <- function(periodt, site){
  
  plot <- DO_all_events %>%
    filter(period == periodt, NAME == site) %>%
    ggplot(aes(julianday, Year, fill = intensity)) +
    geom_tile() +
    scale_fill_distiller(palette = "YlOrRd", direction = 1, 
                         na.value = "skyblue") +
    theme_classic() +
    ggtitle(paste(periodt, "DO events at", site)) 
    
ggsave(here(paste("./figs/sevintdur/","DO_events", periodt, site,".png",sep="_")), plot)

}


#central CA
DO_sev_int_dur("historic", "Ano Nuevo SMR")
DO_sev_int_dur("midcen", "Ano Nuevo SMR")
DO_sev_int_dur("endcen", "Ano Nuevo SMR")

#northern CA
DO_sev_int_dur("historic", "Point Arena SMR")
DO_sev_int_dur("midcen", "Point Arena SMR")
DO_sev_int_dur("endcen", "Point Arena SMR")

#southern CA
DO_sev_int_dur("historic", "South La Jolla SMR")
DO_sev_int_dur("midcen", "South La Jolla SMR")
DO_sev_int_dur("endcen", "South La Jolla SMR")

#channel
DO_sev_int_dur("historic", "Anacapa Island FMR")
DO_sev_int_dur("midcen", "Anacapa Island FMR")
DO_sev_int_dur("endcen", "Anacapa Island FMR")
```

## Time series to put in context

```{r}

#this fig shows time series in julian days w/ envelope wrt given period's SD (NOT historic, as it should be for calculating anomalous event!)

make_temp_time_series <- function(periodt, site, title){

mpa %>% 
  filter(period == periodt, 
         NAME == site) %>%
  mutate(sliding_sd = roll_sd(Temp, 30, fill = NA)) %>% 
  group_by(julianday) %>%
  mutate(clim_mean = mean(Temp, na.rm = TRUE),
         clim_sd = mean(sliding_sd, na.rm = TRUE),
         clim_upr = clim_mean + 2 * clim_sd,
         clim_lwr = clim_mean - 2 * clim_sd) %>% 
  ungroup() %>%
  
  ggplot(aes(julianday)) +
  geom_ribbon(aes(ymin = clim_lwr, ymax = clim_upr),
              fill = "black", alpha = 0.75) +
  geom_line(aes(y = Temp, color = Year, group = Year),
            alpha = 0.5) +
  geom_line(aes(y = clim_mean), size = 1, color = "blue") +
  scale_color_viridis_c() +
  theme_classic() +
    ggtitle(title)
}

make_pH_time_series <- function(df, periodt, site, title){

df %>% 
  filter(period == periodt, 
         NAME == site) %>%
  mutate(sliding_sd = roll_sd(pH, 30, fill = NA)) %>% 
  group_by(julianday) %>%
  mutate(clim_mean = mean(pH, na.rm = TRUE),
         clim_sd = mean(sliding_sd, na.rm = TRUE),
         clim_upr = clim_mean + 2 * clim_sd,
         clim_lwr = clim_mean - 2 * clim_sd) %>% 
  ungroup() %>%
  
  ggplot(aes(julianday)) +
  geom_ribbon(aes(ymin = clim_lwr, ymax = clim_upr),
              fill = "black", alpha = 0.75) +
  geom_line(aes(y = pH, color = Year, group = Year),
            alpha = 0.5) +
  geom_line(aes(y = clim_mean), size = 1, color = "blue") +
  scale_color_viridis_c() +
  theme_classic() +
    ggtitle(title)
}

make_DO_time_series <- function(df, periodt, site, title){

df %>% 
  filter(period == periodt, 
         NAME == site) %>%
  mutate(sliding_sd = roll_sd(DO, 30, fill = NA)) %>% 
  group_by(julianday) %>%
  mutate(clim_mean = mean(DO, na.rm = TRUE),
         clim_sd = mean(sliding_sd, na.rm = TRUE),
         clim_upr = clim_mean + 2 * clim_sd,
         clim_lwr = clim_mean - 2 * clim_sd) %>% 
  ungroup() %>%
  
  ggplot(aes(julianday)) +
  geom_ribbon(aes(ymin = clim_lwr, ymax = clim_upr),
              fill = "black", alpha = 0.75) +
  geom_line(aes(y = DO, color = Year, group = Year),
            alpha = 0.5) +
  geom_line(aes(y = clim_mean), size = 1, color = "blue") +
  scale_color_viridis_c() +
  theme_classic() +
    ggtitle(title)
  
}



make_temp_time_series("historic","Ano Nuevo SMR", title = "Historic Temp at Ano Nuevo") 
make_temp_time_series("midcen","Ano Nuevo SMR", title = "Midcen Temp at Ano Nuevo") 
make_temp_time_series("endcen","Ano Nuevo SMR", title = "Endcen Temp at Ano Nuevo") 

make_pH_time_series(mpa, "historic","Ano Nuevo SMR", title = "Historic pH at Ano Nuevo") 
make_pH_time_series("midcen","Ano Nuevo SMR", title = "Midcen pH at Ano Nuevo") 
make_pH_time_series("endcen","Ano Nuevo SMR", title = "Endcen pH at Ano Nuevo") 

make_DO_time_series(mpa, "historic","Ano Nuevo SMR", title = "Historic DO at Ano Nuevo") 
make_DO_time_series("midcen","Ano Nuevo SMR", title = "Midcen DO at Ano Nuevo") 
make_DO_time_series("endcen","Ano Nuevo SMR", title = "Endcen DO at Ano Nuevo")



```

Making figs (severity/intensity/duration and time series for individual MPAs

```{r}

```
