---
title: "MPA Exposure Analysis"
format: 
  html:
    code-fold: true
editor: visual
---

```{r}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

------------------------------------------------------------------------

Code to characterize exposure of California MPAs to future pH, DO, and temperature using IPSL model predictions. Note: this setup is only for one projection (IPSL). Change every occurence of "IPSL" to "GFDL" or "HADLEY" in this report to do same for diff projection.

## Setup

Load libraries and read data. Convert DO units and filter for years 2090-2100.

```{r}
#| label: libraries
#| message: false

library(tidyverse)
library(lubridate)
library(data.table)
library(factoextra)
library(broom)
library(cowplot)
library(respR)
library(here)
library(lattice)
library(RcppRoll)
library(RColorBrewer)
library(gplots)
library(ggpmisc)
```

```{r}
#| label: load_data
#| message: false

#Read mpa data, convert DO units, add a column for time period, and one for difference between mid/end cen and historic.
mpa <- read_csv(here("data/processeddata/model/IPSLmpa.csv")) %>% 
  mutate(DO_mmolL = DO_surf/1000,
         DO_mgL = convert_DO(DO_mmolL, from = "mmol/L", to = "mg/L")) %>% 
  filter(Year <= 2020|
         Year >= 2040 & Year <= 2060|
         Year >= 2080 & Year <= 2100) %>%
  mutate(period = case_when((Year <= 2020) ~ "historic", 
                           (Year %in% c(2040:2060)) ~ "midcen",
                           (Year %in% c(2080:2100)) ~ "endcen")) %>%
  select(-T_bot, -DO_bot, -pH_bot,-DO_surf,-DO_mmolL) 
mpa$File <- substr(mpa$File, 1, nchar(mpa$File)-4)


```

```{r}
#| label: regions
#| message: false
#add regions here!
mpa_centroids<- read_csv(here("data/rawdata/MPA_polygons.csv"))
mpa_centroids$File <- sub("^", "tphdo_mpa_", mpa_centroids$OBJECTID )

mpa_centroids <- mpa_centroids %>%
  mutate(region = ifelse(degy >= 37.29, "norca", 
                         ifelse(degy > 34.8, "centralca", 
                                ifelse("degy" < 34.274 & "degx" < -119.220, "channelisl" , 
                                       "socal"))))

channel <- c("Anacapa Island FMCA", "Anacapa Island FMR", "Anacapa Island SMCA", "Anacapa Island SMR", "Anacapa Island Special Closure", 
             "Arrow Point to Lion Head Point SMCA", "Begg Rock SMR", "Blue Cavern Offshore SMCA", "Blue Cavern Onshore SMCA (No-Take)", 
             "Carrington Point SMR", "Casino Point SMCA (No-Take)", "Cat Harbor SMCA", "Farnsworth Offshore SMCA", "Farnsworth Onshore SMCA", 
             "Footprint FMR", "Footprint SMR", "Gull Island FMR", "Gull Island SMR", "Harris Point FMR", "Harris Point SMR", "Judith Rock SMR", 
             "Long Point SMR", "Lover’s Cove SMCA", "Painted Cave SMCA", "Richardson Rock FMR", "Richardson Rock SMR", "San Miguel Island Special Closure", 
             "Santa Barbara Island FMR", "Santa Barbara Island SMR", "Scorpion FMR", "Scorpion SMR", "Skunk Point SMR", "South Point FMR", "South Point SMR")

mpa_centroids$region[mpa_centroids$NAME %in% channel]  <- "channel" 
mpa <- merge(mpa, mpa_centroids,  by = "File")

mpa_historic <- mpa %>%
  filter(period == "historic")
mpa_midcen <- mpa %>%
  filter(period == "midcen")
mpa_endcen <- mpa %>%
  filter(period == "endcen")
```

## Create summary stats

Find mean and lower 10th percentile of pH, temp, and DO for each MPA

```{r}
#| label: summary_stats

sum_historic <- mpa_historic %>% 
  group_by(File) %>% 
  summarize(across(c(T_surf, DO_mgL, pH_surf), 
                   list(mean = mean, quantile = ~ quantile(.x, 0.1)))) %>% 
  rename(T_mean = T_surf_mean, 
         DO_mean = DO_mgL_mean, 
         pH_mean = pH_surf_mean, 
         T_low10 = T_surf_quantile, 
         DO_low10 = DO_mgL_quantile, 
         pH_low10 = pH_surf_quantile) 
sum_historic <- merge(sum_historic, mpa_centroids,  by = "File")


sum_midcen <- mpa_midcen %>% 
  group_by(File) %>% 
  summarize(across(c(T_surf, DO_mgL, pH_surf), 
                   list(mean = mean, quantile = ~ quantile(.x, 0.1)))) %>% 
  rename(T_mean = T_surf_mean, 
         DO_mean = DO_mgL_mean, 
         pH_mean = pH_surf_mean, 
         T_low10 = T_surf_quantile, 
         DO_low10 = DO_mgL_quantile, 
         pH_low10 = pH_surf_quantile)
sum_midcen <- merge(sum_midcen, mpa_centroids,  by = "File")


sum_endcen <- mpa_endcen %>% 
  group_by(File) %>% 
  summarize(across(c(T_surf, DO_mgL, pH_surf), 
                   list(mean = mean, quantile = ~ quantile(.x, 0.1)))) %>% 
  rename(T_mean = T_surf_mean, 
         DO_mean = DO_mgL_mean, 
         pH_mean = pH_surf_mean, 
         T_low10 = T_surf_quantile, 
         DO_low10 = DO_mgL_quantile, 
         pH_low10 = pH_surf_quantile)
sum_endcen <- merge(sum_endcen, mpa_centroids,  by = "File")
```

## Calculate seasonal variation

Create a climatology (Jan averaged over every year), so each MPA has one value per month that is averaged over all of the years. Find SD of this dataframe, to get seasonal variation.

```{r}
#| label: clim_SD
#| message: false

mpa_historic_climatology <- mpa_historic %>%
  group_by(File, Month) %>%
  summarise(T_clim = mean(T_surf), T_clim_sd = sd(T_surf), pH_clim = mean(pH_surf), DO_clim =   mean(DO_mgL))
mpa_midcen_climatology <- mpa_midcen %>%
  group_by(File, Month) %>%
  summarise(T_clim = mean(T_surf), pH_clim = mean(pH_surf), DO_clim = mean(DO_mgL))
mpa_endcen_climatology <- mpa_endcen %>%
  group_by(File, Month) %>%
  summarise(T_clim = mean(T_surf), pH_clim = mean(pH_surf), DO_clim = mean(DO_mgL))

historic_seasonalSD <- mpa_historic_climatology %>%
  group_by(File) %>%
  summarise(T_seasonalSD = sd(T_clim),
            pH_seasonalSD = sd(pH_clim),
            DO_seasonalSD = sd(DO_clim))
midcen_seasonalSD <- mpa_midcen_climatology %>%
  group_by(File) %>%
  summarise(T_seasonalSD = sd(T_clim),
            pH_seasonalSD = sd(pH_clim),
            DO_seasonalSD = sd(DO_clim))
endcen_seasonalSD <- mpa_endcen_climatology %>%
  group_by(File) %>%
  summarise(T_seasonalSD = sd(T_clim),
            pH_seasonalSD = sd(pH_clim),
            DO_seasonalSD = sd(DO_clim))
```

## Calculate event-based variation

Create a mock dataset interpolating climatologies to get daily values in the absence of natural event-based variability. Subtract actual daily values. Find standard deviation of these differences (on a daily scale).

Setup for interpolation:

```{r}
#| label: inter_setup
#| message: false
#| warning: false


#First: Need to create a day1 and day365 proxies. Approx function can only interpolate not extrapolate so without this, you can interpolate days 1-14 and 350-365. #below creates empty vectors as big as we need (365 days per mpa) for each variable and each mpa. julianday is dates 1-365 as many mpa times

mpaslist = unique(mpa$File) 
mpas = rep(NA, 365*121)
julianday = rep(1:365, 121)
T_clim = rep(NA, 365*121)
pH_clim = rep(NA, 365*121)
DO_clim = rep(NA, 365*121)

#Set up a vector of julian day assignment for the 15th of each month and the first and last day of the year
x_in <- yday(as.Date(c("2000-01-01", "2000-01-15","2000-02-15","2000-03-15","2000-04-15","2000-05-15","2000-06-15",
                       "2000-07-15","2000-08-15","2000-09-15","2000-10-15","2000-11-15","2000-12-15", "2000-12-31")))

# creating a list of all the days of the year not included in x_in to interpolate to.
x_out <- (1:365)
x_out <- x_out[!(x_out %in% x_in)] #removing the days of the year we already have values for
```

Interpolation for temp:

```{r,warning=FALSE, message=FALSE, results='hide'}
#| label: inter_temp
#| message: false
#| warning: false

#HISTORIC
for (i in 1:length(mpaslist)){
  print(i)
  d = mpa_historic_climatology %>% filter(File == mpaslist[i]) %>% select(T_clim)
  #use a weighted average to get these
  Dec31 = as.numeric(( ((16/30) * (d[12,2])) + ((14/30) * d[1,2]) ))
  Jan1 = as.numeric(( ((14/30) * (d[12,2])) + ((16/30) * d[1,2]) ))
  # a list of y-values of the climatological temp on each of the days in x_in
  y_in <- c(Jan1, d$T_clim, Dec31)

  mod = approx(x = x_in, y = y_in, xout = x_out) 
  mpas[((i-1)*365+1):(i*365)] <- mpaslist[i] #rep(mpaslist[1], 365)
  T_clim[((i-1)*365+1):(i*365)] <- mod$y
}

historic_tempdata = data.frame(mpas, julianday,T_clim)
historic_tempdata <- historic_tempdata %>%
  rename(File = mpas)

#Later we create a julian day variable in our OG mpa dataset, to merge this to all the rows on the dataset by filename and day and then have the mean daily climatological value for each row. 
#MIDCEN
for (i in 1:length(mpaslist)){
  print(i)
  d = mpa_midcen_climatology %>% filter(File == mpaslist[i]) %>% select(T_clim)
  #use a weighted average to get these
  Dec31 = as.numeric(( ((16/30) * (d[12,2])) + ((14/30) * d[1,2]) ))
  Jan1 = as.numeric(( ((14/30) * (d[12,2])) + ((16/30) * d[1,2]) ))
  # a list of y-values of the climatological temp on each of the days in x_in
  y_in <- c(Jan1, d$T_clim, Dec31)

  mod = approx(x = x_in, y = y_in, xout = x_out) 
  mpas[((i-1)*365+1):(i*365)] <- mpaslist[i] #rep(mpaslist[1], 365)
  T_clim[((i-1)*365+1):(i*365)] <- mod$y
}

midcen_tempdata = data.frame(mpas, julianday,T_clim)
midcen_tempdata <- midcen_tempdata %>%
  rename(File = mpas)

#ENDCEN
for (i in 1:length(mpaslist)){
  print(i)
  d = mpa_endcen_climatology %>% filter(File == mpaslist[i]) %>% select(T_clim)
  #use a weighted average to get these
  Dec31 = as.numeric(( ((16/30) * (d[12,2])) + ((14/30) * d[1,2]) ))
  Jan1 = as.numeric(( ((14/30) * (d[12,2])) + ((16/30) * d[1,2]) ))
  # a list of y-values of the climatological temp on each of the days in x_in
  y_in <- c(Jan1, d$T_clim, Dec31)

  mod = approx(x = x_in, y = y_in, xout = x_out) 
  mpas[((i-1)*365+1):(i*365)] <- mpaslist[i] #rep(mpaslist[1], 365)
  T_clim[((i-1)*365+1):(i*365)] <- mod$y
}

endcen_tempdata = data.frame(mpas, julianday,T_clim)
endcen_tempdata <- endcen_tempdata %>%
  rename(File = mpas)
```

Interpolation for pH

```{r, warning=FALSE, message=FALSE, results='hide'}
#| label: inter_pH
#| message: false
#| warning: false

#HISTORIC
for (i in 1:length(mpaslist)){
  print(i)
  d = mpa_historic_climatology %>% filter(File == mpaslist[i]) %>% select(pH_clim)
  Dec31 = as.numeric(( ((16/30) * (d[12,2])) + ((14/30) * d[1,2]) ))
  Jan1 = as.numeric(( ((14/30) * (d[12,2])) + ((16/30) * d[1,2]) ))

  y_in <- c(Jan1, d$pH_clim, Dec31)

  mod = approx(x = x_in, y = y_in, xout = x_out) 
  mpas[((i-1)*365+1):(i*365)] <- mpaslist[i]
  pH_clim[((i-1)*365+1):(i*365)] <- mod$y
}

historic_pHdata = data.frame(mpas, julianday,pH_clim)
historic_pHdata <- historic_pHdata %>%
  rename(File = mpas)

#MIDCEN
for (i in 1:length(mpaslist)){
  print(i)
  d = mpa_midcen_climatology %>% filter(File == mpaslist[i]) %>% select(pH_clim)
  Dec31 = as.numeric(( ((16/30) * (d[12,2])) + ((14/30) * d[1,2]) ))
  Jan1 = as.numeric(( ((14/30) * (d[12,2])) + ((16/30) * d[1,2]) ))

  y_in <- c(Jan1, d$pH_clim, Dec31)

  mod = approx(x = x_in, y = y_in, xout = x_out) 
  mpas[((i-1)*365+1):(i*365)] <- mpaslist[i]
  pH_clim[((i-1)*365+1):(i*365)] <- mod$y
}

midcen_pHdata = data.frame(mpas, julianday,pH_clim)
midcen_pHdata <- midcen_pHdata %>%
  rename(File = mpas)

#ENDCEN
for (i in 1:length(mpaslist)){
  print(i)
  d = mpa_endcen_climatology %>% filter(File == mpaslist[i]) %>% select(pH_clim)
  Dec31 = as.numeric(( ((16/30) * (d[12,2])) + ((14/30) * d[1,2]) ))
  Jan1 = as.numeric(( ((14/30) * (d[12,2])) + ((16/30) * d[1,2]) ))

  y_in <- c(Jan1, d$pH_clim, Dec31)

  mod = approx(x = x_in, y = y_in, xout = x_out) 
  mpas[((i-1)*365+1):(i*365)] <- mpaslist[i]
  pH_clim[((i-1)*365+1):(i*365)] <- mod$y
}

endcen_pHdata = data.frame(mpas, julianday,pH_clim)
endcen_pHdata <- endcen_pHdata %>%
  rename(File = mpas)
```

Interpolation for DO:

```{r,warning=FALSE, message=FALSE, results='hide'}
#| label: inter_DO
#| message: false
#| warning: false

#HISTORIC
for (i in 1:length(mpaslist)){
  print(i)
  d = mpa_historic_climatology %>% filter(File == mpaslist[i]) %>% select(DO_clim)
  Dec31 = as.numeric(( ((16/30) * (d[12,2])) + ((14/30) * d[1,2]) ))
  Jan1 = as.numeric(( ((14/30) * (d[12,2])) + ((16/30) * d[1,2]) ))
  y_in <- c(Jan1, d$DO_clim, Dec31)
  
  mod = approx(x = x_in, y = y_in, xout = x_out) 
  mpas[((i-1)*365+1):(i*365)] <- mpaslist[i]
  DO_clim[((i-1)*365+1):(i*365)] <- mod$y
}

historic_DOdata = data.frame(mpas, julianday,DO_clim)
historic_DOdata <- historic_DOdata %>%
  rename(File = mpas)

#MIDCEN
for (i in 1:length(mpaslist)){
  print(i)
  d = mpa_midcen_climatology %>% filter(File == mpaslist[i]) %>% select(DO_clim)
  Dec31 = as.numeric(( ((16/30) * (d[12,2])) + ((14/30) * d[1,2]) ))
  Jan1 = as.numeric(( ((14/30) * (d[12,2])) + ((16/30) * d[1,2]) ))
  y_in <- c(Jan1, d$DO_clim, Dec31)
  
  mod = approx(x = x_in, y = y_in, xout = x_out) 
  mpas[((i-1)*365+1):(i*365)] <- mpaslist[i]
  DO_clim[((i-1)*365+1):(i*365)] <- mod$y
}

midcen_DOdata = data.frame(mpas, julianday,DO_clim)
midcen_DOdata <- midcen_DOdata %>%
  rename(File = mpas)

#ENDCEN
for (i in 1:length(mpaslist)){
  print(i)
  d = mpa_endcen_climatology %>% filter(File == mpaslist[i]) %>% select(DO_clim)
  Dec31 = as.numeric(( ((16/30) * (d[12,2])) + ((14/30) * d[1,2]) ))
  Jan1 = as.numeric(( ((14/30) * (d[12,2])) + ((16/30) * d[1,2]) ))
  y_in <- c(Jan1, d$DO_clim, Dec31)
  
  mod = approx(x = x_in, y = y_in, xout = x_out) 
  mpas[((i-1)*365+1):(i*365)] <- mpaslist[i]
  DO_clim[((i-1)*365+1):(i*365)] <- mod$y
}

endcen_DOdata = data.frame(mpas, julianday,DO_clim)
endcen_DOdata <- endcen_DOdata %>%
  rename(File = mpas)
```

Merge interpolated values to MPA dataset and find SD

```{r}
#| label: inter_merge

#adding julianday variable
mpa_historic <- mpa_historic %>%
  mutate(Date = make_date(Year, Month, Day)) %>%
  mutate(julianday = yday(Date))
mpa_midcen <- mpa_midcen %>%
  mutate(Date = make_date(Year, Month, Day)) %>%
  mutate(julianday = yday(Date))
mpa_endcen <- mpa_endcen %>%
  mutate(Date = make_date(Year, Month, Day)) %>%
  mutate(julianday = yday(Date))

#mpa merge to ph/do/temp
historic_merge <- merge(mpa_historic, historic_tempdata, by = c("File", "julianday"))
historic_merge <- merge(historic_merge, historic_DOdata, by = c("File", "julianday"))
historic_merge <- merge(historic_merge, historic_pHdata, by = c("File", "julianday"))

midcen_merge <- merge(mpa_midcen, midcen_tempdata, by = c("File", "julianday"))
midcen_merge <- merge(midcen_merge, midcen_DOdata, by = c("File", "julianday"))
midcen_merge <- merge(midcen_merge, midcen_pHdata, by = c("File", "julianday"))

endcen_merge <- merge(mpa_endcen, endcen_tempdata, by = c("File", "julianday"))
endcen_merge <- merge(endcen_merge, endcen_DOdata, by = c("File", "julianday"))
endcen_merge <- merge(endcen_merge, endcen_pHdata, by = c("File", "julianday"))

#subtract climatology . actual - clim
#filtered for only upwelling months.
historic_merge <- historic_merge %>%
  mutate(noclim_T = T_surf - T_clim,
         noclim_pH = pH_surf - pH_clim,
         noclim_DO = DO_mgL - DO_clim) %>%
  filter(Month == c(5,6,7,8,9))

midcen_merge <- midcen_merge %>%
  mutate(noclim_T = T_surf - T_clim,
         noclim_pH = pH_surf - pH_clim,
         noclim_DO = DO_mgL - DO_clim) %>%
  filter(Month == c(5,6,7,8,9))

endcen_merge <- endcen_merge %>%
  mutate(noclim_T = T_surf - T_clim,
         noclim_pH = pH_surf - pH_clim,
         noclim_DO = DO_mgL - DO_clim) %>%
  filter(Month == c(5,6,7,8,9))

#find event based SD (removed clim)
historic_eventSD <- historic_merge %>%
  group_by(File) %>%
  summarise(T_eventSD = sd(noclim_T),
            pH_eventSD = sd(noclim_pH),
            DO_eventSD = sd(noclim_DO))
midcen_eventSD <- midcen_merge %>%
  group_by(File) %>%
  summarise(T_eventSD = sd(noclim_T),
            pH_eventSD = sd(noclim_pH),
            DO_eventSD = sd(noclim_DO))
endcen_eventSD <- endcen_merge %>%
  group_by(File) %>%
  summarise(T_eventSD = sd(noclim_T),
            pH_eventSD = sd(noclim_pH),
            DO_eventSD = sd(noclim_DO))
```

## Append SD to summary stats

```{r}
#| label: append_sum

#merging eventSD and seasonalSD to mpa_sum
sum_historic <- merge(sum_historic, historic_seasonalSD,  by = "File")
sum_historic <- merge(sum_historic, historic_eventSD,  by = "File")

sum_midcen <- merge(sum_midcen, midcen_seasonalSD,  by = "File")
sum_midcen <- merge(sum_midcen, midcen_eventSD,  by = "File")

sum_endcen <- merge(sum_endcen, endcen_seasonalSD,  by = "File")
sum_endcen <- merge(sum_endcen, endcen_eventSD,  by = "File")


#create a new df that for each summary stat has: historic, (midcen-historic), (endcen-historic) - this is only really going to be used for heatmaps. so only need sum
#renaming columns


sum_diff <- merge(sum_historic,sum_midcen, by = "File") %>%
  merge(sum_endcen) %>%
  rename(historic_T_mean=T_mean.x,
         historic_T_low10=T_low10.x,
         historic_DO_mean=DO_mean.x,
         historic_DO_low10=DO_low10.x,
         historic_pH_mean=pH_mean.x,
         historic_pH_low10=pH_low10.x,
         historic_T_seasonalSD=T_seasonalSD.x,
         historic_pH_seasonalSD=pH_seasonalSD.x,
         historic_DO_seasonalSD=DO_seasonalSD.x,
         historic_T_eventSD= T_eventSD.x,
         historic_pH_eventSD=pH_eventSD.x,
         historic_DO_eventSD=DO_eventSD.x,
         midcen_T_mean=T_mean.y,
         midcen_T_low10=T_low10.y,
         midcen_DO_mean=DO_mean.y,
         midcen_DO_low10=DO_low10.y,
         midcen_pH_mean=pH_mean.y,
         midcen_pH_low10=pH_low10.y,
         midcen_T_seasonalSD=T_seasonalSD.y,
         midcen_pH_seasonalSD=pH_seasonalSD.y,
         midcen_DO_seasonalSD= DO_seasonalSD.y,
         midcen_T_eventSD=T_eventSD.y,
         midcen_pH_eventSD=pH_eventSD.y,
         midcen_DO_eventSD=DO_eventSD.y,
         endcen_T_mean=T_mean,
         endcen_T_low10=T_low10,
         endcen_DO_mean=DO_mean,
         endcen_DO_low10=DO_low10,
         endcen_pH_mean=pH_mean,
         endcen_pH_low10=pH_low10,
         endcen_T_seasonalSD=T_seasonalSD,
         endcen_pH_seasonalSD=pH_seasonalSD,
         endcen_DO_seasonalSD=DO_seasonalSD,
         endcen_T_eventSD=T_eventSD,
         endcen_pH_eventSD=pH_eventSD,
         endcen_DO_eventSD=DO_eventSD) %>%
  mutate(mid_diff_T_mean = midcen_T_mean - historic_T_mean,
         mid_diff_T_low10 = midcen_T_low10 - historic_T_low10,
         mid_diff_T_seasonalSD = midcen_T_seasonalSD - historic_T_seasonalSD,
         mid_diff_T_eventSD = midcen_T_eventSD - historic_T_eventSD,
         mid_diff_pH_mean = midcen_pH_mean - historic_pH_mean,
         mid_diff_pH_low10 = midcen_pH_low10 - historic_pH_low10,
         mid_diff_pH_seasonalSD = midcen_pH_seasonalSD - historic_pH_seasonalSD,
         mid_diff_pH_eventSD = midcen_pH_eventSD - historic_pH_eventSD,
         mid_diff_DO_mean = midcen_DO_mean - historic_DO_mean,
         mid_diff_DO_low10 = midcen_DO_low10 - historic_DO_low10,
         mid_diff_DO_seasonalSD = midcen_DO_seasonalSD - historic_DO_seasonalSD,
         mid_diff_DO_eventSD = midcen_DO_eventSD - historic_DO_eventSD,
         end_diff_T_mean = endcen_T_mean - historic_T_mean,
         end_diff_T_low10 = endcen_T_low10 - historic_T_low10,
         end_diff_T_seasonalSD = endcen_T_seasonalSD - historic_T_seasonalSD,
         end_diff_T_eventSD = endcen_T_eventSD - historic_T_eventSD,
         end_diff_pH_mean = endcen_pH_mean - historic_pH_mean,
         end_diff_pH_low10 = endcen_pH_low10 - historic_pH_low10,
         end_diff_pH_seasonalSD = endcen_pH_seasonalSD - historic_pH_seasonalSD,
         end_diff_pH_eventSD = endcen_pH_eventSD - historic_pH_eventSD,
         end_diff_DO_mean = endcen_DO_mean - historic_DO_mean,
         end_diff_DO_low10 = endcen_DO_low10 - historic_DO_low10,
         end_diff_DO_seasonalSD = endcen_DO_seasonalSD - historic_DO_seasonalSD,
         end_diff_DO_eventSD = endcen_DO_eventSD - historic_DO_eventSD)

```

## PCA

PCA with all summary stats (mean, lower 10th percentile, seasonal SD, event SD)

```{r}
#| label: pca

sumsub_historic <- sum_historic %>% select(-OBJECTID,-NAME, -File,  -Area_sq_mi, -Type, -SHORTNAME, -degx, -degy, -region) 
sumsub_midcen <- sum_midcen %>% select(-OBJECTID,-File, -NAME, -Area_sq_mi, -Type, -SHORTNAME, -degx, -degy, -region)
sumsub_endcen <- sum_endcen %>% select(-OBJECTID,-File, -NAME, -Area_sq_mi, -Type, -SHORTNAME, -degx, -degy, -region) 


pca_historic <- prcomp(sumsub_historic, scale = TRUE)
fviz_pca_biplot(pca_historic, repel = TRUE,
                col.var = "black",
                col.ind = sum_historic$region,
                label ="var",
                labelsize = 3,
                addEllipses = TRUE,
                title = "IPSL Historic") 

pca_midcen <- prcomp(sumsub_midcen, scale = TRUE)
fviz_pca_biplot(pca_midcen, repel = TRUE,
                col.var = "black",
                col.ind = sum_midcen$region,
                label ="var",
                labelsize = 3,
                addEllipses = TRUE,
                title = "IPSL Mid cen") 

pca_endcen <- prcomp(sumsub_endcen, scale = TRUE) #REMOVE [,-1] IF DONT WANT TO ROTATE
fviz_pca_biplot(pca_historic, repel = TRUE,
                col.var = "black",
                col.ind = sum_endcen$region,
                label ="var",
                labelsize = 3,
                addEllipses = TRUE,
                title = "IPSL End cen") 


fviz_pca_ind(pca_historic, label="none", habillage=sum_historic$region,
     addEllipses=TRUE,  col.ind = sum_endcen$region)
```

## Heatmaps with all variables

Create latitudinaly arranged heatmaps showing all summary stats for all MPAs

```{r}
#| label: heatmap

#HISTORIC
#ordering by latitude so rows make some sense in heatmap
historic_selectvar <- sum_historic[order(sum_historic$degy),] %>%
  select(NAME,T_mean,pH_mean,DO_mean,T_low10,pH_low10,DO_low10, T_seasonalSD,pH_seasonalSD,           DO_seasonalSD, T_eventSD, pH_eventSD,DO_eventSD, region)

#selecting data we want and converting first column into rownames to label rows later in the heatmap
historic_selectvar_norca <- historic_selectvar %>%
  filter(region == "norca") %>% select(-region)
historic_selectvar_centralca <- historic_selectvar %>%
  filter(region == "centralca") %>% select(-region)
historic_selectvar_channel <- historic_selectvar %>%
  filter(region == "channel") %>% select(-region)
historic_selectvar_socal <- historic_selectvar %>%
  filter(region == "socal") %>% select(-region)

historic_rnames_norca <- historic_selectvar_norca[,1]
historic_rnames_centralca <- historic_selectvar_centralca[,1]
historic_rnames_channel <- historic_selectvar_channel[,1]
historic_rnames_socal <- historic_selectvar_socal[,1]

col <- colorRampPalette(brewer.pal(9,"YlOrRd"))(256)

#heatmaps
historic_selectvar_norca <- as.matrix(historic_selectvar_norca[,2:ncol(historic_selectvar_norca)])
rownames(historic_selectvar_norca) <- historic_rnames_norca
heatmap(historic_selectvar_norca, col=col,Rowv = NA, Colv = NA, 
        scale="column", main = "Historic IPSL Northern CA") 
      
historic_selectvar_centralca <- as.matrix(historic_selectvar_centralca[,2:ncol(historic_selectvar_centralca)])
rownames(historic_selectvar_centralca) <- historic_rnames_centralca
heatmap(historic_selectvar_centralca,col=col, Rowv = NA, Colv = NA, 
        scale="column", main = "Historic IPSL Central CA") 

historic_selectvar_channel <- as.matrix(historic_selectvar_channel[,2:ncol(historic_selectvar_channel)])
rownames(historic_selectvar_channel) <- historic_rnames_channel
heatmap(historic_selectvar_channel,col=col, Rowv = NA, Colv = NA, 
        scale="column", main = "Historic IPSL Channel Islands") 

historic_selectvar_socal <- as.matrix(historic_selectvar_socal[,2:ncol(historic_selectvar_socal)])
rownames(historic_selectvar_socal) <- historic_rnames_socal
heatmap(historic_selectvar_socal, col=col,Rowv = NA, Colv = NA,  
        scale="column", main = "Historic IPSL Southern CA") 


#MIDCEN
#ordering by latitude so rows make some sense in heatmap
midcen_selectvar <- sum_diff[order(sum_historic$degy),] %>%
  select(NAME.x,mid_diff_T_mean,mid_diff_pH_mean,mid_diff_DO_mean,mid_diff_T_low10,mid_diff_pH_low10,mid_diff_DO_low10, mid_diff_T_seasonalSD,mid_diff_pH_seasonalSD,           mid_diff_DO_seasonalSD, mid_diff_T_eventSD, mid_diff_pH_eventSD,mid_diff_DO_eventSD, region)

#selecting data we want and converting first column into rownames to label rows later in the heatmap
midcen_selectvar_norca <- midcen_selectvar %>%
  filter(region == "norca") %>% select(-region)
midcen_selectvar_centralca <- midcen_selectvar %>%
  filter(region == "centralca") %>% select(-region)
midcen_selectvar_channel <- midcen_selectvar %>%
  filter(region == "channel") %>% select(-region)
midcen_selectvar_socal <- midcen_selectvar %>%
  filter(region == "socal") %>% select(-region)

midcen_rnames_norca <- midcen_selectvar_norca[,1]
midcen_rnames_centralca <- midcen_selectvar_centralca[,1]
midcen_rnames_channel <- midcen_selectvar_channel[,1]
midcen_rnames_socal <- midcen_selectvar_socal[,1]

#heatmaps
midcen_selectvar_norca <- as.matrix(midcen_selectvar_norca[,2:ncol(midcen_selectvar_norca)])
rownames(midcen_selectvar_norca) <- midcen_rnames_norca
heatmap(midcen_selectvar_norca, Rowv = NA, Colv = NA, 
        scale="column", col=col,main = "Mid cen diff IPSL Northern CA") 
      
midcen_selectvar_centralca <- as.matrix(midcen_selectvar_centralca[,2:ncol(midcen_selectvar_centralca)])
rownames(midcen_selectvar_centralca) <- midcen_rnames_centralca
heatmap(midcen_selectvar_centralca, Rowv = NA, Colv = NA, 
        scale="column", col=col,main = "Mid cen diff IPSL Central CA") 

midcen_selectvar_channel <- as.matrix(midcen_selectvar_channel[,2:ncol(midcen_selectvar_channel)])
rownames(midcen_selectvar_channel) <- midcen_rnames_channel
heatmap(midcen_selectvar_channel, Rowv = NA, Colv = NA, 
        scale="column", main = "Mid cen diff IPSL Channel Islands") 

midcen_selectvar_socal <- as.matrix(midcen_selectvar_socal[,2:ncol(midcen_selectvar_socal)])
rownames(midcen_selectvar_socal) <- midcen_rnames_socal
heatmap(midcen_selectvar_socal, Rowv = NA, Colv = NA,  
        scale="column", main = "Mid cen diff IPSL Southern CA") 


#ENDCEN
#ordering by latitude so rows make some sense in heatmap
endcen_selectvar <- sum_diff[order(sum_endcen$degy),] %>%
  select(NAME.x,end_diff_T_mean,end_diff_pH_mean,end_diff_DO_mean,end_diff_T_low10,end_diff_pH_low10,end_diff_DO_low10, end_diff_T_seasonalSD,end_diff_pH_seasonalSD,           end_diff_DO_seasonalSD, end_diff_T_eventSD, end_diff_pH_eventSD,end_diff_DO_eventSD, region)

#selecting data we want and converting first column into rownames to label rows later in the heatmap
endcen_selectvar_norca <- endcen_selectvar %>%
  filter(region == "norca") %>% select(-region)
endcen_selectvar_centralca <- endcen_selectvar %>%
  filter(region == "centralca") %>% select(-region)
endcen_selectvar_channel <- endcen_selectvar %>%
  filter(region == "channel") %>% select(-region)
endcen_selectvar_socal <- endcen_selectvar %>%
  filter(region == "socal") %>% select(-region)

endcen_rnames_norca <- endcen_selectvar_norca[,1]
endcen_rnames_centralca <- endcen_selectvar_centralca[,1]
endcen_rnames_channel <- endcen_selectvar_channel[,1]
endcen_rnames_socal <- endcen_selectvar_socal[,1]

#heatmaps
endcen_selectvar_norca <- as.matrix(endcen_selectvar_norca[,2:ncol(endcen_selectvar_norca)])
rownames(endcen_selectvar_norca) <- endcen_rnames_norca
heatmap(endcen_selectvar_norca, Rowv = NA, Colv = NA, 
        scale="column", col=col,main = "End cen diff IPSL Northern CA") 
      
endcen_selectvar_centralca <- as.matrix(endcen_selectvar_centralca[,2:ncol(endcen_selectvar_centralca)])
rownames(endcen_selectvar_centralca) <- endcen_rnames_centralca
heatmap(endcen_selectvar_centralca, Rowv = NA, Colv = NA, 
        scale="column", col=col,main = "End cen diff IPSL Central CA") 

endcen_selectvar_channel <- as.matrix(endcen_selectvar_channel[,2:ncol(endcen_selectvar_channel)])
rownames(endcen_selectvar_channel) <- endcen_rnames_channel
heatmap(endcen_selectvar_channel, Rowv = NA, Colv = NA, 
        scale="column", col=col,main = "End cen diff IPSL Channel Islands") 

endcen_selectvar_socal <- as.matrix(endcen_selectvar_socal[,2:ncol(endcen_selectvar_socal)])
rownames(endcen_selectvar_socal) <- endcen_rnames_socal
heatmap(endcen_selectvar_socal, Rowv = NA, Colv = NA,  
        scale="column", col=col,main = "End cen diff IPSL Southern CA") 



```

## Heatmaps by variable

```{r}
col <- colorRampPalette(brewer.pal(9,"YlOrRd"))(256)
invert_col <- colorRampPalette(rev(brewer.pal(9,"YlOrRd")))(256)
```

Heatmaps across all regions for T sumary stats

```{r}
#T mean all regions
sum_diff_T_mean <- sum_diff[order(sum_diff$degy.x),] %>%
  select(NAME,historic_T_mean, midcen_T_mean, endcen_T_mean) 

rownames(sum_diff_T_mean) <- sum_diff_T_mean[,1]

Tmean <- as.matrix(sum_diff_T_mean[,2:4])

heatmap.2(Tmean,Rowv = FALSE, Colv = FALSE, dendrogram = "none", 
          main = "T mean",tracecol=NA, revC= TRUE,
          margins = c(5,10), col=col, srtCol = 45)

#T low 10 all regions
sum_diff_T_low10 <- sum_diff[order(sum_diff$degy.x),] %>%
  select(NAME,historic_T_low10, midcen_T_low10, endcen_T_low10) 

rownames(sum_diff_T_low10) <- sum_diff_T_low10[,1]

Tlow10 <- as.matrix(sum_diff_T_low10[,2:4])

heatmap.2(Tlow10,Rowv = FALSE, Colv = FALSE, dendrogram = "none", 
          main = "T low10",tracecol=NA, revC= TRUE,
          margins = c(5,10), col=col, srtCol = 45)

#T seasonalSD all regions
sum_diff_T_seasonalSD <- sum_diff[order(sum_diff$degy.x),] %>%
  select(NAME,historic_T_seasonalSD, midcen_T_seasonalSD, endcen_T_seasonalSD) 

rownames(sum_diff_T_seasonalSD) <- sum_diff_T_seasonalSD[,1]

TseasonalSD <- as.matrix(sum_diff_T_seasonalSD[,2:4])

heatmap.2(TseasonalSD,Rowv = FALSE, Colv = FALSE, dendrogram = "none", 
          main = "T seasonalSD",tracecol=NA, revC= TRUE,
          margins = c(5,10), col=col, srtCol = 45)

#T eventSD all regions
sum_diff_T_eventSD <- sum_diff[order(sum_diff$degy.x),] %>%
  select(NAME,historic_T_eventSD, midcen_T_eventSD, endcen_T_eventSD) 

rownames(sum_diff_T_eventSD) <- sum_diff_T_eventSD[,1]

TeventSD <- as.matrix(sum_diff_T_eventSD[,2:4])

heatmap.2(TeventSD,Rowv = FALSE, Colv = FALSE, dendrogram = "none", 
          main = "T eventSD",tracecol=NA, revC= TRUE,
          margins = c(5,10), col=col, srtCol = 45)
```

Heatmaps across all regions for DO summary stats

```{r}
#DO mean all regions
sum_diff_DO_mean <- sum_diff[order(sum_diff$degy.x),] %>%
  select(NAME,historic_DO_mean, midcen_DO_mean, endcen_DO_mean) 

rownames(sum_diff_DO_mean) <- sum_diff_DO_mean[,1]

DOmean <- as.matrix(sum_diff_DO_mean[,2:4])

heatmap.2(DOmean,Rowv = FALSE, Colv = FALSE, dendrogram = "none", 
          main = "DO mean",tracecol=NA, revC= TRUE,
          margins = c(5,10), col=invert_col, srtCol = 45)

#DO low 10 all regions
sum_diff_DO_low10 <- sum_diff[order(sum_diff$degy.x),] %>%
  select(NAME,historic_DO_low10, midcen_DO_low10, endcen_DO_low10) 

rownames(sum_diff_DO_low10) <- sum_diff_DO_low10[,1]

DOlow10 <- as.matrix(sum_diff_DO_low10[,2:4])

heatmap.2(DOlow10,Rowv = FALSE, Colv = FALSE, dendrogram = "none", 
          main = "DO low10",tracecol=NA, revC= TRUE,
          margins = c(5,10), col=invert_col, srtCol = 45)

#DO seasonalSD all regions
sum_diff_DO_seasonalSD <- sum_diff[order(sum_diff$degy.x),] %>%
  select(NAME,historic_DO_seasonalSD, midcen_DO_seasonalSD, endcen_DO_seasonalSD) 

rownames(sum_diff_DO_seasonalSD) <- sum_diff_DO_seasonalSD[,1]

DOseasonalSD <- as.matrix(sum_diff_DO_seasonalSD[,2:4])

heatmap.2(DOseasonalSD,Rowv = FALSE, Colv = FALSE, dendrogram = "none", 
          main = "DO seasonalSD",tracecol=NA, revC= TRUE,
          margins = c(5,10), col=col, srtCol = 45)

#DO eventSD all regions
sum_diff_DO_eventSD <- sum_diff[order(sum_diff$degy.x),] %>%
  select(NAME,historic_DO_eventSD, midcen_DO_eventSD, endcen_DO_eventSD) 

rownames(sum_diff_DO_eventSD) <- sum_diff_DO_eventSD[,1]

DOeventSD <- as.matrix(sum_diff_DO_eventSD[,2:4])

heatmap.2(DOeventSD,Rowv = FALSE, Colv = FALSE, dendrogram = "none", 
          main = "DO eventSD",tracecol=NA, revC= TRUE,
          margins = c(5,10), col=col, srtCol = 45)

```

Heatmaps across all regions for pH summary stats

```{r}
#pH mean all regions
sum_diff_pH_mean <- sum_diff[order(sum_diff$degy.x),] %>%
  select(NAME,historic_pH_mean, midcen_pH_mean, endcen_pH_mean) 

rownames(sum_diff_pH_mean) <- sum_diff_pH_mean[,1]

pHmean <- as.matrix(sum_diff_pH_mean[,2:4])

heatmap.2(pHmean,Rowv = FALSE, Colv = FALSE, dendrogram = "none", 
          main = "pH mean",tracecol=NA, revC= TRUE,
          margins = c(5,10), col=invert_col, srtCol = 45)

#pH low 10 all regions
sum_diff_pH_low10 <- sum_diff[order(sum_diff$degy.x),] %>%
  select(NAME,historic_pH_low10, midcen_pH_low10, endcen_pH_low10) 

rownames(sum_diff_pH_low10) <- sum_diff_pH_low10[,1]

pHlow10 <- as.matrix(sum_diff_pH_low10[,2:4])

heatmap.2(pHlow10,Rowv = FALSE, Colv = FALSE, dendrogram = "none", 
          main = "pH low10",tracecol=NA, revC= TRUE,
          margins = c(5,10), col=invert_col, srtCol = 45)

#pH seasonalSD all regions
sum_diff_pH_seasonalSD <- sum_diff[order(sum_diff$degy.x),] %>%
  select(NAME,historic_pH_seasonalSD, midcen_pH_seasonalSD, endcen_pH_seasonalSD) 

rownames(sum_diff_pH_seasonalSD) <- sum_diff_pH_seasonalSD[,1]

pHseasonalSD <- as.matrix(sum_diff_pH_seasonalSD[,2:4])

heatmap.2(pHseasonalSD,Rowv = FALSE, Colv = FALSE, dendrogram = "none", 
          main = "pH seasonalSD",tracecol=NA, revC= TRUE,
          margins = c(5,10), col=col, srtCol = 45)

#pH eventSD all regions
sum_diff_pH_eventSD <- sum_diff[order(sum_diff$degy.x),] %>%
  select(NAME,historic_pH_eventSD, midcen_pH_eventSD, endcen_pH_eventSD) 

rownames(sum_diff_pH_eventSD) <- sum_diff_pH_eventSD[,1]

pHeventSD <- as.matrix(sum_diff_pH_eventSD[,2:4])

heatmap.2(pHeventSD,Rowv = FALSE, Colv = FALSE, dendrogram = "none", 
          main = "pH eventSD",tracecol=NA, revC= TRUE,
          margins = c(5,10), col=col, srtCol = 45)

```

## Regressions

Supplementing heatmaps by looking at correlation between historic and future time periods for a given summary stat

Temp:

```{r}
# T -----------------------------------------------------------------------

#historic T_mean vs. midcen T_mean
ggplot(data = sum_diff, aes(x = historic_T_mean, y = midcen_T_mean)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Midcen T_mean", x = "Historic T_mean") +
  ggtitle("Historic vs. midcen T_mean")

#historic T_mean vs. endcen T_mean
ggplot(data = sum_diff, aes(x = historic_T_mean, y = endcen_T_mean)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Endcen T_mean", x = "Historic T_mean") +
  ggtitle("Historic vs. endcen T_mean")

#historic T_seasonalSD vs. midcen T_seasonalSD
ggplot(data = sum_diff, aes(x = historic_T_seasonalSD, y = midcen_T_seasonalSD)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Midcen T_seasonalSD", x = "Historic T_seasonalSD") +
  ggtitle("Historic vs. midcen T_seasonalSD")

#historic T_seasonalSD vs. endcen T_seasonalSD
ggplot(data = sum_diff, aes(x = historic_T_seasonalSD, y = endcen_T_seasonalSD)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Endcen T_seasonalSD", x = "Historic T_seasonalSD") +
  ggtitle("Historic vs. endcen T_seasonalSD")

#historic T_eventSD vs. midcen T_eventSD
ggplot(data = sum_diff, aes(x = historic_T_eventSD, y = midcen_T_eventSD)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Midcen T_eventSD", x = "Historic T_eventSD") +
  ggtitle("Historic vs. midcen T_eventSD")

#historic T_eventSD vs. endcen T_eventSD
ggplot(data = sum_diff, aes(x = historic_T_eventSD, y = endcen_T_eventSD)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Endcen T_eventSD", x = "Historic T_eventSD") +
  ggtitle("Historic vs. endcen T_eventSD")

```

pH

```{r}
#historic pH_mean vs. midcen pH_mean
ggplot(data = sum_diff, aes(x = historic_pH_mean, y = midcen_pH_mean)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Midcen pH_mean", x = "Historic pH_mean") +
  ggtitle("Historic vs. midcen pH_mean")

#historic pH_mean vs. endcen pH_mean
ggplot(data = sum_diff, aes(x = historic_pH_mean, y = endcen_pH_mean)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Endcen pH_mean", x = "Historic pH_mean") +
  ggtitle("Historic vs. endcen pH_mean")

#historic pH_seasonalSD vs. midcen pH_seasonalSD
ggplot(data = sum_diff, aes(x = historic_pH_seasonalSD, y = midcen_pH_seasonalSD)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Midcen pH_seasonalSD", x = "Historic pH_seasonalSD") +
  ggtitle("Historic vs. midcen pH_seasonalSD")

#historic pH_seasonalSD vs. endcen pH_seasonalSD
ggplot(data = sum_diff, aes(x = historic_pH_seasonalSD, y = endcen_pH_seasonalSD)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Endcen pH_seasonalSD", x = "Historic pH_seasonalSD") +
  ggtitle("Historic vs. endcen pH_seasonalSD")

#historic pH_eventSD vs. midcen pH_eventSD
ggplot(data = sum_diff, aes(x = historic_pH_eventSD, y = midcen_pH_eventSD)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Midcen pH_eventSD", x = "Historic pH_eventSD") +
  ggtitle("Historic vs. midcen pH_eventSD")

#historic pH_eventSD vs. endcen pH_eventSD
ggplot(data = sum_diff, aes(x = historic_pH_eventSD, y = endcen_pH_eventSD)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Endcen pH_eventSD", x = "Historic pH_eventSD") +
  ggtitle("Historic vs. endcen pH_eventSD")

```

DO

```{r}
#historic DO_mean vs. midcen DO_mean
ggplot(data = sum_diff, aes(x = historic_DO_mean, y = midcen_DO_mean)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Midcen DO_mean", x = "Historic DO_mean") +
  ggtitle("Historic vs. midcen DO_mean")

#historic DO_mean vs. endcen DO_mean
ggplot(data = sum_diff, aes(x = historic_DO_mean, y = endcen_DO_mean)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Endcen DO_mean", x = "Historic DO_mean") +
  ggtitle("Historic vs. endcen DO_mean")

#historic DO_seasonalSD vs. midcen DO_seasonalSD
ggplot(data = sum_diff, aes(x = historic_DO_seasonalSD, y = midcen_DO_seasonalSD)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Midcen DO_seasonalSD", x = "Historic DO_seasonalSD") +
  ggtitle("Historic vs. midcen DO_seasonalSD")

#historic DO_seasonalSD vs. endcen DO_seasonalSD
ggplot(data = sum_diff, aes(x = historic_DO_seasonalSD, y = endcen_DO_seasonalSD)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Endcen DO_seasonalSD", x = "Historic DO_seasonalSD") +
  ggtitle("Historic vs. endcen DO_seasonalSD")

#historic DO_eventSD vs. midcen DO_eventSD
ggplot(data = sum_diff, aes(x = historic_DO_eventSD, y = midcen_DO_eventSD)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Midcen DO_eventSD", x = "Historic DO_eventSD") +
  ggtitle("Historic vs. midcen DO_eventSD")

#historic DO_eventSD vs. endcen DO_eventSD
ggplot(data = sum_diff, aes(x = historic_DO_eventSD, y = endcen_DO_eventSD)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Endcen DO_eventSD", x = "Historic DO_eventSD") +
  ggtitle("Historic vs. endcen DO_eventSD")

```

## Anomalous pH/DO Event Analysis

Find percentage of unsaturated/ low DO days over all years using a biological threshold. Separated by upwelling and non-upwelling months. Output is used to make maps in QGIS.

```{r}
# Create labels for events (e.g., consecutive days with low pH or low DO) using run 
# length encoding
label_events <- function(is_event) {
  event_rle <- rle(is_event)
  labels <- ifelse(event_rle$values, cumsum(event_rle$values), NA)
  rep(labels, event_rle$lengths)
}

#historic =================================================================================

historic_pH_events <- mpa_historic %>% 
  mutate(is_pH_low = pH_surf < 7.75,
         pH_event = label_events(is_pH_low)) 
historic_pH_event_summary <- historic_pH_events %>% 
  drop_na(pH_event) %>% 
  group_by(File, pH_event) %>% 
  summarize(duration_days = n(),
            event_begin = min(Date),
            mean_pH = mean(pH_surf),
            intensity = 7.75 - mean_pH,
            severity = intensity*duration_days,
            .groups = "drop")
historic_pH_mpa_summary <- historic_pH_event_summary %>%
  group_by(File) %>%
  summarise(num_event = n(),
         mean_event_duration = mean(duration_days),
         max_event_duration = max(duration_days),
         mean_event_mean_pH = mean(mean_pH),
         mean_event_intensity = mean(intensity),
         mean_event_severity = mean(severity)) %>% 
  merge(mpa_centroids, by = "File")
#total days below threshold + avg ph below threhsold per mpa
historic_pH_mpa_summary2 <- historic_pH_events %>%
  filter(is_pH_low == TRUE) %>%
  group_by(File, Year) %>%
  summarise(annual_days_belowthresh = n(),
            annual_avg_ph_belowthresh = mean(pH_surf))
historic_pH_mpa_summary3 <- historic_pH_mpa_summary2 %>%
  group_by(File)%>%
  summarise(meanannual_days_belowthresh = mean(annual_days_belowthresh),
            meanannual_avg_pH_belowthresh = mean(annual_avg_ph_belowthresh)) %>% 
  mutate(period = "historic") %>%
  merge(historic_pH_mpa_summary, by = "File") 

#midcen =====================================================================================

midcen_pH_events <- mpa_midcen %>% 
  mutate(is_pH_low = pH_surf < 7.75,
         pH_event = label_events(is_pH_low)) 
midcen_pH_event_summary <- midcen_pH_events %>% 
  drop_na(pH_event) %>% 
  group_by(File, pH_event) %>% 
  summarize(duration_days = n(),
            event_begin = min(Date),
            mean_pH = mean(pH_surf),
            intensity = 7.75 - mean_pH,
            severity = intensity*duration_days,
            .groups = "drop")
midcen_pH_mpa_summary <- midcen_pH_event_summary %>%
  group_by(File) %>%
  summarise(num_event = n(),
            mean_event_duration = mean(duration_days),
            max_event_duration = max(duration_days),
            mean_event_mean_pH = mean(mean_pH),
            mean_event_intensity = mean(intensity),
            mean_event_severity = mean(severity)) %>% 
  merge(mpa_centroids, by = "File")
#total days below threshold + avg ph below threhsold per mpa
midcen_pH_mpa_summary2 <- midcen_pH_events %>%
  filter(is_pH_low == TRUE) %>%
  group_by(File, Year) %>%
  summarise(annual_days_belowthresh = n(),
            annual_avg_ph_belowthresh = mean(pH_surf))
midcen_pH_mpa_summary3 <- midcen_pH_mpa_summary2 %>%
  group_by(File)%>%
  summarise(meanannual_days_belowthresh = mean(annual_days_belowthresh),
            meanannual_avg_pH_belowthresh = mean(annual_avg_ph_belowthresh)) %>% 
  mutate(period = "midcen") %>%
  merge(midcen_pH_mpa_summary, by = "File")  

#endcen =====================================================================================

endcen_pH_events <- mpa_endcen %>% 
  mutate(is_pH_low = pH_surf < 7.75,
         pH_event = label_events(is_pH_low)) 
endcen_pH_event_summary <- endcen_pH_events %>% 
  drop_na(pH_event) %>% 
  group_by(File, pH_event) %>% 
  summarize(duration_days = n(),
            event_begin = min(Date),
            mean_pH = mean(pH_surf),
            intensity = 7.75 - mean_pH,
            severity = intensity*duration_days,
            .groups = "drop")
endcen_pH_mpa_summary <- endcen_pH_event_summary %>%
  group_by(File) %>%
  summarise(num_event = n(),
            mean_event_duration = mean(duration_days),
            max_event_duration = max(duration_days),
            mean_event_mean_pH = mean(mean_pH),
            mean_event_intensity = mean(intensity),
            mean_event_severity = mean(severity)) %>% 
  merge(mpa_centroids, by = "File")
#total days below threshold + avg ph below threhsold per mpa
endcen_pH_mpa_summary2 <- endcen_pH_events %>%
  filter(is_pH_low == TRUE) %>%
  group_by(File, Year) %>%
  summarise(annual_days_belowthresh = n(),
            annual_avg_ph_belowthresh = mean(pH_surf))
endcen_pH_mpa_summary3 <- endcen_pH_mpa_summary2 %>%
  group_by(File)%>%
  summarise(meanannual_days_belowthresh = mean(annual_days_belowthresh),
            meanannual_avg_pH_belowthresh = mean(annual_avg_ph_belowthresh)) %>% 
  mutate(period = "endcen") %>%
  merge(endcen_pH_mpa_summary, by = "File")   


#merging pH event hist/midcen/endcen
pH_event <- full_join(historic_pH_mpa_summary3, midcen_pH_mpa_summary3) %>%
  full_join(endcen_pH_mpa_summary3)
```

```{r}
historic_DO_events <- mpa_historic %>% 
  mutate(is_DO_low = DO_mgL < 4.6,
         DO_event = label_events(is_DO_low)) 
historic_DO_event_summary <- historic_DO_events %>% 
  drop_na(DO_event) %>% 
  group_by(File, DO_event) %>% 
  summarize(duration_days = n(),
            event_begin = min(Date),
            mean_DO = mean(DO_mgL),
            intensity = 4.6 - mean_DO,
            severity = intensity*duration_days,
            .groups = "drop")
historic_DO_mpa_summary <- historic_DO_event_summary %>%
  group_by(File) %>%
  summarise(histDO_num_event = n(),
            histDO_mean_event_duration = mean(duration_days),
            histDO_max_event_duration = max(duration_days),
            histDO_mean_event_mean_DO = mean(mean_DO),
            histDO_mean_event_intensity = mean(intensity),
            histDO_mean_event_severity = mean(severity)) %>% 
  merge(mpa_centroids, by = "File")
#total days below threshold + avg DO below threhsold per mpa
historic_DO_mpa_summary2 <- historic_DO_events %>%
  filter(is_DO_low == TRUE) %>%
  group_by(File, Year) %>%
  summarise(histDO_annual_days_belowthresh = n(),
            histDO_annual_avg_DO_belowthresh = mean(DO_mgL))
historic_DO_mpa_summary3 <- historic_DO_mpa_summary2 %>%
  group_by(File)%>%
  summarise(histDO_meanannual_days_belowthresh = mean(histDO_annual_days_belowthresh),
            histDO_meanannual_avg_DO_belowthresh = mean(histDO_annual_avg_DO_belowthresh)) %>%
  merge(historic_DO_mpa_summary, by = "File") 

#midcen =====================================================================================

midcen_DO_events <- mpa_midcen %>% 
  mutate(is_DO_low = DO_mgL < 4.6,
         DO_event = label_events(is_DO_low)) 
midcen_DO_event_summary <- midcen_DO_events %>% 
  drop_na(DO_event) %>% 
  group_by(File, DO_event) %>% 
  summarize(duration_days = n(),
            event_begin = min(Date),
            mean_DO = mean(DO_mgL),
            intensity = 4.6 - mean_DO,
            severity = intensity*duration_days,
            .groups = "drop")
midcen_DO_mpa_summary <- midcen_DO_event_summary %>%
  group_by(File) %>%
  summarise(midcenDO_num_event = n(),
            midcenDO_mean_event_duration = mean(duration_days),
            midcenDO_max_event_duration = max(duration_days),
            midcenDO_mean_event_mean_DO = mean(mean_DO),
            midcenDO_mean_event_intensity = mean(intensity),
            midcenDO_mean_event_severity = mean(severity)) %>% 
  merge(mpa_centroids, by = "File")
#total days below threshold + avg DO below threhsold per mpa
midcen_DO_mpa_summary2 <- midcen_DO_events %>%
  filter(is_DO_low == TRUE) %>%
  group_by(File, Year) %>%
  summarise(midcenDO_annual_days_belowthresh = n(),
            midcenDO_annual_avg_DO_belowthresh = mean(DO_mgL))
midcen_DO_mpa_summary3 <- midcen_DO_mpa_summary2 %>%
  group_by(File)%>%
  summarise(midcenDO_meanannual_days_belowthresh = mean(midcenDO_annual_days_belowthresh),
            midcenDO_meanannual_avg_DO_belowthresh = mean(midcenDO_annual_avg_DO_belowthresh)) %>%
  merge(midcen_DO_mpa_summary, by = "File") 

#endcen =====================================================================================

endcen_DO_events <- mpa_endcen %>% 
  mutate(is_DO_low = DO_mgL < 4.6,
         DO_event = label_events(is_DO_low)) 
endcen_DO_event_summary <- endcen_DO_events %>% 
  drop_na(DO_event) %>% 
  group_by(File, DO_event) %>% 
  summarize(duration_days = n(),
            event_begin = min(Date),
            mean_DO = mean(DO_mgL),
            intensity = 4.6 - mean_DO,
            severity = intensity*duration_days,
            .groups = "drop")
endcen_DO_mpa_summary <- endcen_DO_event_summary %>%
  group_by(File) %>%
  summarise(endcenDO_num_event = n(),
            endcenDO_mean_event_duration = mean(duration_days),
            endcenDO_max_event_duration = max(duration_days),
            endcenDO_mean_event_mean_DO = mean(mean_DO),
            endcenDO_mean_event_intensity = mean(intensity),
            endcenDO_mean_event_severity = mean(severity)) %>% 
  merge(mpa_centroids, by = "File")
#total days below threshold + avg DO below threhsold per mpa
endcen_DO_mpa_summary2 <- endcen_DO_events %>%
  filter(is_DO_low == TRUE) %>%
  group_by(File, Year) %>%
  summarise(endcenDO_annual_days_belowthresh = n(),
            endcenDO_annual_avg_DO_belowthresh = mean(DO_mgL))
endcen_DO_mpa_summary3 <- endcen_DO_mpa_summary2 %>%
  group_by(File)%>%
  summarise(endcenDO_meanannual_days_belowthresh = mean(endcenDO_annual_days_belowthresh),
            endcenDO_meanannual_avg_DO_belowthresh = mean(endcenDO_annual_avg_DO_belowthresh)) %>%
  merge(endcen_DO_mpa_summary, by = "File") 


#merging DO event hist/midcen/endcen
DO_event <- full_join(historic_DO_mpa_summary3, midcen_DO_mpa_summary3) %>%
  full_join(endcen_DO_mpa_summary3)
```

## Anomalous temp analysis

Take climatological average for every day and climatological SD in a sliding 30 day window to figure out temperature deviations

```{r}
#historic ===================================================================================
#Historical climatology = mpa_historic_climatology. Append this to historic T dataset
mpa_historic_climatology_select <- mpa_historic_climatology %>%
  select(-pH_clim, -DO_clim) %>%
  merge(mpa_historic, by = c("File", "Month"))

historic_temp_events <- mpa_historic_climatology_select %>% 
  mutate(is_temp_high = T_surf > (T_clim + 2*T_clim_sd),
         temp_event = label_events(is_temp_high)) 
historic_temp_event_summary <- historic_temp_events %>% 
  drop_na(temp_event) %>% 
  group_by(File, temp_event) %>% 
  summarize(duration_days = n(),
            event_begin = min(Date),
            mean_temp = mean(T_surf),
            intensity = T_surf - (T_clim + 2*T_clim_sd),
            severity = intensity*duration_days,
            .groups = "drop")
historic_temp_mpa_summary <- historic_temp_event_summary %>%
  group_by(File) %>%
  summarise(histT_num_event = n(),
         histT_mean_event_duration = mean(duration_days),
         histT_max_event_duration = max(duration_days),
         histT_mean_event_mean_pH = mean(mean_temp),
         histT_mean_event_intensity = mean(intensity),
         histT_mean_event_severity = mean(severity)) %>% 
  merge(mpa_centroids, by = "File")
#total days below threshold + avg ph below threhsold per mpa
historic_temp_mpa_summary2 <- historic_temp_events %>%
  filter(is_temp_high == TRUE) %>%
  group_by(File, Year) %>%
  summarise(histT_annual_days_belowthresh = n(),
            histT_annual_avg_temp_belowthresh = mean(T_surf))
historic_temp_mpa_summary3 <- historic_temp_mpa_summary2 %>%
  group_by(File)%>%
  summarise(histT_meanannual_days_belowthresh = mean(histT_annual_days_belowthresh),
            histT_meanannual_avg_T_belowthresh = mean(histT_annual_avg_temp_belowthresh)) %>%
  merge(historic_temp_mpa_summary, by = "File") 


#midcen ===================================================================================
#Historical climatology = mpa_historic_climatology. Append historical clim and SD to midcen T dataset
mpa_midcen_climatology_select <- mpa_historic_climatology %>%
  select(-pH_clim, -DO_clim) %>%
  merge(mpa_midcen, by = c("File", "Month"))

midcen_temp_events <- mpa_midcen_climatology_select %>% 
  mutate(is_temp_high = T_surf > (T_clim + 2*T_clim_sd),
         temp_event = label_events(is_temp_high)) 
midcen_temp_event_summary <- midcen_temp_events %>% 
  drop_na(temp_event) %>% 
  group_by(File, temp_event) %>% 
  summarize(duration_days = n(),
            event_begin = min(Date),
            mean_temp = mean(T_surf),
            intensity = T_surf - (T_clim + 2*T_clim_sd),
            severity = intensity*duration_days,
            .groups = "drop")
midcen_temp_mpa_summary <- midcen_temp_event_summary %>%
  group_by(File) %>%
  summarise(midcenT_num_event = n(),
         midcenT_mean_event_duration = mean(duration_days),
         midcenT_max_event_duration = max(duration_days),
         midcenT_mean_event_mean_pH = mean(mean_temp),
         midcenT_mean_event_intensity = mean(intensity),
         midcenT_mean_event_severity = mean(severity)) %>% 
  merge(mpa_centroids, by = "File")
#total days below threshold + avg ph below threhsold per mpa
midcen_temp_mpa_summary2 <- midcen_temp_events %>%
  filter(is_temp_high == TRUE) %>%
  group_by(File, Year) %>%
  summarise(midcenT_annual_days_belowthresh = n(),
            midcenT_annual_avg_temp_belowthresh = mean(T_surf))
midcen_temp_mpa_summary3 <- midcen_temp_mpa_summary2 %>%
  group_by(File)%>%
  summarise(midcenT_meanannual_days_belowthresh = mean(midcenT_annual_days_belowthresh),
            midcenT_meanannual_avg_T_belowthresh = mean(midcenT_annual_avg_temp_belowthresh)) %>%
  merge(midcen_temp_mpa_summary, by = "File") 

#endcen ===================================================================================
#Historical climatology = mpa_historic_climatology. Append historical clim and SD to endcen T dataset
mpa_endcen_climatology_select <- mpa_historic_climatology %>%
  select(-pH_clim, -DO_clim) %>%
  merge(mpa_endcen, by = c("File", "Month"))

endcen_temp_events <- mpa_endcen_climatology_select %>% 
  mutate(is_temp_high = T_surf > (T_clim + 2*T_clim_sd),
         temp_event = label_events(is_temp_high)) 
endcen_temp_event_summary <- endcen_temp_events %>% 
  drop_na(temp_event) %>% 
  group_by(File, temp_event) %>% 
  summarize(duration_days = n(),
            event_begin = min(Date),
            mean_temp = mean(T_surf),
            intensity = T_surf - (T_clim + 2*T_clim_sd),
            severity = intensity*duration_days,
            .groups = "drop")
endcen_temp_mpa_summary <- endcen_temp_event_summary %>%
  group_by(File) %>%
  summarise(endcenT_num_event = n(),
            endcenT_mean_event_duration = mean(duration_days),
            endcenT_max_event_duration = max(duration_days),
            endcenT_mean_event_mean_pH = mean(mean_temp),
            endcenT_mean_event_intensity = mean(intensity),
            endcenT_mean_event_severity = mean(severity)) %>% 
  merge(mpa_centroids, by = "File")
#total days below threshold + avg ph below threhsold per mpa
endcen_temp_mpa_summary2 <- endcen_temp_events %>%
  filter(is_temp_high == TRUE) %>%
  group_by(File, Year) %>%
  summarise(endcenT_annual_days_belowthresh = n(),
            endcenT_annual_avg_temp_belowthresh = mean(T_surf))
endcen_temp_mpa_summary3 <- endcen_temp_mpa_summary2 %>%
  group_by(File)%>%
  summarise(endcenT_meanannual_days_belowthresh = mean(endcenT_annual_days_belowthresh),
            endcenT_meanannual_avg_T_belowthresh = mean(endcenT_annual_avg_temp_belowthresh)) %>%
  merge(endcen_temp_mpa_summary, by = "File") 

#merging temp event hist/midcen/endcen
temp_event <- full_join(historic_temp_mpa_summary3, midcen_temp_mpa_summary3) %>%
  full_join(endcen_temp_mpa_summary3)

```

```{r}
#intensity/severity/duration figs

historic_temp_events %>% 
  left_join(historic_temp_event_summary, by = c("File", "temp_event")) %>% 
  filter(NAME == "Ano Nuevo SMR") %>% 
  ggplot(aes(julianday, Year, fill = intensity)) +
  geom_tile() +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, na.value = "skyblue") +
  theme_classic()

midcen_temp_events %>% 
  left_join(midcen_temp_event_summary, by = c("File", "temp_event")) %>% 
  filter(NAME == "Ano Nuevo SMR") %>% 
  ggplot(aes(julianday, Year, fill = intensity)) +
  geom_tile() +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, na.value = "skyblue") +
  theme_classic()

endcen_temp_events %>% 
  left_join(endcen_temp_event_summary, by = c("File", "temp_event")) %>% 
  filter(NAME == "Ano Nuevo SMR") %>% 
  ggplot(aes(julianday, Year, fill = intensity)) +
  geom_tile() +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, na.value = "skyblue") +
  theme_classic()
```

## Anom contd: regressions

Regressions of historical env var on x axis and mid/end cen on y axis, each point is an MPA. to visualize bright spots and hot spots. Need to figure out how to normalize data

```{r}
#first, for mean temp. historical vs. mid cen. next we can do historical vs. diff btwn historical and mid.

sum_diff_normalized <- sum_diff %>%
  mutate(historic_T_mean_norm = historic_T_mean/log(historic_T_mean),
         middiff_T_mean_norm = mid_diff_T_mean/log(mid_diff_T_mean))

#historic pH_mean vs. midcen pH_mean
ggplot(data = sum_diff, aes(x = historic_pH_mean, y = midcen_pH_mean)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Midcen pH_mean", x = "Historic pH_mean") +
  ggtitle("Historic vs. midcen pH_mean")

#historic pH_mean vs. midcen pH_mean
ggplot(data = sum_diff, aes(x = historic_pH_mean, y = midcen_pH_mean)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point() +
  theme_classic() +
  labs(y= "Midcen pH_mean", x = "Historic pH_mean") +
  ggtitle("Historic vs. midcen pH_mean")


```
