---
title: "MPA Exposure Analysis"
format: 
  html:
    code-fold: true
    toc: true
    page-layout: full
editor: visual
---

```{r}
#| echo: false

knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

------------------------------------------------------------------------

*Characterization of exposure of California MPAs to future pH, DO, and temperature using IPSL model predictions.*

## Setup

Load required libraries, read IPSL MPA projections file, and assign MPAs to regions (Northern CA, Central CA, Southern CA, Channel Islands).

```{r}
#| label: libraries
#| message: false
#| echo: false

library(tidyverse)
library(lubridate)
library(data.table)
library(factoextra)
library(broom)
library(cowplot)
library(respR)
library(here)
library(lattice)
library(RcppRoll)
library(RColorBrewer)
library(gplots)
library(ggpmisc)
library(dendextend)
library(colourvalues)
library(zoo)
library(terra) #needed for tmap?
library(tmap) 
library(sf) 
library(stringr)
library(heatmaply)
library(ggpubr)
library(ggrepel)
library(flextable)
```

```{r}
#| label: load_data
#| message: false

#Read mpa data, convert DO units, add a column for time period, and one for difference between mid/end cen and historic.
mpa <- read_csv(here("data/processeddata/model/IPSLmpa.csv")) %>% 
  mutate(DO_mmolL = DO_surf/1000,
         DO_mgL = convert_DO(DO_mmolL, from = "mmol/L", to = "mg/L")) %>% 
  filter(Year <= 2020|
         Year >= 2040 & Year <= 2060|
         Year >= 2080 & Year <= 2100) %>%
  mutate(period = case_when((Year <= 2020) ~ "historic", 
                           (Year %in% c(2040:2060)) ~ "midcen",
                           (Year %in% c(2080:2100)) ~ "endcen")) %>%
  select(-T_bot, -DO_bot, -pH_bot,-DO_surf,-DO_mmolL) %>%
  rename(Temp = T_surf,
         DO = DO_mgL,
         pH = pH_surf) %>%
  mutate(Date = make_date(Year, Month, Day)) %>%
  mutate(julianday = yday(Date))

mpa$File <- substr(mpa$File, 1, nchar(mpa$File)-4)

```

```{r}
#| label: regions
#| message: false
#add regions here!
mpa_centroids<- read_csv(here("data/rawdata/MPA_polygons.csv")) %>%
  select(-Area_sq_mi, -Type)
mpa_centroids$File <- sub("^", "tphdo_mpa_", mpa_centroids$OBJECTID )

mpa_centroids <- mpa_centroids %>%
  mutate(region = ifelse(degy >= 37.29, "norca", 
                         ifelse(degy > 34.8, "centralca", 
                                ifelse("degy" < 34.274 & "degx" < -119.220, "channelisl" , 
                                       "socal"))))

channel <- c("Anacapa Island FMCA", "Anacapa Island FMR", "Anacapa Island SMCA", "Anacapa Island SMR", "Anacapa Island Special Closure", 
             "Arrow Point to Lion Head Point SMCA", "Begg Rock SMR", "Blue Cavern Offshore SMCA", "Blue Cavern Onshore SMCA (No-Take)", 
             "Carrington Point SMR", "Casino Point SMCA (No-Take)", "Cat Harbor SMCA", "Farnsworth Offshore SMCA", "Farnsworth Onshore SMCA", 
             "Footprint FMR", "Footprint SMR", "Gull Island FMR", "Gull Island SMR", "Harris Point FMR", "Harris Point SMR", "Judith Rock SMR", 
             "Long Point SMR", "Lover’s Cove SMCA", "Painted Cave SMCA", "Richardson Rock FMR", "Richardson Rock SMR", "San Miguel Island Special Closure", 
             "Santa Barbara Island FMR", "Santa Barbara Island SMR", "Scorpion FMR", "Scorpion SMR", "Skunk Point SMR", "South Point FMR", "South Point SMR")

mpa_centroids$region[mpa_centroids$NAME %in% channel]  <- "channel" 
mpa <- merge(mpa, mpa_centroids,  by = "File")

```

## Calculate climatology

Create a climatology (for each MPA within a time period, Jan values averaged over all years - same for each month), so each MPA has one value per month that is averaged over all of the years. Seasonal SD is calculated later in summary stat section, and is the variation across these 12 climatologies for each MPA, representing the seasonal cycle.

```{r}
#| label: climatology
#| message: false

#calculate climatology for each MPA for each time period
mpa_climatology <- mpa %>%
  group_by(File, Month, period) %>%
    summarise(T_clim = mean(Temp), 
              pH_clim = mean(pH), 
              DO_clim =   mean(DO))

```

## Calculate event-based variation

To understand event-based variation (not attributed to the seasonal cycle), we need to find the deviation of actual daily values from a climatological value. To do so, a mock dataset with julian day climatologies is created by interpolating monthly climatology values. Actual daily values are subtracted from this interpolated dataset at the daily scale. The standard deviation of this difference is the event-based SD, which is calculated in the summary stat section.

*Q: instead of interpolating from monthly climatologies to create daily climatologies, why not average daily values to get daily climatologies?*

```{r}
#| label: inter_setup
#| message: false
#| warning: false

#' Setup for interpolation:
#' Need to create a day 1 and day 365 proxy. The approx function can only interpolate not extrapolate. The approx function is used to interpolate between the 15th day of each month, and day 1 and day 365 are used to interpolate the first 15 and last 15 days of the year.


##below creates empty vectors as big as we need (365 days per mpa) for each variable and each mpa. julianday is dates 1-365 as many mpa times

mpaslist = unique(mpa$File) 
mpas = rep(NA, 365*121)
julianday = rep(1:365, 121)
interp = rep(NA, 365*121)

#Set up a vector of julian day assignment for the 15th of each month and the first and last day of the year
x_in <- yday(as.Date(c("2001-01-01", 
                       "2001-01-15","2001-02-15","2001-03-15","2001-04-15",
                       "2001-05-15","2001-06-15","2001-07-15","2001-08-15",
                       "2001-09-15","2001-10-15","2001-11-15","2001-12-15", 
                       "2001-12-31")))

# creating a list of all the days of the year to interpolate to.
x_out <- (1:365) 

```

```{r}
#| label: interpolation

interpolate <- function(periodt, variable){
  
  mpa_period_climatology <- mpa_climatology %>%
    filter(period == periodt)
  
  for (i in 1:length(mpaslist)){ #for each mpa
    
    d <- mpa_period_climatology %>% #for given mpa iterating through, selects var from clim
      filter(File == mpaslist[i]) %>% 
      select(File,Month, !!sym(variable))
    
#use a weighted average. for dec 31, slightly more than half weight to dec (denoted by [12,3] index - 12 is the month, third column is the variable's clim value), little less than   half weight to jan. for jan 1, slightly more weight to jan, little less weight to dec. 
    Dec31 = as.numeric(( ((16/30) * (d[12,3])) +  
                           ((14/30) * d[1,3]) )) 
    Jan1 = as.numeric(( ((14/30) * (d[12,3])) + 
                          ((16/30) * d[1,3]) ))

    y_in <- c(Jan1, d[[variable]], Dec31)
    
    mod <- approx(x = x_in, #days to interpolate from
                 y = y_in, #temp values per day in x_in
                 xout = x_out) #1-365 days to interpolate out to
    
    #I don't understand this part
    mpas[((i-1)*365+1):(i*365)] <- mpaslist[i] #rep(mpaslist[1], 365)
    interp[((i-1)*365+1):(i*365)] <- mod$y
  }
  
  df <- data.frame(mpas, julianday, interp) %>%
    rename(File = mpas) %>%
    mutate(period = periodt) %>%
    rename(!!sym(variable) := interp)
  
  assign(x = paste("interp",variable,periodt, sep="_"), value = df, envir = globalenv())
  
}
```

```{r}
#| label: interpolation_function_call


interpolate("historic", "T_clim")
interpolate("midcen", "T_clim")
interpolate("endcen", "T_clim")

interpolate("historic", "pH_clim")
interpolate("midcen", "pH_clim")
interpolate("endcen", "pH_clim")

interpolate("historic", "DO_clim")
interpolate("midcen", "DO_clim")
interpolate("endcen", "DO_clim")
```

```{r}
#| label: combine_interp_dataframes

interpolated_temp <- rbind(interp_T_clim_historic,
                           interp_T_clim_midcen,
                           interp_T_clim_endcen)

interpolated_pH <- rbind(interp_pH_clim_historic,
                           interp_pH_clim_midcen,
                           interp_pH_clim_endcen)

interpolated_DO <- rbind(interp_DO_clim_historic,
                           interp_DO_clim_midcen,
                           interp_DO_clim_endcen)
```

## Create summary statistics

Generate summary statistics for T, pH, and DO.

*Seasonal SD* is across climatologies (monthly values averaged over all years resulting in 12 values per MPA - one per month - **the seasonal SD captures the seasonal cycle,** and is the variation across these values).

*Event SD* is calculated based on the deviation of daily values from interpolated daily values. The standard deviation per MPA is across all of the daily differences between daily value and interpolated values filtered for upwelling months May, June, July, August, and September (*filtered*, not *grouped by* upwelling month - the **Event SD captures variation of the deviations at a daily scale**).

Additional summary stats include the mean, lower 10th percentile, and upper 10th percentile.

```{r}
#| label: summary_stats
#| warning: false

#calculate seasonal SD 
seasonal_SD <- mpa_climatology %>%
  group_by(File, period) %>%
  summarise(T_seasonalSD = sd(T_clim),
            pH_seasonalSD = sd(pH_clim),
            DO_seasonalSD = sd(DO_clim))

#put all summary stats together - first, calculate event SD
sum <- mpa %>%
 
  #merge interpolated climatologies to mpa. (used to calculate event sd)
  merge(interpolated_temp, by = c("File", "julianday", "period")) %>%
  merge(interpolated_pH, by = c("File", "julianday", "period")) %>%
  merge(interpolated_DO, by = c("File", "julianday", "period")) %>%
  
  #subtract actual values - interpolated climatology values 
  #filtered for only upwelling months!!
  mutate(temp_deviation = Temp - T_clim,
         pH_deviation = pH - pH_clim,
         DO_deviation = DO - DO_clim) %>%
  filter(Month == c(5,6,7,8,9)) %>%
  
  #find event based SD 
  group_by(File, period) %>%
    mutate(T_eventSD = sd(temp_deviation),
            pH_eventSD = sd(pH_deviation),
            DO_eventSD = sd(DO_deviation)) %>%
  ungroup() %>%
    select(-T_clim, -pH_clim, -DO_clim, -temp_deviation,
         -pH_deviation, -DO_deviation) %>%
  
    #merge seasonal SD 
    merge(seasonal_SD, by = c("File", "period")) %>%
  
    #adding other summary stats - mean, low 10th and upper 10th percentiles
    group_by(File, period) %>%
    mutate(across(c(Temp, DO, pH), 
                   list(mean = mean, 
                        low10 = ~ quantile(.x, 0.1),
                        high10 = ~quantile(.x,0.9)))) %>%
  
  # to get to scale of one row per mpa
  select(-...1, -Temp, -pH, -DO, -julianday, -Year, -Month, -Day, -Date) %>%
  distinct(File, .keep_all = TRUE) %>% ungroup()

```

## PCA

PCA of all summary stats (mean, lower 10th percentile, upper 10th percentile, seasonal SD, event SD) clustered by regions within specified time period.

```{r}
# Remove metadata
sumsub <- sum %>%
  select(-OBJECTID, -NAME, -File, -SHORTNAME,
         -degx, -degy, -region, -period)

# PCA on all observations
pca_all <- prcomp(sumsub, scale. = TRUE)

#varimax rotation
n_comp <- 2
raw_loadings <- pca_all$rotation[, 1:n_comp]
varimax_res  <- varimax(raw_loadings)
pca_all$rotation[, 1:n_comp] <- varimax_res$loadings
scaled_data <- scale(sumsub)
pca_all$x[, 1:n_comp] <- scaled_data %*% varimax_res$loadings

#Keep first 3 PCs
scores <- as.data.frame(pca_all$x[,1:3])

#Add metadata back
scores$MPA <- sum$NAME
scores$MPA_num <- sum$OBJECTID
scores$region <- sum$region
scores$period <- sum$period

```

Cluster analysis

```{r}
#deciding how many clusters to use (not too sure about this)
fviz_nbclust(scores[,1:3], FUNcluster=kmeans, k.max = 8) 

#cluster: first set seed and run k means
set.seed(123)
km <- kmeans(scores[,1:2], centers = 4, nstart = 50)
scores$cluster <- factor(km$cluster)

cluster_table <- scores %>%
  select(MPA, MPA_num, region, period, cluster) %>%
  pivot_wider(names_from = period, values_from = cluster) %>%
  mutate(region = recode(region, "channel" = "Channel Islands","norca" = "Northern CA",
                         "socal" = "Southern CA", "centralca" = "Central CA"))

#count cluster changes
cluster_table$n_changes <-
  (cluster_table$historic != cluster_table$midcen) +
  (cluster_table$midcen != cluster_table$endcen)

#visualizing summary for cluster changes
general_cluster_table_summary <- cluster_table %>%
  group_by(region) %>%
  summarise(
    total_MPAs = n(),
    num_MPAs_changed = sum(n_changes > 0),
    mean_changes = mean(n_changes),
    .groups = "drop")

#making pretty table
ft_general_cluster_table <- flextable(general_cluster_table_summary) %>%
  set_header_labels(
    region = "Region", total_MPAs = "Total\nMPAs", num_MPAs_changed = "MPAs\nChanged",mean_changes = "Mean Cluster\nChanges") %>%
  colformat_double(j = "mean_changes", digits = 2) %>%
  theme_booktabs() %>%
  bold(part = "header") %>%
  align(align = "center", j = c("total_MPAs", "num_MPAs_changed", "mean_changes"),
        part = "all") %>%
  align(align = "left",j = "region",
        part = "all") %>%
  autofit() %>%
  fontsize(size = 10, part = "all")
ft_general_cluster_table


mpa_cluster_table_summary <- scores %>%
  select(MPA, MPA_num, region, period, cluster) %>%
  pivot_wider(
    names_from = period,
    values_from = cluster,
    names_glue = "{period}_cluster"
  ) %>%
  # Reorder columns
  select(region, historic_cluster, midcen_cluster, endcen_cluster, MPA, MPA_num) %>%
  # Keep only MPAs that changed cluster
  filter(historic_cluster != midcen_cluster | historic_cluster != endcen_cluster 
         |midcen_cluster != endcen_cluster) %>%
  # Collapse MPAs with identical trajectories
  group_by(region,historic_cluster,midcen_cluster,endcen_cluster) %>%
  summarise(MPAs = paste(MPA, collapse = ", "),MPA_nums = paste(MPA_num, collapse = ", "),
    .groups = "drop")  %>%
  arrange(region,historic_cluster,midcen_cluster,endcen_cluster)

#making a pretty table
ft_mpa_cluster_table <- flextable(mpa_cluster_table_summary) %>%
  set_header_labels(
    region = "Region",
    historic_cluster = "Historic\nCluster",
    midcen_cluster = "Mid-century\nCluster",
    endcen_cluster = "End-century\nCluster",
    MPAs = "MPAs",
    MPA_nums = "MPA IDs"
  ) %>%
  theme_booktabs() %>%
  autofit() %>%
  align(
    j = c("historic_cluster", "midcen_cluster", "endcen_cluster"),
    align = "center",
    part = "all"
  ) %>%
  align(
    j = c("region", "MPAs", "MPA_nums"),
    align = "left",
    part = "all"
  ) %>%
  merge_v(j = "region") %>%
  valign(j = "region", valign = "top") %>%
  bold(part = "header") %>%
  fontsize(size = 10, part = "all") %>%
  width(j = "region", width = 1.2) %>%
  width(j = c("historic_cluster", "midcen_cluster", "endcen_cluster"), width = 0.8) %>%
  width(j = "MPAs", width = 3.8) %>%
  width(j = "MPA_nums", width = 1.5)

ft_mpa_cluster_table

#ARI someway to show if clusters are similar or different?
scores %>%
  group_by(period) %>%
  summarise(
    ARI = adjustedRandIndex(region, cluster)
  )
```

Plotting results

```{r}
#plotting
hulls <- scores %>%
  group_by(cluster) %>%
  slice(chull(PC1, PC2))

scores$period <- factor(scores$period, 
                        levels = c("historic","midcen","endcen"))

#adding this to draw lines for only mpas that changed clusters
changed_mpas <- cluster_table %>%
  filter(n_changes > 0) %>%
  pull(MPA_num)

scores_segments <- scores %>%
  filter(MPA_num %in% changed_mpas) %>%
  select(MPA_num, region, period, PC1, PC2) %>%
  pivot_wider(
    names_from = period,
    values_from = c(PC1,PC2)
  )

#segments for plot
seg_hist_mid <- cluster_table %>%
  filter(historic != midcen) %>%
  select(MPA_num) %>%
  left_join(scores %>%
      select(MPA_num, period, PC1, PC2, region) %>%
      tidyr::pivot_wider(
        names_from = period,
        values_from = c(PC1, PC2)
      ),
    by = "MPA_num"
  )

seg_mid_end <- cluster_table %>%
  filter(midcen != endcen) %>%
  select(MPA_num) %>%
  left_join(scores %>%
              select(MPA_num, period, PC1, PC2, region) %>%
              tidyr::pivot_wider(
                names_from = period,
                values_from = c(PC1, PC2)
              ),
            by = "MPA_num"
  )

#pca biplot
p <- fviz_pca_biplot(
  pca_all,
  axes = c(1,2), #plots varimax rotated
  repel = TRUE,
  col.var = "black",
  col.ind = sum$region,
  pointshape = 19,
  label = "var",
  labelsize = 3,
  addEllipses = FALSE,
  invisible = "quali"
) 
p

#labeling cluster colors
cluster_cols <- c(
  "1" = "#8da0cb",
  "2" = "#66c2a5",
  "3" = "#fc8d62",
  "4" = "#e78ac3"
)
region_cols <- c(
  "centralca" = "#fc8d62",
  "channel" = "#66c2a5",
  "norca" = "#8da0cb",
  "socal" = "#e78ac3"
)

#overlaying pca biplot with the hulls from clusters
p +
  geom_polygon(
    data = hulls,
    aes(x = PC1, y = PC2, 
        group = cluster, fill = cluster),
    alpha = 0.3,
    color = NA,
    inherit.aes = FALSE
  ) +
  
  geom_segment(
    data = seg_hist_mid,
    aes(x = PC1_historic,
        y = PC2_historic,
        xend = PC1_midcen,
        yend = PC2_midcen,
        color = region),
    linewidth = 0.4,
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  
  geom_segment(
    data = seg_mid_end,
    aes(x = PC1_midcen,
        y = PC2_midcen,
        xend = PC1_endcen,
        yend = PC2_endcen,
        color = region),
    linewidth = 0.4,
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  
  geom_point(
    data = scores, 
    aes(PC1, PC2,
        color = region,
        shape = period),
    size = 2,
    inherit.aes = FALSE
  ) + 

  scale_fill_manual(values = cluster_cols) +
  scale_color_manual(values = region_cols) +
    
  
  labs(title = "All Period MPAs")
```

## NMDS Plots

```{r}
#saving factor scores from pca of whole dataset (above chunk) and only select PC1 (most explanatory)
factors <- as.data.frame(pca$x) %>%
  select("PC1") %>%
#I think the rows correspond to each individual mpa for a certain time period, corresponds to sum, need to merge this and only select pc1 since its most explanatory (but need to confirm this)
  cbind(sum) %>%
#convert from long to wide (for period) 
  pivot_wider(id_cols = c("File","region","SHORTNAME"),
              names_from = "period",
              values_from = "PC1",
              names_prefix = "PC1_") %>%
  select(File, region, SHORTNAME, PC1_historic, PC1_midcen, PC1_endcen)

#running lm model of historic vs. midcen and historic vs. endcen
hist_midcen_mod <- lm(PC1_midcen ~ PC1_historic, data = factors)
hist_endcen_mod <- lm(PC1_endcen ~ PC1_historic, data = factors)
midcen_endcen_mod <- lm( PC1_endcen ~ PC1_midcen, data = factors)

summary(hist_midcen_mod)
summary(hist_endcen_mod)

#saving residuals and fitted values. - scaled so wrt 0
factors2 <- factors %>% mutate(predicted_hist_midcen = predict(hist_midcen_mod),
                  residual_hist_midcen = residuals(hist_midcen_mod),
                  predicted_hist_endcen = predict(hist_endcen_mod),
                  residual_hist_endcen = residuals(hist_endcen_mod),
                  predicted_midcen_endcen = predict(midcen_endcen_mod),
                  residual_midcen_endcen = residuals(midcen_endcen_mod),
                  
                  scaled_predicted_hist_midcen = scale(predict(hist_midcen_mod)),
                  scaled_residual_hist_midcen = scale(residuals(hist_midcen_mod)),
                  scaled_predicted_hist_endcen = scale(predict(hist_endcen_mod)),
                  scaled_residual_hist_endcen =  scale(residuals(hist_endcen_mod)),
                  scaled_predicted_midcen_endcen = scale(predict(midcen_endcen_mod)),
                  scaled_residual_midcen_endcen =  scale(residuals(midcen_endcen_mod)),
                  scaled_PC1_historic = scale(factors$PC1_historic))

#running an lm between scaled_predicted_hist_midcen and scaled_predicted_hist_endcen bc this is the line in p1, and want slope to extend the line
bestfit <-lm(factors2$scaled_predicted_hist_endcen ~ factors2$scaled_predicted_hist_midcen)

intercept <- coef(bestfit)[1]
slope <- coef(bestfit)[2]

#trying to make labels onle for points a certain distance away from line
#factors3 <- factors2 %>%
  #mutate(dist_to_line = abs(slope * 
                              #(scaled_predicted_hist_midcen+scaled_residual_hist_midcen) - 
                              #(scaled_predicted_hist_endcen+scaled_residual_hist_endcen)) + 
                              #intercept / sqrt(slope^2 + (-1)^2),
        # label = case_when(dist_to_line > 1.3 ~ SHORTNAME,
                         #  abs(scaled_predicted_hist_midcen+scaled_residual_hist_midcen) > 1.5 &
                       #     abs(scaled_predicted_hist_endcen+scaled_residual_hist_endcen) > 1.5 ~ SHORTNAME,
                          # .default = NA))


#historic vs midcen residual plot - added scaled predicted to scaled residual
p1 <- ggplot(factors2, aes(x = (scaled_predicted_hist_midcen+scaled_residual_hist_midcen), 
                    y = (scaled_predicted_hist_endcen+scaled_residual_hist_endcen))) +
  geom_point(aes(color = region)) +  
  geom_abline(slope = slope, intercept = intercept,
            linewidth = 1) +
  geom_hline(yintercept = 0, color = "gray") +
  geom_vline(xintercept = 0, color = "gray")  +
  ggtitle("Mid century and End century PC1 residuals with fitted model line") +
  geom_text_repel(aes(label = label, color = region), size = 2.5) +
  theme_bw()

#historic vs midcen residual plot - just scaled residual
p1.5 <- ggplot(factors2, aes(x = scaled_residual_hist_midcen, 
                    y = scaled_residual_hist_endcen)) +
  geom_point(aes(color = region)) +  
  geom_line(aes(x = scaled_predicted_hist_midcen, 
                y = scaled_predicted_hist_endcen),
            linewidth = 1) +
  geom_hline(yintercept = 0, color = "gray") +
  geom_vline(xintercept = 0, color = "gray")  +
  ggtitle("Mid century and End century PC1 residuals relative to historic plotted with fitted model line") +
  theme_bw()


#hist, midcen, endcen plot - not scaled
p2 <- ggplot(factors2, aes(x = PC1_historic)) +
  #historic
  geom_point(aes(y = PC1_historic), color = "red") +  
  geom_line(aes(y = PC1_historic), color = "red") +
  #midcen
  geom_point(aes(y = (PC1_historic+residual_hist_midcen)), color = "blue") +
  geom_line(aes(y = predicted_hist_midcen), color = "blue") +
  #endcen
  geom_point(aes(y = (PC1_historic+residual_hist_endcen)), color = "green") +
  geom_line(aes(y = predicted_hist_endcen), color = "green") +

  geom_hline(yintercept = 0, color = "gray") +
  geom_vline(xintercept = 0, color = "gray")  +
  ggtitle("Historic vs. mid century PC1 residuals plotted with fitted model line ") +
  theme_bw()




```

## Anomalous pH/DO/temp Event Analysis

Identify pH/DO/temp anomalous events. Events for pH are defined as 2+ days where pH \< 7.75, and events for DO are 2+ days where DO \< 4.6. Events for temp defined as 2+ days where daily temp exceeds historical climatology's temp for that month + 2\* historical climatology SD.

Main summary statistic: difference in average number of days in a year an MPA experiences an event across time periods

Function for labeling events:

```{r}
# Create labels for events (e.g., consecutive days with low pH or low DO) using run length encoding

label_events <- function(is_event) {
  event_rle <- rle(is_event)
  labels <- ifelse(event_rle$values, cumsum(event_rle$values), NA)
  rep(labels, event_rle$lengths)
}
```

pH and DO events:

```{r}
ph_do_events <- mpa %>%
  group_by(period) %>%
  select(File, Date, Year,NAME, pH, DO, period, degx, degy, julianday) %>%
  mutate(is_pH_low = pH < 7.75,
         pH_event_id = ifelse(is_pH_low == "FALSE", NA, label_events(is_pH_low)),
         is_DO_low = DO < 4.6,
         DO_event_id = ifelse(is_DO_low == "FALSE", NA, label_events(is_DO_low))) 

#pH event summary
ph_do_event_summary <- ph_do_events %>%

  #ph events (only calculate summary stats when an event took place)
  #note: events with 1 day are changed to NA (bc event defined as 2+ days)
  group_by(File, pH_event_id, period) %>% 
  mutate(pH_duration_days = if_else(is_pH_low == "TRUE", n(), NA),
         pH_duration_days = if_else(pH_duration_days == 1, NA, pH_duration_days), 
         pH_event_id = if_else(pH_duration_days == 0, NA, pH_event_id),
         pH_event_mean = if_else(pH_duration_days > 1, mean(pH, na.rm = TRUE), NA)) %>%
  ungroup() %>%
  
  #do events
  group_by(File, DO_event_id, period) %>% 
  mutate(DO_duration_days = if_else(is_DO_low == "TRUE", n(), NA),
         DO_duration_days = if_else(DO_duration_days == 1, NA, DO_duration_days), 
         DO_event_id = if_else(DO_duration_days == 0, NA, DO_event_id),
         DO_event_mean = if_else(DO_duration_days > 1, mean(DO, na.rm = TRUE), NA)) %>%
  ungroup() 

#since an event can span multiple years, annual sum is n() in that year (not duration days since that can show duration spanning multiple years)
ph_do_annual_event_counts <- ph_do_event_summary %>%
  #per mpa per year
  group_by(File, Year, period) %>%
  mutate(annual_days_below_pH_thresh = sum(!is.na(pH_event_id)),
         annual_days_below_DO_thresh = sum(!is.na(DO_event_id))) %>%
  ungroup() %>%
  #per mpa
  group_by(File, period) %>%
  mutate(avg_annual_days_below_pH_thresh = mean(annual_days_below_pH_thresh, na.rm=TRUE),
         avg_annual_days_below_DO_thresh = mean(annual_days_below_DO_thresh, na.rm=TRUE)) %>%
  ungroup() %>%
  distinct(File, NAME, degx, degy, period, avg_annual_days_below_pH_thresh, 
           avg_annual_days_below_DO_thresh)
```

Temperature events:

```{r}
temp_events <- mpa %>%
  
  #first calculating reference historic climatology
  filter(period == "historic") %>%
  group_by(File, Month) %>%
  summarise(histclim_Temp_mean = mean(Temp), 
            histclim_Temp_sd = sd(Temp)) %>% #this sd is temp over all days in Jan - rather than average of each jan's SD - does it make a difference? 
 
   #bringing back to scale of julian days and all time periods
  merge(mpa, by = c("File", "Month")) %>%
  select(File, Date, Year, NAME, Temp, period, degx, degy, julianday, histclim_Temp_mean, histclim_Temp_sd) %>%
  
  #labelling events
  group_by(period) %>%
  mutate(is_temp_high = Temp > (histclim_Temp_mean + 2*histclim_Temp_sd),
         temp_event_id = ifelse(is_temp_high == "FALSE", NA, label_events(is_temp_high))) %>%
  select(-histclim_Temp_mean, -histclim_Temp_sd) %>%
         
  #temp event summary 
  group_by(File, temp_event_id, period) %>% 
  mutate(temp_duration_days = if_else(is_temp_high == "TRUE", n(), NA),
         temp_duration_days = if_else(temp_duration_days == 1, NA, temp_duration_days), 
         temp_event_id = if_else(temp_duration_days == 0, NA, temp_event_id),
         temp_event_mean = if_else(temp_duration_days > 1, mean(Temp, na.rm = TRUE), NA)) %>%
  ungroup() 
```

Combining all events:

```{r}
#putting it all together
annual_events <- temp_events %>%
  #first finding annual averages for temp
  #per mpa per year
  group_by(File, Year, period) %>%
  mutate(annual_days_above_temp_thresh = sum(!is.na(temp_event_id))) %>%
  ungroup() %>%
  
  #per mpa
  group_by(File, period) %>%
  summarise(avg_annual_days_above_temp_thresh = mean(annual_days_above_temp_thresh, na.rm=TRUE)) %>%
  ungroup() %>%
  right_join(ph_do_annual_event_counts, by = c("File", "period"))
```

Plotting maps:

Setting scale and legend breaks

```{r}
CA_shp <- st_read(here("./data/rawdata/shp/CA_Counties/CA_Counties_TIGER2016.shp"))

make_anom_map(annual_events, "avg_annual_days_above_temp_thresh", "historic", diff = TRUE)

make_anom_map <- function(mpa_summary_df, sumstat, periodt, 
                          palette = "YlOrRd", diff = FALSE)
  
 
#converting df to spatial data format
anom_points <- annual_events %>% 
  st_as_sf(coords = c("degx","degy"), crs = 4326) %>%
  st_transform(st_crs(CA_shp))
                          
#creating period diff df: calculating diff btwn time points for diff sumstat
period_diff <- annual_events %>% 
  
  pivot_wider(names_from = period,
              values_from = c(avg_annual_days_above_temp_thresh,
                              avg_annual_days_below_pH_thresh,
                              avg_annual_days_below_DO_thresh),
              names_glue = "{period}_{.value}") %>%
  
  mutate(midhist_temp_days = midcen_avg_annual_days_above_temp_thresh -
                             historic_avg_annual_days_above_temp_thresh,
         endmid_temp_days =  endcen_avg_annual_days_above_temp_thresh -
                             midcen_avg_annual_days_above_temp_thresh,
         
         midhist_pH_days = midcen_avg_annual_days_below_pH_thresh -
                           historic_avg_annual_days_below_pH_thresh,
         endmid_pH_days =  endcen_avg_annual_days_below_pH_thresh -
                           midcen_avg_annual_days_below_pH_thresh,
        
         midhist_DO_days = midcen_avg_annual_days_below_DO_thresh -
                           historic_avg_annual_days_below_DO_thresh,
         endmid_DO_days =  endcen_avg_annual_days_below_DO_thresh -
                           midcen_avg_annual_days_below_DO_thresh) %>%
  
  mutate(across((14:19), as.integer))
                           
                          
#converting period diff df to spatial data format
period_diff_anom_points <- st_as_sf(period_diff, coords = c("degx","degy"), crs = 4326) %>%
                           st_transform(st_crs(CA_shp))
  
#finding equal interval breaks
#instead of all_diff_breaks
temp_diffs <- c(period_diff$midhist_temp_days, period_diff$endmid_temp_days)
temp_breaks <- classInt::classIntervals(temp_diffs, 10, "equal")
pH_diffs <- c(period_diff$midhist_pH_days, period_diff$endmid_pH_days)
pH_breaks <- classInt::classIntervals(pH_diffs, 10, "equal")
DO_diffs <- c(period_diff$midhist_DO_days, period_diff$endmid_DO_days)
DO_breaks <- classInt::classIntervals(DO_diffs, 10, "equal")
```

Plots - diff in avg annual days crossing threshold between periods

```{r}
#Temp: mid-hist diff map
tm_shape(CA_shp) + 
  tm_polygons(fill = "#ccebc5", lwd = 0) + 
  tm_shape(period_diff_anom_points) + 
  tm_dots(fill = "midhist_temp_days", 
          size = 0.4, 
          fill.scale = tm_scale_intervals(values = "YlOrRd",
                                          breaks = temp_breaks$brks)) +
  tm_layout(bg.color = "#a6cee3") +
  tm_title(
    "Increase in avg annual no. days above temp threshold from historic to mid-century",
    position = tm_pos_out("center", "top"))
#Temp: end-mid diff
tm_shape(CA_shp) + 
  tm_polygons(fill = "#ccebc5", lwd = 0) + 
  tm_shape(period_diff_anom_points) + 
  tm_dots(fill = "endmid_temp_days", 
          size = 0.4, 
          fill.scale = tm_scale_intervals(values = "YlOrRd",
                                          breaks = temp_breaks$brks)) +
  tm_layout(bg.color = "#a6cee3") +
  tm_title(
    "Increase in avg annual no. days above temp threshold from mid-century to end-century",
    position = tm_pos_out("center", "top"))


#pH:mid-hist diff
tm_shape(CA_shp) + 
  tm_polygons(fill = "#ccebc5", lwd = 0) + 
  tm_shape(period_diff_anom_points) + 
  tm_dots(fill = "midhist_pH_days", 
          size = 0.4, 
          fill.scale = tm_scale_intervals(values = "YlOrRd",
                                          breaks = pH_breaks$brks)) +
  tm_layout(bg.color = "#a6cee3") +
  tm_title(
    "Increase in avg annual no. days below pH threshold from historic to mid-century",
    position = tm_pos_out("center", "top"))
#pH:end-mid diff
tm_shape(CA_shp) + 
  tm_polygons(fill = "#ccebc5", lwd = 0) + 
  tm_shape(period_diff_anom_points) + 
  tm_dots(fill = "endmid_pH_days", 
          size = 0.4, 
          fill.scale = tm_scale_intervals(values = "YlOrRd",
                                          breaks = pH_breaks$brks)) +
  tm_layout(bg.color = "#a6cee3") +
  tm_title(
    "Increase in avg annual no. days below pH threshold from mid-century to end-century",
    position = tm_pos_out("center", "top"))


#DO: mid-hist diff
tm_shape(CA_shp) + 
  tm_polygons(fill = "#ccebc5", lwd = 0) + 
  tm_shape(period_diff_anom_points) + 
  tm_dots(fill = "midhist_DO_days", 
          size = 0.4, 
          fill.scale = tm_scale_intervals(values = "YlOrRd",
                                          breaks = DO_breaks$brks)) +
  tm_layout(bg.color = "#a6cee3") +
  tm_title(
    "Increase in avg annual no. days below DO threshold from historic to mid-century",
    position = tm_pos_out("center", "top"))
#DO: end-mid diff
tm_shape(CA_shp) + 
  tm_polygons(fill = "#ccebc5", lwd = 0) + 
  tm_shape(period_diff_anom_points) + 
  tm_dots(fill = "endmid_DO_days", 
          size = 0.4, 
          fill.scale = tm_scale_intervals(values = "YlOrRd",
                                          breaks = DO_breaks$brks)) +
  tm_layout(bg.color = "#a6cee3") +
  tm_title(
    "Increase in avg annual no. days below DO threshold from mid-century to end-century",
    position = tm_pos_out("center", "top"))
```

Plots - actual avg annual days per period and variable

```{r}

#setting new breaks:
all_temp <- c(period_diff$historic_avg_annual_days_above_temp_thresh, period_diff$midcen_avg_annual_days_above_temp_thresh, period_diff$endcen_avg_annual_days_above_temp_thresh) 
all_pH <- c(period_diff$historic_avg_annual_days_below_pH_thresh, period_diff$midcen_avg_annual_days_below_pH_thresh, period_diff$endcen_avg_annual_days_below_pH_thresh)
all_DO <- c(period_diff$historic_avg_annual_days_below_DO_thresh, period_diff$midcen_avg_annual_days_below_DO_thresh, period_diff$endcen_avg_annual_days_below_DO_thresh)

all_temp_breaks <- classInt::classIntervals(all_temp, 10, "equal")
all_pH_breaks <- classInt::classIntervals(all_pH, 10, "equal")
all_DO_breaks <- classInt::classIntervals(all_DO, 10, "equal")

#same story as diff map
#Temp - historic
tm_shape(CA_shp) + 
  tm_polygons(fill = "#ccebc5", lwd = 0) + 
  tm_shape(period_diff_anom_points) + 
  tm_dots(fill = "historic_avg_annual_days_above_temp_thresh", size = 0.4, 
          fill.scale = tm_scale_intervals(values = "YlOrRd", breaks = all_temp_breaks$brks)) +
  tm_layout(bg.color = "#a6cee3") +
  tm_title(
    "Historic avg annual days above temp thresh",
    position = tm_pos_out("center", "top"))
#Temp - midcen
tm_shape(CA_shp) + 
  tm_polygons(fill = "#ccebc5", lwd = 0) + 
  tm_shape(period_diff_anom_points) + 
  tm_dots(fill = "midcen_avg_annual_days_above_temp_thresh", size = 0.4, 
          fill.scale = tm_scale_intervals(values = "YlOrRd", breaks = all_temp_breaks$brks)) +
  tm_layout(bg.color = "#a6cee3") +
  tm_title(
    "Midcen avg annual days above temp thresh",
    position = tm_pos_out("center", "top"))
#Temp - endcen
tm_shape(CA_shp) + 
  tm_polygons(fill = "#ccebc5", lwd = 0) + 
  tm_shape(period_diff_anom_points) + 
  tm_dots(fill = "endcen_avg_annual_days_above_temp_thresh", size = 0.4, 
          fill.scale = tm_scale_intervals(values = "YlOrRd", breaks = all_temp_breaks$brks)) +
  tm_layout(bg.color = "#a6cee3") +
  tm_title(
    "Endcen avg annual days above temp thresh",
    position = tm_pos_out("center", "top"))

#shows diff story than diff map -- in midcen, more days below thresh in norcal. in endcen, much more days everywhere. but relatively the increase is more in socal.
#pH - historic
tm_shape(CA_shp) + 
  tm_polygons(fill = "#ccebc5", lwd = 0) + 
  tm_shape(period_diff_anom_points) + 
  tm_dots(fill = "historic_avg_annual_days_below_pH_thresh", size = 0.4, 
          fill.scale = tm_scale_intervals(values = "YlOrRd", breaks = all_pH_breaks$brks)) +
  tm_layout(bg.color = "#a6cee3") +
  tm_title(
    "Historic avg annual days below pH thresh",
    position = tm_pos_out("center", "top"))
#pH - midcen
tm_shape(CA_shp) + 
  tm_polygons(fill = "#ccebc5", lwd = 0) + 
  tm_shape(period_diff_anom_points) + 
  tm_dots(fill = "midcen_avg_annual_days_below_pH_thresh", size = 0.4, 
          fill.scale = tm_scale_intervals(values = "YlOrRd", breaks = all_pH_breaks$brks)) +
  tm_layout(bg.color = "#a6cee3") +
  tm_title(
    "Midcen avg annual days below pH thresh",
    position = tm_pos_out("center", "top"))
#pH - endcen
tm_shape(CA_shp) + 
  tm_polygons(fill = "#ccebc5", lwd = 0) + 
  tm_shape(period_diff_anom_points) + 
  tm_dots(fill = "endcen_avg_annual_days_below_pH_thresh", size = 0.4, 
          fill.scale = tm_scale_intervals(values = "YlOrRd", breaks = all_pH_breaks$brks)) +
  tm_layout(bg.color = "#a6cee3") +
  tm_title(
    "Endcen avg annual days below pH thresh",
    position = tm_pos_out("center", "top"))

#same story as diff maps
#DO - historic
tm_shape(CA_shp) + 
  tm_polygons(fill = "#ccebc5", lwd = 0) + 
  tm_shape(period_diff_anom_points) + 
  tm_dots(fill = "historic_avg_annual_days_below_DO_thresh", size = 0.4, 
          fill.scale = tm_scale_intervals(values = "YlOrRd", breaks = all_DO_breaks$brks)) +
  tm_layout(bg.color = "#a6cee3") +
  tm_title(
    "Historic avg annual days below DO thresh",
    position = tm_pos_out("center", "top"))
#DO - midcen
tm_shape(CA_shp) + 
  tm_polygons(fill = "#ccebc5", lwd = 0) + 
  tm_shape(period_diff_anom_points) + 
  tm_dots(fill = "midcen_avg_annual_days_below_DO_thresh", size = 0.4, 
          fill.scale = tm_scale_intervals(values = "YlOrRd", breaks = all_DO_breaks$brks)) +
  tm_layout(bg.color = "#a6cee3") +
  tm_title(
    "Midcen avg annual days below DO thresh",
    position = tm_pos_out("center", "top"))
#DO - endcen
tm_shape(CA_shp) + 
  tm_polygons(fill = "#ccebc5", lwd = 0) + 
  tm_shape(period_diff_anom_points) + 
  tm_dots(fill = "endcen_avg_annual_days_below_DO_thresh", size = 0.4, 
          fill.scale = tm_scale_intervals(values = "YlOrRd", breaks = all_DO_breaks$brks)) +
  tm_layout(bg.color = "#a6cee3") +
  tm_title(
    "Endcen avg annual days below DO thresh",
    position = tm_pos_out("center", "top"))
```
