---
title: "MPA Exposure Analysis"
format: 
  html:
    code-fold: true
    toc: true
    page-layout: full
editor: visual
---

```{r}
#| echo: false

knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

------------------------------------------------------------------------

*Characterization of exposure of California MPAs to future pH, DO, and temperature using IPSL model predictions.*

## Setup

Load required libraries, read IPSL MPA projections file, and assign MPAs to regions (Northern CA, Central CA, Southern CA, Channel Islands).

```{r}
#| label: libraries
#| message: false
#| echo: false

library(tidyverse)
library(lubridate)
library(data.table)
library(factoextra)
library(broom)
library(cowplot)
library(respR)
library(here)
library(lattice)
library(RcppRoll)
library(RColorBrewer)
library(gplots)
library(ggpmisc)
library(dendextend)
library(colourvalues)
library(zoo)
library(terra) #needed for tmap?
library(tmap) 
library(sf) 
library(stringr)
library(heatmaply)
library(ggpubr)
library(ggrepel)
```

```{r}
#| label: load_data
#| message: false

#Read mpa data, convert DO units, add a column for time period, and one for difference between mid/end cen and historic.
mpa <- read_csv(here("data/processeddata/model/IPSLmpa.csv")) %>% 
  mutate(DO_mmolL = DO_surf/1000,
         DO_mgL = convert_DO(DO_mmolL, from = "mmol/L", to = "mg/L")) %>% 
  filter(Year <= 2020|
         Year >= 2040 & Year <= 2060|
         Year >= 2080 & Year <= 2100) %>%
  mutate(period = case_when((Year <= 2020) ~ "historic", 
                           (Year %in% c(2040:2060)) ~ "midcen",
                           (Year %in% c(2080:2100)) ~ "endcen")) %>%
  select(-T_bot, -DO_bot, -pH_bot,-DO_surf,-DO_mmolL) %>%
  rename(Temp = T_surf,
         DO = DO_mgL,
         pH = pH_surf) %>%
  mutate(Date = make_date(Year, Month, Day)) %>%
  mutate(julianday = yday(Date))

mpa$File <- substr(mpa$File, 1, nchar(mpa$File)-4)

```

```{r}
#| label: regions
#| message: false
#add regions here!
mpa_centroids<- read_csv(here("data/rawdata/MPA_polygons.csv")) %>%
  select(-Area_sq_mi, -Type)
mpa_centroids$File <- sub("^", "tphdo_mpa_", mpa_centroids$OBJECTID )

mpa_centroids <- mpa_centroids %>%
  mutate(region = ifelse(degy >= 37.29, "norca", 
                         ifelse(degy > 34.8, "centralca", 
                                ifelse("degy" < 34.274 & "degx" < -119.220, "channelisl" , 
                                       "socal"))))

channel <- c("Anacapa Island FMCA", "Anacapa Island FMR", "Anacapa Island SMCA", "Anacapa Island SMR", "Anacapa Island Special Closure", 
             "Arrow Point to Lion Head Point SMCA", "Begg Rock SMR", "Blue Cavern Offshore SMCA", "Blue Cavern Onshore SMCA (No-Take)", 
             "Carrington Point SMR", "Casino Point SMCA (No-Take)", "Cat Harbor SMCA", "Farnsworth Offshore SMCA", "Farnsworth Onshore SMCA", 
             "Footprint FMR", "Footprint SMR", "Gull Island FMR", "Gull Island SMR", "Harris Point FMR", "Harris Point SMR", "Judith Rock SMR", 
             "Long Point SMR", "Lover’s Cove SMCA", "Painted Cave SMCA", "Richardson Rock FMR", "Richardson Rock SMR", "San Miguel Island Special Closure", 
             "Santa Barbara Island FMR", "Santa Barbara Island SMR", "Scorpion FMR", "Scorpion SMR", "Skunk Point SMR", "South Point FMR", "South Point SMR")

mpa_centroids$region[mpa_centroids$NAME %in% channel]  <- "channel" 
mpa <- merge(mpa, mpa_centroids,  by = "File")

```

## Calculate climatology

Create a climatology (for each MPA within a time period, Jan values averaged over all years - same for each month), so each MPA has one value per month that is averaged over all of the years. Seasonal SD is calculated later in summary stat section, and is the variation across these 12 climatologies for each MPA, representing the seasonal cycle.

```{r}
#| label: climatology
#| message: false

#calculate climatology for each MPA for each time period
mpa_climatology <- mpa %>%
  group_by(File, Month, period) %>%
    summarise(T_clim = mean(Temp), 
              pH_clim = mean(pH), 
              DO_clim =   mean(DO))

```

## Calculate event-based variation

To understand event-based variation (not attributed to the seasonal cycle), we need to find the deviation of actual daily values from a climatological value. To do so, a mock dataset with julian day climatologies is created by interpolating monthly climatology values. Actual daily values are subtracted from this interpolated dataset at the daily scale. The standard deviation of this difference is the event-based SD, which is calculated in the summary stat section.

*Q: instead of interpolating from monthly climatologies to create daily climatologies, why not average daily values to get daily climatologies?*

```{r}
#| label: inter_setup
#| message: false
#| warning: false

#' Setup for interpolation:
#' Need to create a day 1 and day 365 proxy. The approx function can only interpolate not extrapolate. The approx function is used to interpolate between the 15th day of each month, and day 1 and day 365 are used to interpolate the first 15 and last 15 days of the year.


##below creates empty vectors as big as we need (365 days per mpa) for each variable and each mpa. julianday is dates 1-365 as many mpa times

mpaslist = unique(mpa$File) 
mpas = rep(NA, 365*121)
julianday = rep(1:365, 121)
interp = rep(NA, 365*121)

#Set up a vector of julian day assignment for the 15th of each month and the first and last day of the year
x_in <- yday(as.Date(c("2001-01-01", 
                       "2001-01-15","2001-02-15","2001-03-15","2001-04-15",
                       "2001-05-15","2001-06-15","2001-07-15","2001-08-15",
                       "2001-09-15","2001-10-15","2001-11-15","2001-12-15", 
                       "2001-12-31")))

# creating a list of all the days of the year to interpolate to.
x_out <- (1:365) 

```

```{r}
#| label: interpolation

interpolate <- function(periodt, variable){
  
  mpa_period_climatology <- mpa_climatology %>%
    filter(period == periodt)
  
  for (i in 1:length(mpaslist)){ #for each mpa
    
    d <- mpa_period_climatology %>% #for given mpa iterating through, selects var from clim
      filter(File == mpaslist[i]) %>% 
      select(File,Month, !!sym(variable))
    
#use a weighted average. for dec 31, slightly more than half weight to dec (denoted by [12,3] index - 12 is the month, third column is the variable's clim value), little less than   half weight to jan. for jan 1, slightly more weight to jan, little less weight to dec. 
    Dec31 = as.numeric(( ((16/30) * (d[12,3])) +  
                           ((14/30) * d[1,3]) )) 
    Jan1 = as.numeric(( ((14/30) * (d[12,3])) + 
                          ((16/30) * d[1,3]) ))

    y_in <- c(Jan1, d[[variable]], Dec31)
    
    mod <- approx(x = x_in, #days to interpolate from
                 y = y_in, #temp values per day in x_in
                 xout = x_out) #1-365 days to interpolate out to
    
    #I don't understand this part
    mpas[((i-1)*365+1):(i*365)] <- mpaslist[i] #rep(mpaslist[1], 365)
    interp[((i-1)*365+1):(i*365)] <- mod$y
  }
  
  df <- data.frame(mpas, julianday, interp) %>%
    rename(File = mpas) %>%
    mutate(period = periodt) %>%
    rename(!!sym(variable) := interp)
  
  assign(x = paste("interp",variable,periodt, sep="_"), value = df, envir = globalenv())
  
}
```

```{r}
#| label: interpolation_function_call


interpolate("historic", "T_clim")
interpolate("midcen", "T_clim")
interpolate("endcen", "T_clim")

interpolate("historic", "pH_clim")
interpolate("midcen", "pH_clim")
interpolate("endcen", "pH_clim")

interpolate("historic", "DO_clim")
interpolate("midcen", "DO_clim")
interpolate("endcen", "DO_clim")
```

```{r}
#| label: combine_interp_dataframes

interpolated_temp <- rbind(interp_T_clim_historic,
                           interp_T_clim_midcen,
                           interp_T_clim_endcen)

interpolated_pH <- rbind(interp_pH_clim_historic,
                           interp_pH_clim_midcen,
                           interp_pH_clim_endcen)

interpolated_DO <- rbind(interp_DO_clim_historic,
                           interp_DO_clim_midcen,
                           interp_DO_clim_endcen)
```

## Create summary statistics

Generate summary statistics for T, pH, and DO.

*Seasonal SD* is across climatologies (monthly values averaged over all years resulting in 12 values per MPA - one per month - **the seasonal SD captures the seasonal cycle,** and is the variation across these values).

*Event SD* is calculated based on the deviation of daily values from interpolated daily values. The standard deviation per MPA is across all of the daily differences between daily value and interpolated values filtered for upwelling months May, June, July, August, and September (*filtered*, not *grouped by* upwelling month - the **Event SD captures variation of the deviations at a daily scale**).

Additional summary stats include the mean, lower 10th percentile, and upper 10th percentile.

```{r}
#| label: summary_stats
#| warning: false

#calculate seasonal SD 
seasonal_SD <- mpa_climatology %>%
  group_by(File, period) %>%
  summarise(T_seasonalSD = sd(T_clim),
            pH_seasonalSD = sd(pH_clim),
            DO_seasonalSD = sd(DO_clim))

#put all summary stats together - first, calculate event SD
sum <- mpa %>%
 
  #merge interpolated climatologies to mpa. (used to calculate event sd)
  merge(interpolated_temp, by = c("File", "julianday", "period")) %>%
  merge(interpolated_pH, by = c("File", "julianday", "period")) %>%
  merge(interpolated_DO, by = c("File", "julianday", "period")) %>%
  
  #subtract actual values - interpolated climatology values 
  #filtered for only upwelling months!!
  mutate(temp_deviation = Temp - T_clim,
         pH_deviation = pH - pH_clim,
         DO_deviation = DO - DO_clim) %>%
  filter(Month == c(5,6,7,8,9)) %>%
  
  #find event based SD 
  group_by(File, period) %>%
    mutate(T_eventSD = sd(temp_deviation),
            pH_eventSD = sd(pH_deviation),
            DO_eventSD = sd(DO_deviation)) %>%
  ungroup() %>%
    select(-T_clim, -pH_clim, -DO_clim, -temp_deviation,
         -pH_deviation, -DO_deviation) %>%
  
    #merge seasonal SD 
    merge(seasonal_SD, by = c("File", "period")) %>%
  
    #adding other summary stats - mean, low 10th and upper 10th percentiles
    group_by(File, period) %>%
    mutate(across(c(Temp, DO, pH), 
                   list(mean = mean, 
                        low10 = ~ quantile(.x, 0.1),
                        high10 = ~quantile(.x,0.9)))) %>%
  
  # to get to scale of one row per mpa
  select(-...1, -Temp, -pH, -DO, -julianday, -Year, -Month, -Day) %>%
  distinct(File, .keep_all = TRUE) %>% ungroup()

```

## PCA

PCA of all summary stats (mean, lower 10th percentile, upper 10th percentile, seasonal SD, event SD) clustered by regions within specified time period.

```{r}
#| label: pca_function


make_pca <- function(df, periodt){
  sum_period <- df %>% filter(period == periodt) 
  sumsub <- sum_period %>% select(-OBJECTID,-NAME, -File, -SHORTNAME, -degx, 
                                  -degy, -region, -period, -Date) 
  
  #only historic time period's pc axis needs to be rotated 
  if (periodt == "endcen"){
  pca <- prcomp(sumsub[,-1], scale = TRUE)
  } else {
  pca <- prcomp(sumsub, scale = TRUE)
  }

  
  fviz_pca_biplot(pca, repel = TRUE,
                  col.var = "black",
                  col.ind = sum_period$region,
                  label ="var",
                  labelsize = 3,
                  addEllipses = TRUE) 
  #fviz_pca_ind(pca, label="none", habillage=sum_period$region,
               #addEllipses=TRUE,  col.ind = sum_period$region)
  
}
```

```{r}
#| label: plot_pca
#| echo: FALSE
#| layout-ncol: 3
#| fig-cap: 
#|   - "Historical period"
#|   - "Mid-century period"
#|   - "End-century period"

make_pca(sum, "historic")
make_pca(sum, "midcen")
make_pca(sum, "endcen") #need to rotate by changing sumsub to sumsub[,-1] inside prcomp
```

## Heatmaps by variable

Create heatmaps per summary stat showing time periods on x-axis, and all MPAs on y-axis to pick up on finer scale geographical patterns in change of summary stats over time.

```{r}
#| label: heatmap_function


col <- colorRampPalette(brewer.pal(9,"YlOrRd"))(256)
invert_col <- colorRampPalette(rev(brewer.pal(9,"YlOrRd")))(256)


make_heatmap <- function(sumstat, color_palette = invert_col, title){
  
  #make matrix for variable of interest
  matrix <- sum %>%
  mutate(color = case_when((region == "centralca") ~ "#ffff99", 
                             (region == "norca") ~ "#beaed4",
                             (region == "socal") ~ "#fdc086",
                             (region == "channel") ~ "#7fc97f")) %>%
  select(!!sym(sumstat), NAME, degy, period, region, color) %>%
  pivot_wider(names_from = period, values_from = sumstat) %>%
  arrange(degy)
  
   

  #making matrix numeric for heatmap to work
  matrix_numeric <- matrix %>%
  arrange(degy) %>%
  select(-NAME, -region, -degy, -color) %>%
  select(historic, midcen, endcen) %>%
  as.matrix()

  row.names(matrix_numeric) <- matrix$NAME
  
  #jpeg(file=here(paste("./figs/heatmap/heatmap",sumstat,".png", 
                       #sep = "_")))
  
  heatmap <- heatmap.2(matrix_numeric, Rowv = FALSE, Colv = FALSE, dendrogram = "none", 
          main = title, tracecol=NA, revC= TRUE,
          margins = c(3,5), col= color_palette, srtCol = 360, 
          labRow = matrix$NAME,cexRow = 0.5, cexCol = 1, keysize = 0.95, 
          key.title = sumstat, key.xlab = NA, key.ylab = NA, par(cex.main=0.5)) #+
    
  #RowSideColors = matrix$color - add this within above - but it is inverted need to fix.
    
    #legend("left",      
    #legend = c("Central CA", "Northern CA", "Southern CA", "Channel Isl"),
    #col = unique(matrix$color), 
    #lty= 1,             
    #lwd = 5,           
    #cex=.7
    #)
  
  #ggsave(here(paste("./figs/heatmap/heatmapTEST.png")), plot)
  
  #dev.off()

}
```

#### **Temp:**

```{r}
#| label: plot_heatmap_temp
#| echo: false
#| layout-ncol: 3

#Temp heatmaps
make_heatmap("Temp_mean", col, title = "Temp (C) mean across periods" )
make_heatmap("Temp_high10", col, title = "Temp highest 10% across periods")
make_heatmap("Temp_low10", col, title = "Temp lowest 10% across periods")
make_heatmap("T_eventSD", col, title = "Temp event SD across periods")
make_heatmap("T_seasonalSD", col, title = "Temp seasonal SD across periods")

```

#### **pH**

```{r}
#| label: plot_heatmap_pH
#| layout-ncol: 3
#| echo: false

#pH heatmaps
make_heatmap("pH_mean", title = "pH mean across periods")
make_heatmap("pH_high10",title = "pH highest 10% across periods" )
make_heatmap("pH_low10", title = "pH lowest 10% across periods")
make_heatmap("pH_eventSD",col, title = "pH event SD across periods")
make_heatmap("pH_seasonalSD",col, title = "pH seasonal SD across periods")
```

#### **DO**

```{r}
#| label: plot_heatmap_DO
#| layout-ncol: 3
#| echo: false

#DO heatmaps
make_heatmap("DO_mean", title = "DO mean across periods")
make_heatmap("DO_high10",title = "DO highest 10% across periods" )
make_heatmap("DO_low10", title = "DO lowest 10% across periods")
make_heatmap("DO_eventSD",col, title = "DO event SD across periods")
make_heatmap("DO_seasonalSD",col, title = "DO seasonal SD across periods")
```

## Regressions

Supplements heatmaps by looking at the correlation between historic and future time periods for a given summary stat. Each point is an MPA, and values for given summary stat within specified time periods are on the axes.

```{r}
#| label: regression_function

make_regression <- function(sumstat, period1, period2){
 
  plot_matrix <- sum %>%
   filter(period == period1 | period == period2) %>%
   select(sumstat, period, NAME, File, OBJECTID, region) %>%
   pivot_wider(names_from = period, values_from = sumstat) 
 
  ggplot(data = plot_matrix, aes_string(x = period1, y = period2)) +
   stat_poly_line(colour = "black") +
   stat_poly_eq(use_label(c("eq", "R2"))) +
   geom_point(aes(colour = region), alpha = 0.7) +
   theme_classic()
 
}
```

#### **Temp**

```{r}
#| label: plot_regression_temp
#| layout-ncol: 2
#| fig-cap: 
#|    - "Historic vs. mid century temp mean"
#|    - "Historic vs. end century temp mean"
#|    - "Historic vs. mid century temp seasonal SD"
#|    - "Historic vs. end century temp seasonal SD"
#|    - "Historic vs. mid century temp event SD"
#|    - "Historic vs. end century temp event SD"
#| echo: false

#temp
make_regression("Temp_mean", "historic", "midcen")
make_regression("Temp_mean", "historic", "endcen")
make_regression("T_seasonalSD", "historic", "midcen")
make_regression("T_seasonalSD", "historic", "endcen")
make_regression("T_eventSD", "historic", "midcen")
make_regression("T_eventSD", "historic", "endcen")
```

#### **pH**

```{r}
#| label: plot_regression_pH_mean
#| echo: false
#| layout-ncol: 2
#| fig-cap: 
#|  - "Historic vs. midcen pH mean"
#|  - "Historic vs. endcen pH mean"
#|  - "Historic vs. midcen seasonal SD"
#|  - "Historic vs. endcen seasonal SD"
#|  - "Historic vs. midcen event SD"
#|  - "Historic vs. endcen event SD" 

 #pH
make_regression("pH_mean", "historic", "midcen")
make_regression("pH_mean", "historic", "endcen")
make_regression("pH_seasonalSD", "historic", "midcen")
make_regression("pH_seasonalSD", "historic", "endcen")
make_regression("pH_eventSD", "historic", "midcen")
make_regression("pH_eventSD", "historic", "endcen")
```

#### **DO**

```{r}
#| label: plot_regression_DO_mean
#| echo: false
#| layout-ncol: 2
#| fig-cap: 
#|  - "Historic vs. midcen DO mean"
#|  - "Historic vs. endcen DO mean"
#|  - "Historic vs. midcen seasonal SD"
#|  - "Historic vs. endcen seasonal SD"
#|  - "Historic vs. midcen event SD"
#|  - "Historic vs. endcen event SD"

#DO
make_regression("DO_mean", "historic", "midcen")
make_regression("DO_mean", "historic", "endcen")
make_regression("DO_seasonalSD", "historic", "midcen")
make_regression("DO_seasonalSD", "historic", "endcen")
make_regression("DO_eventSD", "historic", "midcen")
make_regression("DO_eventSD", "historic", "endcen")
```

## Anomalous pH/DO/temp Event Analysis

Identify pH/DO/temp anomalous events. Events for pH are defined as 2+ days where pH \< 7.75, and events for DO are 2+ days where DO \< 4.6. Events for temp defined as 1+ days where daily temp exceeds historical climatology's temp for that month + 2\* historical climatology SD.

For each [event]{.underline}, the following are calculated: 1) Duration of days, 2) Date event began, 3) Mean pH/DO/T during the event, 4) Intensity of the event (threshold - event mean), 5) Severity of the event (intensity\*duration).

For each [MPA]{.underline}, the following are calculated: 1) Number of events, 2) Mean event duration, 3) Mean of average event pH/DO/T, 4) Mean event intensity, 5) Mean event severity, 6) Average number of events within a year, 7) Average annual pH/DO/T during event

```{r}
#| label: label_events
# Create labels for events (e.g., consecutive days with low pH or low DO) using run 
# length encoding

label_events <- function(is_event) {
  event_rle <- rle(is_event)
  labels <- ifelse(event_rle$values, cumsum(event_rle$values), NA)
  rep(labels, event_rle$lengths)
}

#all event categorization for pH and DO 
events <- mpa %>%
  group_by(period) %>%
  select(File, Date, Year,NAME, pH, DO, period, degx, degy, julianday) %>%
  mutate(is_pH_low = pH < 7.75,
         pH_event = label_events(is_pH_low),
        
         is_DO_low = DO < 4.6,
         DO_event = label_events(is_DO_low),
         
         is_pH_and_DO_low = ifelse(is_pH_low == TRUE & 
                                     is_DO_low == TRUE, TRUE, FALSE),
         pH_and_DO_event = label_events(is_pH_and_DO_low)) 
```

```{r}
#| label: pH_event_summary

#pH event summary
pH_event_summary <- events %>%
  filter(is_pH_low == TRUE) %>% 
  select(-DO_event, -DO, -is_DO_low, -is_pH_and_DO_low, -pH_and_DO_event) %>%
  group_by(File, pH_event) %>% 
  mutate(duration_days = n(),
         event_begin = min(Date),
         event_mean = mean(pH), #mean during the individual event 
         intensity = 7.75 - event_mean,
         severity = intensity*duration_days) %>%
  filter(duration_days > 1) %>%
  ungroup() %>%

  #mpa summary (across periods)
  group_by(File, period) %>%
  mutate(num_event = n_distinct(pH_event), 
         mean_event_duration = mean(duration_days),  
         max_event_duration = max(duration_days),
         mean_event_mean_pH = mean(event_mean), 
         mean_event_intensity = mean(intensity),
         mean_event_severity = mean(severity)) %>%
  ungroup() %>%

  #mpa summary (annual)
  group_by(File, Year, period) %>%
  mutate(annual_days_belowthresh = n(), annual_avg_ph_belowthresh = mean(pH)) %>%
  ungroup() %>%
  group_by(File, period) %>%
  mutate(meanannual_days_belowthresh = mean(annual_days_belowthresh),
         meanannual_avg_pH_belowthresh = mean(annual_avg_ph_belowthresh)) 
  
#now removing event-scale variables not needed to condense to mpa scale
pH_mpa_summary <- pH_event_summary %>%
  select(-duration_days, -event_begin, -event_mean, -intensity, 
         -severity, -is_pH_low,-pH_event, -julianday, -annual_days_belowthresh, -annual_avg_ph_belowthresh) %>%
  distinct(period, File, .keep_all = TRUE) 

#used later
pH_all_events <- left_join(events, pH_event_summary,
                         by = c("File", "Date", "Year", "NAME", "pH", 
                                "period", "degx", "degy", "julianday",
                                "is_pH_low", "pH_event"))
```

```{r}
#| label: DO_event_summary


#DO event summary
DO_event_summary <- events %>%
  filter(is_DO_low == TRUE) %>% 
  select(-pH_event, -pH, -is_pH_low, -is_pH_and_DO_low, -pH_and_DO_event) %>%
  group_by(File, DO_event) %>% 
  mutate(duration_days = n(),
         event_begin = min(Date),
         event_mean = mean(DO), #mean during the individual event 
         intensity = 7.75 - event_mean,
         severity = intensity*duration_days) %>%
    filter(duration_days > 1) %>%
  ungroup() %>%

  #mpa summary (across periods)
  group_by(File, period) %>%
  mutate(num_event = n_distinct(DO_event), 
         mean_event_duration = mean(duration_days),  
         max_event_duration = max(duration_days),
         mean_event_mean_pH = mean(event_mean), 
         mean_event_intensity = mean(intensity),
         mean_event_severity = mean(severity)) %>%
  ungroup() %>%

  #mpa summary (annual)
  group_by(File, Year, period) %>%
  mutate(annual_days_belowthresh = n(), annual_avg_DO_belowthresh = mean(DO)) %>%
  ungroup() %>%
  group_by(File, period) %>%
  mutate(meanannual_days_belowthresh = mean(annual_days_belowthresh),
         meanannual_avg_DO_belowthresh = mean(annual_avg_DO_belowthresh)) 
  
#now removing event-scale variables not needed to condense to mpa scale
DO_mpa_summary <- DO_event_summary %>% 
  select(-duration_days, -event_begin, -event_mean, -intensity, 
         -severity, -is_DO_low, -DO_event, -julianday, -annual_days_belowthresh, -annual_avg_DO_belowthresh) %>%
   distinct(File, period, .keep_all = TRUE) 

#used later
DO_all_events <- left_join(events, DO_event_summary,
                         by = c("File", "Date", "Year", "NAME", "DO", 
                                "period", "degx", "degy", "julianday",
                                "is_DO_low", "DO_event"))
```

```{r}
#| label: temp_event_summary

#temp:
mpa_with_histclimsd <- mpa %>%
  filter(period == "historic") %>%
  group_by(File, Month) %>%
  summarise(hist_T_clim = mean(Temp), hist_T_clim_sd = sd(Temp)) %>% #this sd is temp over all days in Jan - rather than average of each jan's SD - does it make a difference? 
  merge(mpa, by = c("File", "Month"))

temp_event_summary <- mpa_with_histclimsd %>% 
  group_by(period) %>%
  select(File, Date, Year,NAME, Temp, hist_T_clim, hist_T_clim_sd, period, 
         degx, degy, julianday) %>%
  mutate(is_temp_high = Temp > (hist_T_clim + 2*hist_T_clim_sd),
         temp_event = label_events(is_temp_high)) %>%
  
  filter(is_temp_high == TRUE) %>%
  group_by(File, temp_event, period) %>% 
  mutate(duration_days = n(),
         event_begin = min(Date),
         event_mean = mean(Temp),  
         intensity = Temp - (hist_T_clim + 2*hist_T_clim_sd),
         severity = intensity*duration_days) %>%
    filter(duration_days > 1) %>%
  ungroup() %>%

  #mpa summary (across periods)
  group_by(File, period) %>%
  mutate(num_event = n_distinct(temp_event), 
         mean_event_duration = mean(duration_days),  
         max_event_duration = max(duration_days),
         mean_event_mean_temp = mean(event_mean), 
         mean_event_intensity = mean(intensity),
         mean_event_severity = mean(severity)) %>%
  ungroup() %>%
  
  #mpa summary (annual)
  group_by(File, Year, period) %>%
  mutate(annual_days_abovethresh = n(), 
         annual_avg_temp_abovethresh = mean(Temp)) %>%
  ungroup() %>%
  group_by(File, period) %>%
  mutate(meanannual_days_abovethresh = mean(annual_days_abovethresh),
         meanannual_avg_temp_abovethresh = mean(annual_avg_temp_abovethresh)) 
  
  #now removing event-scale variables not needed to condense to mpa scale
temp_mpa_summary <- temp_event_summary %>% 
  select(-Temp,-duration_days, -event_begin, -event_mean, -intensity, 
         -severity, -is_temp_high, -temp_event, -julianday, -annual_days_abovethresh, -annual_avg_temp_abovethresh) %>%
  distinct(period, File, .keep_all = TRUE)

#used later
temp_all_events <- left_join(mpa, temp_event_summary,
                         by = c("File", "Date", "Year", "NAME", "Temp", "period", 
                                "degx", "degy", "julianday"))
```

## Map of anomalous events

Maps showing anomalous event summaries in CA, including average number days crossed threshold in a year, and average value when threshold is crossed. Same color scheme fixed across time periods, using equal breaks classification of values across all periods.

```{r}
#| label: anom_map_function
#| message: FALSE


CA_shp <- st_read(here("./data/rawdata/shp/CA_Counties/CA_Counties_TIGER2016.shp"))


 make_anom_map <- function(mpa_summary_df, sumstat, periodt, 
                          palette = "YlOrRd", diff = FALSE){
  
  #converting df to spatial data format
  anom_points <- st_as_sf((mpa_summary_df %>% filter(period == periodt)), 
                          coords = c("degx","degy"))
  #creating period diff df: calculating diff btwn time points for given sumstat
  period_diff <- temp_mpa_summary %>%
                    pivot_wider(names_from = period,
                                values_from = meanannual_days_abovethresh,
                                id_cols = c(File, degx, degy)) %>%
                    mutate(mid_hist_diff = midcen - historic,
                           end_hist_diff = endcen - historic,
                           end_mid_diff = endcen - midcen)
  #converting period diff df to spatial data format
  period_diff_anom_points <- st_as_sf(period_diff,coords = c("degx","degy"))
  
  #finding equal interval breaks based on all time periods
  all_breaks <- classInt::classIntervals(mpa_summary_df[[sumstat]], 8, "equal")
  all_diff_breaks <- pivot_longer(period_diff, c(mid_hist_diff,end_hist_diff,end_mid_diff))
  all_diff_breaks <- classInt::classIntervals(all_diff_breaks$value, 10, "equal")

   #if diff parameter is false: create regular maps of sumstat
    if (diff == FALSE){
      tm_shape(CA_shp) + #basemap
        tm_fill(col = "#ccebc5") +
        
        tm_shape(anom_points) +
        tm_dots(col = sumstat, palette = palette, size = 0.2, 
                breaks = all_breaks[["brks"]]) +
        
        tm_layout(bg.color = "#a6cee3",
                  legend.title.size = 1,
                  legend.text.size = .8) 

  #if diff parameter is true: create 3 maps of diff between each time period
   } else {
     #mid-hist diff
     mid_hist <- tm_shape(CA_shp) + #basemap
       tm_fill(col = "#ccebc5") +
       
       tm_shape(period_diff_anom_points) +
       tm_dots(col = "mid_hist_diff", palette = palette, size = 0.2, 
               breaks = all_diff_breaks[["brks"]]) +
       
       tm_layout(bg.color = "#a6cee3",
                 legend.title.size = 1,
                 legend.text.size = .8) 
     
     #end-hist diff
     end_hist <- tm_shape(CA_shp) + #basemap
       tm_fill(col = "#ccebc5") +
       
       tm_shape(period_diff_anom_points) +
       tm_dots(col = "end_hist_diff", palette = palette, size = 0.2, 
               breaks = all_diff_breaks[["brks"]]) +
       
       tm_layout(bg.color = "#a6cee3",
                 legend.title.size = 1,
                 legend.text.size = .8)
     
     #end-mid diff
      end_mid <- tm_shape(CA_shp) + #basemap
       tm_fill(col = "#ccebc5") +
       
       tm_shape(period_diff_anom_points) +
       tm_dots(col = "end_mid_diff", palette = palette, size = 0.2, 
               breaks = all_diff_breaks[["brks"]]) +
       
       tm_layout(bg.color = "#a6cee3",
                 legend.title.size = 1,
                 legend.text.size = .8) 
      
      return(list(mid_hist = mid_hist, end_hist = end_hist, end_mid = end_mid))
   }
  }
 
```

#### **Heat events**

```{r}
#| label: plot_anom_map_temp
#| layout-ncol: 3
#| fig-cap: 
#| - "Historic: Avg # days above threshold in a year"
#| - "Midcen: Avg # days above threshold in a year"
#| - "Endcen: Avg # days above threshold in a year"
#| - "Historic: Mean temperature during event"
#| - "Midcen: Mean temperature during event"
#| - "Endcen: Mean temperature during event"
#| column: screen-inset
#| echo: false


#TEMP-----------------------------------------------------
#mean annual num events for all periods for temp 
make_anom_map(temp_mpa_summary, sumstat = "meanannual_days_abovethresh",periodt =  "historic")
make_anom_map(temp_mpa_summary, sumstat = "meanannual_days_abovethresh",periodt = "midcen") 
make_anom_map(temp_mpa_summary, sumstat = "meanannual_days_abovethresh",periodt = "endcen")

#mean temp during event for all periods
make_anom_map(temp_mpa_summary, "meanannual_avg_temp_abovethresh", "historic")
make_anom_map(temp_mpa_summary, "meanannual_avg_temp_abovethresh","midcen") 
make_anom_map(temp_mpa_summary, "meanannual_avg_temp_abovethresh", "endcen")
```

##### Difference in heat events: change between historical and future time periods

```{r, echo=FALSE,results='hide',fig.keep='all'}
#| label: plot_anom_map_temp_diff
#| layout-ncol: 3
#| column: screen-inset
#| fig-cap: 
#| - "Difference between midcen and historic avg days above threshold in a yr"
#| - "Difference between endcen and historic avg days above threshold in a yr"
#| - "Difference between endcen and midcen avg days above threshold in a yr"

make_anom_map(temp_mpa_summary, sumstat = "meanannual_days_abovethresh",period = "historic",diff = TRUE)

```

#### **pH events**

```{r}
#| label: plot_anom_map_pH
#| layout-ncol: 3
#| fig-cap: 
#| - "Historic: Avg # days below threshold in a year"
#| - "Midcen: Avg # days below threshold in a year"
#| - "Endcen: Avg # days below threshold in a year"
#| - "Historic: Mean pH during event"
#| - "Midcen: Mean pH during event"
#| - "Endcen: Mean pH during event"
#| column: screen-inset
#| echo: false


#pH---------------------------------------------------------
make_anom_map(pH_mpa_summary, "meanannual_days_belowthresh", "historic")
make_anom_map(pH_mpa_summary, "meanannual_days_belowthresh","midcen") 
make_anom_map(pH_mpa_summary, "meanannual_days_belowthresh","endcen")

#mean pH during event for all periods 
make_anom_map(pH_mpa_summary, "meanannual_avg_pH_belowthresh", "historic",palette = "-YlOrRd")
make_anom_map(pH_mpa_summary, "meanannual_avg_pH_belowthresh","midcen",palette = "-YlOrRd") 
make_anom_map(pH_mpa_summary, "meanannual_avg_pH_belowthresh","endcen",palette = "-YlOrRd")

```

##### Difference in pH events: change between historical and future time periods

```{r, echo=FALSE,results='hide',fig.keep='all'}
#| label: plot_anom_map_pH_diff
#| layout-ncol: 3
#| column: screen-inset
#| fig-cap: 
#| - "Difference between midcen and historic avg days below threshold in a yr"
#| - "Difference between endcen and historic avg days below threshold in a yr"
#| - "Difference between endcen and midcen avg days below threshold in a yr"

make_anom_map(pH_mpa_summary, sumstat = "meanannual_days_belowthresh",period = "historic",diff = TRUE)
```

#### **DO events**

```{r}
#| label: plot_anom_map_DO
#| layout-ncol: 3
#| fig-cap: 
#| - "Historic: Avg # days below threshold in a year"
#| - "Midcen: Avg # days below threshold in a year"
#| - "Endcen: Avg # days below threshold in a year"
#| - "Historic: Mean pH during event"
#| - "Midcen: Mean DO during event"
#| - "Endcen: Mean DO during event"
#| column: screen-inset
#| echo: false


#DO---------------------------------------------------------
make_anom_map(DO_mpa_summary, "meanannual_days_belowthresh", "historic")
make_anom_map(DO_mpa_summary, "meanannual_days_belowthresh","midcen") 
make_anom_map(DO_mpa_summary, "meanannual_days_belowthresh","endcen")

#mean DO during event for all periods 
make_anom_map(DO_mpa_summary, "meanannual_avg_DO_belowthresh", "historic",palette = "-YlOrRd")
make_anom_map(DO_mpa_summary, "meanannual_avg_DO_belowthresh","midcen",palette = "-YlOrRd") 
make_anom_map(DO_mpa_summary, "meanannual_avg_DO_belowthresh","endcen",palette = "-YlOrRd")
```

##### Difference in DO events: change between historical and future time periods

```{r, echo=FALSE,results='hide',fig.keep='all'}
#| label: plot_anom_map_DO_diff
#| layout-ncol: 3
#| column: screen-inset
#| fig-cap: 
#| - "Difference between midcen and historic avg days below threshold in a yr"
#| - "Difference between endcen and historic avg days below threshold in a yr"
#| - "Difference between endcen and midcen avg days below threshold in a yr"

make_anom_map(DO_mpa_summary, sumstat = "meanannual_days_belowthresh",period = "historic",diff = TRUE)
```

## Severity, Intensity, Duration

Create figures displaying occurrences of events throughout the year, to determine seasonal influence. Julian days on x-axis, years on y-axis, and events are shaded by intensity (threshold minus event mean).

```{r}
#| label: sev_int_dur_function


#NOTE: when made into one function sev_int_dur for all variables (df as a parameter), it takes FOREVER to load while it is very quick when function is specific to variable (temp_sev_int_dur).

#temp:
temp_sev_int_dur <- function(periodt, site){
  
  temp_all_events %>%
    filter(period == periodt, NAME == site) %>%
    ggplot(aes(julianday, Year, fill = intensity)) +
    geom_tile() +
    scale_fill_distiller(palette = "YlOrRd", direction = 1, 
                         na.value = "skyblue") +
    theme_classic() +
    ggtitle(paste(periodt, "Heat events at", site)) 
    
}

temp_sev_int_dur_regional <- function(periodt, regionn){
  
  temp_all_events %>%
    #this find mean intensity of a given julian day from all MPAs in a region
    group_by(period, region, Year, julianday) %>%
    summarise(mean_intensity = mean(intensity, na.rm=TRUE)) %>%
  
    filter(period == periodt, region == regionn) %>%
    ggplot(aes(julianday, Year, fill = mean_intensity)) +
    geom_tile() +
    scale_fill_distiller(palette = "YlOrRd", direction = 1, 
                         na.value = "skyblue") +
    theme_classic() +
    ggtitle(paste("Mean intensity of heat events in", regionn, "during", periodt, "period")) 
    
}

#pH:
pH_sev_int_dur <- function(periodt, site){
  
  pH_all_events %>%
    filter(period == periodt, NAME == site) %>%
    ggplot(aes(julianday, Year, fill = intensity)) +
    geom_tile() +
    scale_fill_distiller(palette = "YlOrRd", direction = 1, 
                         na.value = "skyblue") +
    theme_classic() +
    ggtitle(paste(periodt, "pH events at", site)) 
    

}

pH_sev_int_dur_regional <- function(periodt, regionn){
  
  pH_all_events %>%
    merge(sum[,c("File","region")], by = c("File")) %>% #to get the region
    #this find mean intensity of a given julian day from all MPAs in a region
    group_by(period, region, Year, julianday) %>%
    summarise(mean_intensity = mean(intensity, na.rm=TRUE)) %>%
  
    filter(period == periodt, region == regionn) %>%
    ggplot(aes(julianday, Year, fill = mean_intensity)) +
    geom_tile() +
    scale_fill_distiller(palette = "YlOrRd", direction = 1, 
                         na.value = "skyblue") +
    theme_classic() +
    ggtitle(paste("Mean intensity of pH events in", regionn, "during", periodt, "period")) 
    
}

#DO
DO_sev_int_dur <- function(periodt, site){
  
  DO_all_events %>%
    filter(period == periodt, NAME == site) %>%
    ggplot(aes(julianday, Year, fill = intensity)) +
    geom_tile() +
    scale_fill_distiller(palette = "YlOrRd", direction = 1, 
                         na.value = "skyblue") +
    theme_classic() +
    ggtitle(paste(periodt, "DO events at", site)) 
    

}

DO_sev_int_dur_regional <- function(periodt, regionn){
  
  DO_all_events %>%
    merge(sum[,c("File","region")], by = c("File")) %>% #to get the region
    #this find mean intensity of a given julian day from all MPAs in a region
    group_by(period, region, Year, julianday) %>%
    summarise(mean_intensity = mean(intensity, na.rm=TRUE)) %>%
  
    filter(period == periodt, region == regionn) %>%
    ggplot(aes(julianday, Year, fill = mean_intensity)) +
    geom_tile() +
    scale_fill_distiller(palette = "YlOrRd", direction = 1, 
                         na.value = "skyblue") +
    theme_classic() +
    ggtitle(paste("Mean intensity of DO events in", regionn, "during", periodt, "period")) 
    
}
```

#### Temp

##### Central CA:

```{r}
#| label: plot_temp_sev_int_dur_centralCA
#| layout-ncol: 3
#| echo: false
#| column: screen-inset


#central CA - all sites within region
temp_sev_int_dur_regional("historic", "centralca")
temp_sev_int_dur_regional("midcen", "centralca")
temp_sev_int_dur_regional("endcen", "centralca")

```

##### Northern CA:

```{r}
#| label: plot_temp_sev_int_dur_norCA
#| layout-ncol: 3
#| echo: false
#| column: screen-inset

temp_sev_int_dur_regional("historic", "norca")
temp_sev_int_dur_regional("midcen", "norca")
temp_sev_int_dur_regional("endcen", "norca")
```

##### Southern CA :

```{r}
#| label: plot_temp_sev_int_dur_southernCA
#| layout-ncol: 3
#| echo: false
#| column: screen-inset

temp_sev_int_dur_regional("historic", "socal")
temp_sev_int_dur_regional("midcen", "socal")
temp_sev_int_dur_regional("endcen", "socal")

```

##### Channel Islands:

```{r}
#| layout-ncol: 3
#| echo: false
#| column: screen-inset

temp_sev_int_dur_regional("historic", "channel")
temp_sev_int_dur_regional("midcen", "channel")
temp_sev_int_dur_regional("endcen", "channel")
```

#### pH

##### Central CA:

```{r}
#| layout-ncol: 3
#| echo: false
#| column: screen-inset

pH_sev_int_dur_regional("historic", "centralca")
pH_sev_int_dur_regional("midcen", "centralca")
pH_sev_int_dur_regional("endcen", "centralca")

```

##### Northern CA:

```{r}
#| layout-ncol: 3
#| echo: false
#| column: screen-inset

pH_sev_int_dur_regional("historic", "norca")
pH_sev_int_dur_regional("midcen", "norca")
pH_sev_int_dur_regional("endcen", "norca")

```

##### Southern CA :

```{r}
#| echo: false
#| column: screen-inset
#| layout-ncol: 3

#southern CA
pH_sev_int_dur_regional("historic", "socal")
pH_sev_int_dur_regional("midcen", "socal")
pH_sev_int_dur_regional("endcen", "socal")
```

##### Channel Islands:

```{r}
#| echo: false
#| column: screen-inset
#| layout-ncol: 3

pH_sev_int_dur_regional("historic", "channel")
pH_sev_int_dur_regional("midcen", "channel")
pH_sev_int_dur_regional("endcen", "channel")
```

#### DO

##### Central CA:

```{r}
#| echo: false
#| column: screen-inset
#| layout-ncol: 3

#central CA
DO_sev_int_dur_regional("historic", "centralca")
DO_sev_int_dur_regional("midcen", "centralca")
DO_sev_int_dur_regional("endcen", "centralca")
```

##### Northern CA:

```{r}
#| echo: false
#| column: screen-inset
#| layout-ncol: 3

#northern CA
DO_sev_int_dur_regional("historic", "norca")
DO_sev_int_dur_regional("midcen", "norca")
DO_sev_int_dur_regional("endcen", "norca")
```

##### Southern CA :

```{r}
#| echo: false
#| column: screen-inset
#| layout-ncol: 3

#southern CA
DO_sev_int_dur_regional("historic", "socal")
DO_sev_int_dur_regional("midcen", "socal")
DO_sev_int_dur_regional("endcen", "socal")
```

##### Channel Islands:

```{r}
#| echo: false
#| column: screen-inset
#| layout-ncol: 3

#channel
DO_sev_int_dur_regional("historic", "channel")
DO_sev_int_dur_regional("midcen", "channel")
DO_sev_int_dur_regional("endcen", "channel")

```

## Multiple stressor analysis

Frequency an MPA encounters multiple events. Note: an event is 1+ days, whereas previously it was defined as 2+ days. Thus, total number of possible events for one stressor in a year is 365.

```{r}
multi_event <- function(periodt,regionn, num_stressors){

all_events <- events %>%
  merge(temp_event_summary, by = c("File", "NAME", "Date","Year", "degx", "degy","period","julianday"), suffixes = c("","temp")) %>%
  merge(sum[,c("File","region")], by = c("File")) %>%
  mutate(is_pH_DO_and_T_low = ifelse(is_pH_low == TRUE & 
                                     is_DO_low == TRUE &
                                     is_temp_high == TRUE, TRUE, FALSE),
         pH_DO_and_T_event = label_events(is_pH_DO_and_T_low))

prob_of_multi_event <- all_events %>%
  group_by(julianday, period, Year, region) %>%
  summarise(pH_DO_prob = mean(n_distinct(pH_and_DO_event), na.rm=TRUE)/n_distinct(File),
            pH_DO_T_prob = (mean(n_distinct(pH_and_DO_event), na.rm=TRUE) + 
                              mean(n_distinct(pH_DO_and_T_event), na.rm=TRUE))/n_distinct(File))

 if (num_stressors == "two"){
filter(period == periodt, region == regionn) %>%
  ggplot(aes(julianday, Year, fill = pH_DO_prob)) +
  geom_tile() +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, #change palette
                        na.value = "skyblue") +
  theme_classic() +
  ggtitle(paste("Frequency of MPAs with pH and DO events","in", periodt, "period", "in", regionn)) 
    
  } else {
filter(period == periodt, region == regionn) %>%
  ggplot(aes(julianday, Year, fill = pH_DO_T_prob)) +
  geom_tile() +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, #change palette
                        na.value = "skyblue") +
  theme_classic() +
  ggtitle(paste("Frequency of MPAs with pH, DO, and T events","in", periodt, "period","in", regionn)) 
    
  }
}

testing <- all_events %>% group_by(julianday, period, Year, region) %>%
     mutate(num_pH_event = n_distinct(pH_DO_and_T_event), 
            num_file = n_distinct(File))

multi_event("historic","centralca","pH_DO_prob")
```

## Time series to put in context

Time series for each time period and variable to confirm trends seen previously in PCA, heatmaps, and anom analysis with respect to the seasonal cycle. Time series for a given period show all years in different colors, julian day on the x-axis, and the value of the variable on the y-axis. The blue line represents the climatological mean for the given time period, and the grey ribbon denotes 2SD above/below the period's climatological mean.

```{r}
#| label: time_series_function

make_time_series <- function(periodt, site, variable){

mpa %>% 
  filter(period == periodt, 
         NAME == site) %>%
  mutate(sliding_sd = roll_sd(!!sym(variable), 30, fill = NA)) %>% 
  group_by(julianday) %>%
  mutate(clim_mean = mean(!!sym(variable), na.rm = TRUE),
         clim_sd = mean(sliding_sd, na.rm = TRUE),
         clim_upr = clim_mean + 2 * clim_sd,
         clim_lwr = clim_mean - 2 * clim_sd) %>% 
  ungroup() %>%
  
  ggplot(aes(julianday)) +
  geom_ribbon(aes(ymin = clim_lwr, ymax = clim_upr),
              fill = "black", alpha = 0.75) +
  geom_line(aes(y = !!sym(variable), color = Year, group = Year),
            alpha = 0.5) +
  geom_line(aes(y = clim_mean), size = 1, color = "blue") +
  scale_color_viridis_c() +
  theme_classic() +
    ggtitle(paste(periodt, variable, "at", site))
}




#this fig shows time series in julian days w/ envelope wrt given period's SD (NOT historic, as it should be for calculating anomalous event!)


make_temp_time_series <- function(periodt, site, title){

mpa %>% 
  filter(period == periodt, 
         NAME == site) %>%
  mutate(sliding_sd = roll_sd(Temp, 30, fill = NA)) %>% 
  group_by(julianday) %>%
  mutate(clim_mean = mean(Temp, na.rm = TRUE),
         clim_sd = mean(sliding_sd, na.rm = TRUE),
         clim_upr = clim_mean + 2 * clim_sd,
         clim_lwr = clim_mean - 2 * clim_sd) %>% 
  ungroup() %>%
  
  ggplot(aes(julianday)) +
  geom_ribbon(aes(ymin = clim_lwr, ymax = clim_upr),
              fill = "black", alpha = 0.75) +
  geom_line(aes(y = Temp, color = Year, group = Year),
            alpha = 0.5) +
  geom_line(aes(y = clim_mean), size = 1, color = "blue") +
  scale_color_viridis_c() +
  theme_classic() +
    ggtitle(title)
}

make_pH_time_series <- function(periodt, site, title){

mpa %>% 
  filter(period == periodt, 
         NAME == site) %>%
  mutate(sliding_sd = roll_sd(pH, 30, fill = NA)) %>% 
  group_by(julianday) %>%
  mutate(clim_mean = mean(pH, na.rm = TRUE),
         clim_sd = mean(sliding_sd, na.rm = TRUE),
         clim_upr = clim_mean + 2 * clim_sd,
         clim_lwr = clim_mean - 2 * clim_sd) %>% 
  ungroup() %>%
  
  ggplot(aes(julianday)) +
  geom_ribbon(aes(ymin = clim_lwr, ymax = clim_upr),
              fill = "black", alpha = 0.75) +
  geom_line(aes(y = pH, color = Year, group = Year),
            alpha = 0.5) +
  geom_line(aes(y = clim_mean), size = 1, color = "blue") +
  scale_color_viridis_c() +
  theme_classic() +
    ggtitle(title)
}

make_DO_time_series <- function(periodt, site, title){

mpa %>% 
  filter(period == periodt, 
         NAME == site) %>%
  mutate(sliding_sd = roll_sd(DO, 30, fill = NA)) %>% 
  group_by(julianday) %>%
  mutate(clim_mean = mean(DO, na.rm = TRUE),
         clim_sd = mean(sliding_sd, na.rm = TRUE),
         clim_upr = clim_mean + 2 * clim_sd,
         clim_lwr = clim_mean - 2 * clim_sd) %>% 
  ungroup() %>%
  
  ggplot(aes(julianday)) +
  geom_ribbon(aes(ymin = clim_lwr, ymax = clim_upr),
              fill = "black", alpha = 0.75) +
  geom_line(aes(y = DO, color = Year, group = Year),
            alpha = 0.5) +
  geom_line(aes(y = clim_mean), size = 1, color = "blue") +
  scale_color_viridis_c() +
  theme_classic() +
    ggtitle(title)
  
}

```

#### Temp

##### Central CA:

```{r}
#| column: screen-inset-shaded
#| layout-ncol: 3
#| echo: false

#central
make_temp_time_series("historic","Ano Nuevo SMR", title = "Historic Temp at Ano Nuevo") 
make_temp_time_series("midcen","Ano Nuevo SMR", title = "Midcen Temp at Ano Nuevo") 
make_temp_time_series("endcen","Ano Nuevo SMR", title = "Endcen Temp at Ano Nuevo")
```

##### Northern CA:

```{r}
#| column: screen-inset-shaded
#| layout-ncol: 3
#| echo: false

#northern 
make_temp_time_series("historic","Point Arena SMR", title = "Historic Temp at Point Arena SMR")
make_temp_time_series("midcen","Point Arena SMR", title = "Midcen Temp at Point Arena SMR") 
make_temp_time_series("endcen","Point Arena SMR", title = "Endcen Temp at Point Arena SMR")
```

##### Southern CA:

```{r}
#| column: screen-inset-shaded
#| layout-ncol: 3
#| echo: false

#southern
make_temp_time_series("historic","South La Jolla SMR", title = "Historic Temp at South La Jolla SMR")
make_temp_time_series("midcen","South La Jolla SMR", title = "Midcen Temp at South La Jolla SMR") 
make_temp_time_series("endcen","South La Jolla SMR", title = "Endcen Temp at South La Jolla SMR")
```

##### Channel Islands

```{r}
#| column: screen-inset-shaded
#| layout-ncol: 3
#| echo: false

#channel
make_temp_time_series("historic","Anacapa Island FMR", title = "Historic Temp at Anacapa Island FMR")
make_temp_time_series("midcen","Anacapa Island FMR", title = "Midcen Temp at Anacapa Island FMR") 
make_temp_time_series("endcen","Anacapa Island FMR", title = "Endcen Temp at Anacapa Island FMR")
```

#### pH

##### Central CA:

```{r}
#| column: screen-inset-shaded
#| layout-ncol: 3
#| echo: false

#central
make_pH_time_series("historic","Ano Nuevo SMR", title = "Historic pH at Ano Nuevo") 
make_pH_time_series("midcen","Ano Nuevo SMR", title = "Midcen pH at Ano Nuevo") 
make_pH_time_series("endcen","Ano Nuevo SMR", title = "Endcen pH at Ano Nuevo")
```

##### Northern CA:

```{r}
#| column: screen-inset-shaded
#| layout-ncol: 3
#| echo: false

#northern
make_pH_time_series("historic","Point Arena SMR", title = "Historic pH at Point Arena SMR") 
make_pH_time_series("midcen","Point Arena SMR", title = "Midcen pH at Point Arena SMR") 
make_pH_time_series("endcen","Point Arena SMR", title = "Endcen pH at Point Arena SMR")
```

##### Southern CA:

```{r}
#| column: screen-inset-shaded
#| layout-ncol: 3
#| echo: false

#southern
make_pH_time_series("historic","South La Jolla SMR", title = "Historic pH at South La Jolla SMR") 
make_pH_time_series("midcen","South La Jolla SMR", title = "Midcen pH at South La Jolla SMR") 
make_pH_time_series("endcen","South La Jolla SMR", title = "Endcen pH at South La Jolla SMR")
```

##### Channel Islands

```{r}
#| column: screen-inset-shaded
#| layout-ncol: 3
#| echo: false

#channel
make_pH_time_series("historic","Anacapa Island FMR", title = "Historic pH at Anacapa Island FMR") 
make_pH_time_series("midcen","Anacapa Island FMR", title = "Midcen pH at Anacapa Island FMR") 
make_pH_time_series("endcen","Anacapa Island FMR", title = "Endcen pH at Anacapa Island FMR")
```

#### DO

##### Central CA:

```{r}
#| column: screen-inset-shaded
#| layout-ncol: 3
#| echo: false

#central
make_DO_time_series("historic","Ano Nuevo SMR", title = "Historic DO at Ano Nuevo") 
make_DO_time_series("midcen","Ano Nuevo SMR", title = "Midcen DO at Ano Nuevo") 
make_DO_time_series("endcen","Ano Nuevo SMR", title = "Endcen DO at Ano Nuevo")
```

##### Northern CA:

```{r}
#| column: screen-inset-shaded
#| layout-ncol: 3
#| echo: false

#northern
make_DO_time_series("historic","Point Arena SMR", title = "Historic DO at Point Arena SMR") 
make_DO_time_series("midcen","Point Arena SMR", title = "Midcen DO at Point Arena SMR") 
make_DO_time_series("endcen","Point Arena SMR", title = "Endcen DO at Point Arena SMR")
```

##### Southern CA :

```{r}
#| column: screen-inset-shaded
#| layout-ncol: 3
#| echo: false

#southern
make_DO_time_series("historic","South La Jolla SMR", title = "Historic DO at South La Jolla SMR") 
make_DO_time_series("midcen","South La Jolla SMR", title = "Midcen DO at South La Jolla SMR") 
make_DO_time_series("endcen","South La Jolla SMR", title = "Endcen DO at South La Jolla SMR")
```

##### Channel Islands:

```{r}
#| column: screen-inset-shaded
#| layout-ncol: 3
#| echo: false

#channel
make_DO_time_series("historic","Anacapa Island FMR", title = "Historic DO at Anacapa Island FMR") 
make_DO_time_series("midcen","Anacapa Island FMR", title = "Midcen DO at Anacapa Island FMR") 
make_DO_time_series("endcen","Anacapa Island FMR", title = "Endcen DO at Anacapa Island FMR")
```
